---
title: "bm_LueckenTheisNatMethods2021_visualization"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../output/")})
output:
    html_document:
      toc: yes
      toc_float:
        collapsed: no
        smooth_scroll: no
      df_print: paged

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.height = 10)
knitr::opts_chunk$set(fig.width = 14)
```

```{r}
.libPaths(.Library)
library(tidyverse)
library(anndataR)
library(SingleCellExperiment)
library(yaml)
library(glue)
source("/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/scripts/ggplot_utils.R")
```

# Load config
```{r}
cfg <- read_yaml("/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/scripts/config_files/LueckenTheis_config.yml")
```

# Load metrics
```{r}
metrics_path <- cfg$visualization$metrics_file
samplesheet_path <- cfg$samplesheet
outplot_bm <- cfg$visualization$outplot_dir_bm_visualization

if (!dir.exists(outplot_bm)) {
  dir.create(outplot_bm, recursive = TRUE, showWarnings = FALSE)
}

metrics <- read.delim(metrics_path)
samplesheet <- read.delim(samplesheet_path)
metrics <- metrics %>%
  left_join(samplesheet, by = join_by(sample == sample_name)) %>%
  mutate(
    gene_selection = factor(gene_selection, levels = c("hvg", "angl")),
    integration_method = factor(integration_method)
  )

head(metrics)
```

# Plot
## Total score: angl vs hvg per sample
```{r}
dodge <- position_dodge(width = 0.7)

for (m in unique(metrics$integration_method)) {
  p <- metrics %>%
    filter(integration_method == m) %>%
    ggplot(aes(x = sample, y = Total, fill = gene_selection)) +
    geom_bar(stat = "identity", position = dodge, width = 0.6) +
    scale_fill_manual(values = c("angl" = "#64487A", "hvg" = "grey")) +
    scale_y_continuous(limits = c(0, 1)) +
    theme_bw() +
    theme(
      text = element_text(size = 20, face = "bold"),
      axis.text.x = element_text(angle = 30, hjust = 1),
      legend.position = "right"
    ) +
    labs(
      x = "Dataset",
      y = "Integration Score (Total)",
      title = paste0("Integration method: ", m),
      fill = "Gene selection"
    )
  print(p)
  ggsave(
    file = file.path(outplot_bm, glue("{today()}_{m}_total_score_barplot.pdf")),
    plot = p,
    width = 14,
    height = 8,
    units = "cm"
  )
}
```

## All scores (Total, Bio conservation, Batch correction) — combined
```{r}
for (m in unique(metrics$integration_method)) {
  p <- metrics %>%
    filter(integration_method == m) %>%
    pivot_longer(
      cols = c(Total, Bio.conservation, Batch.correction),
      names_to = "score_type",
      values_to = "score"
    ) %>%
    mutate(score_type = factor(score_type, levels = c("Total", "Bio.conservation", "Batch.correction"))) %>%
    ggplot(aes(x = sample, y = score, fill = gene_selection)) +
    geom_bar(stat = "identity", position = dodge, width = 0.6) +
    scale_fill_manual(values = c("angl" = "#64487A", "hvg" = "grey")) +
    scale_y_continuous(limits = c(0, 1)) +
    facet_wrap(~score_type, ncol = 1) +
    theme_bw() +
    theme(
      text = element_text(size = 18, face = "bold"),
      axis.text.x = element_text(angle = 30, hjust = 1),
      legend.position = "right",
      strip.text = element_text(size = 16)
    ) +
    labs(
      x = "Dataset",
      y = "Score",
      title = paste0("Integration method: ", m),
      fill = "Gene selection"
    )
  print(p)
  ggsave(
    file = file.path(outplot_bm, glue("{today()}_{m}_all_scores_barplot.pdf")),
    plot = p,
    width = 20,
    height = 24,
    units = "cm"
  )
}
```

## angl vs hvg: dotplot across all samples and methods
```{r}
p <- metrics %>%
  ggplot(aes(x = gene_selection, y = Total, fill = gene_selection)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  geom_jitter(
    aes(fill = gene_selection),
    alpha = 0.8,
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7)
  ) +
  scale_fill_manual(values = c("angl" = "#64487A", "hvg" = "grey")) +
  scale_y_continuous(limits = c(0, 1)) +
  facet_wrap(~integration_method) +
  theme_bw() +
  theme(
    text = element_text(size = 20, face = "bold"),
    aspect.ratio = 1,
    legend.position = "none"
  ) +
  labs(x = "Gene Selection", y = "Total Integration Score")
p
ggsave(
  file = file.path(outplot_bm, glue("{today()}_total_score_boxplot.pdf")),
  plot = p,
  width = 14,
  height = 10,
  units = "cm"
)
```

## Bio conservation
```{r}
p <- metrics %>%
  ggplot(aes(x = gene_selection, y = Bio.conservation, fill = gene_selection)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  geom_jitter(
    aes(fill = gene_selection),
    alpha = 0.8,
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7)
  ) +
  scale_fill_manual(values = c("angl" = "#64487A", "hvg" = "grey")) +
  scale_y_continuous(limits = c(0, 1)) +
  facet_wrap(~integration_method) +
  theme_bw() +
  theme(
    text = element_text(size = 20, face = "bold"),
    aspect.ratio = 1,
    legend.position = "none"
  ) +
  labs(x = "Gene Selection", y = "Bio Conservation Score")
p
ggsave(
  file = file.path(outplot_bm, glue("{today()}_bio_conservation_boxplot.pdf")),
  plot = p,
  width = 14,
  height = 10,
  units = "cm"
)
```

## Batch correction
```{r}
p <- metrics %>%
  ggplot(aes(x = gene_selection, y = Batch.correction, fill = gene_selection)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  geom_jitter(
    aes(fill = gene_selection),
    alpha = 0.8,
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7)
  ) +
  scale_fill_manual(values = c("angl" = "#64487A", "hvg" = "grey")) +
  scale_y_continuous(limits = c(0, 1)) +
  facet_wrap(~integration_method) +
  theme_bw() +
  theme(
    text = element_text(size = 20, face = "bold"),
    aspect.ratio = 1,
    legend.position = "none"
  ) +
  labs(x = "Gene Selection", y = "Batch Correction Score")
p
ggsave(
  file = file.path(outplot_bm, glue("{today()}_batch_correction_boxplot.pdf")),
  plot = p,
  width = 14,
  height = 10,
  units = "cm"
)
```

## Individual metrics overview
```{r}
individual_metrics <- c(
  "Isolated.labels", "Silhouette.label", "cLISI",
  "BRAS", "iLISI", "PCR.comparison", "bNMI", "cms", "ldfDiff"
)
# keep only columns that exist
individual_metrics <- intersect(individual_metrics, colnames(metrics))

metrics_long <- metrics %>%
  pivot_longer(
    cols = all_of(individual_metrics),
    names_to = "metric",
    values_to = "value"
  )

for (m in unique(metrics$integration_method)) {
  p <- metrics_long %>%
    filter(integration_method == m) %>%
    ggplot(aes(x = gene_selection, y = value, fill = gene_selection)) +
    geom_boxplot(width = 0.5, outlier.shape = NA) +
    geom_jitter(
      alpha = 0.7,
      position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7)
    ) +
    scale_fill_manual(values = c("angl" = "#64487A", "hvg" = "grey")) +
    facet_wrap(~metric, scales = "free_y") +
    theme_bw() +
    theme(
      text = element_text(size = 15, face = "bold"),
      legend.position = "none",
      strip.text = element_text(size = 13)
    ) +
    labs(
      x = "Gene Selection",
      y = "Score",
      title = paste0("Individual metrics — ", m)
    )
  print(p)
  ggsave(
    file = file.path(outplot_bm, glue("{today()}_{m}_individual_metrics.pdf")),
    plot = p,
    width = 28,
    height = 20,
    units = "cm"
  )
}
```

# Cell type distributions per batch
## sim2
```{r}
sim2_path <- samplesheet %>%
  filter(sample_name == "sim2_raw") %>%
  pull(file_path)

if (length(sim2_path) > 0) {
  sim2 <- read_h5ad(sim2_path, to = "SingleCellExperiment")
  meta <- sim2@colData %>% as.data.frame()

  p1 <- meta %>%
    summarize(n = n(), .by = c(SubBatch, celltype)) %>%
    ggplot(aes(x = SubBatch, y = n, fill = celltype)) +
    geom_col(position = "stack") +
    theme_bw() +
    theme(text = element_text(size = 15, face = "bold")) +
    labs(x = "SubBatch", y = "Cell count", title = "sim2: cell types per SubBatch")
  print(p1)

  p2 <- meta %>%
    summarize(n = n(), .by = c(SubBatch, celltype)) %>%
    ggplot(aes(x = celltype, y = n, fill = SubBatch)) +
    geom_col(position = "stack") +
    theme_bw() +
    theme(
      text = element_text(size = 15, face = "bold"),
      axis.text.x = element_text(angle = 30, hjust = 1)
    ) +
    labs(x = "Cell type", y = "Cell count", title = "sim2: SubBatch per cell type")
  print(p2)
}
```

# Statistical Testing
## angl vs hvg: paired Wilcoxon test on integration scores

One-sample Wilcoxon test on per-sample deltas (angl − hvg).
H0: median Δ = 0 vs H1: median Δ > 0 (anglemania outperforms HVG).
p-values are FDR-adjusted across integration methods.

```{r stat_overall}
diff_df <- metrics %>%
  filter(gene_selection %in% c("angl", "hvg")) %>%
  select(sample, integration_method, gene_selection, Total, Bio.conservation, Batch.correction) %>%
  pivot_wider(
    names_from  = gene_selection,
    values_from = c(Total, Bio.conservation, Batch.correction)
  ) %>%
  mutate(
    delta_Total = Total_angl             - Total_hvg,
    delta_Bio   = Bio.conservation_angl  - Bio.conservation_hvg,
    delta_Batch = Batch.correction_angl  - Batch.correction_hvg
  )

stat_results <- diff_df %>%
  group_by(integration_method) %>%
  summarise(
    n                  = n(),
    median_delta_Total = median(delta_Total, na.rm = TRUE),
    p_Total            = wilcox.test(delta_Total, mu = 0, alternative = "greater")$p.value,
    median_delta_Bio   = median(delta_Bio,   na.rm = TRUE),
    p_Bio              = wilcox.test(delta_Bio,   mu = 0, alternative = "greater")$p.value,
    median_delta_Batch = median(delta_Batch, na.rm = TRUE),
    p_Batch            = wilcox.test(delta_Batch, mu = 0, alternative = "greater")$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    p_Total_fdr  = p.adjust(p_Total,  method = "BH"),
    p_Bio_fdr    = p.adjust(p_Bio,    method = "BH"),
    p_Batch_fdr  = p.adjust(p_Batch,  method = "BH")
  )

stat_results
```

```{r stat_delta_boxplot, fig.height=5, fig.width=10}
diff_long <- diff_df %>%
  pivot_longer(
    cols      = starts_with("delta_"),
    names_to  = "score_type",
    values_to = "delta"
  ) %>%
  mutate(score_type = factor(
    dplyr::recode(score_type,
      delta_Total = "Total",
      delta_Bio   = "Bio conservation",
      delta_Batch = "Batch correction"
    ),
    levels = c("Total", "Bio conservation", "Batch correction")
  ))

p <- ggplot(diff_long, aes(integration_method, delta, fill = integration_method)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_boxplot(outlier.size = 0.5, alpha = 0.8) +
  geom_jitter(width = 0.1, alpha = 0.6, size = 2) +
  stat_summary(fun = median, geom = "point", shape = 18, size = 4, color = "white") +
  scale_fill_manual(values = c("scvi" = "#4E79A7", "scanorama" = "#F28E2B")) +
  facet_wrap(~score_type) +
  labs(
    title    = "Score difference: anglemania − HVG (paired by dataset)",
    subtitle = "Each point = one dataset; dashed line at 0",
    x = "Integration method", y = "Δ score (angl − hvg)", fill = NULL
  ) +
  theme_bw() +
  theme(text = element_text(size = 14), legend.position = "none")
p
ggsave(
  file.path(outplot_bm, glue("{today()}_stat_delta_boxplot.pdf")),
  p, width = 14, height = 6, units = "cm"
)
```

## Per-metric individual tests

```{r stat_individual_metrics}
individual_metrics <- c(
  "Isolated.labels", "Silhouette.label", "cLISI",
  "BRAS", "iLISI", "PCR.comparison", "bNMI", "cms", "ldfDiff"
)
individual_metrics <- intersect(individual_metrics, colnames(metrics))

diff_individual <- metrics %>%
  filter(gene_selection %in% c("angl", "hvg")) %>%
  select(sample, integration_method, gene_selection, all_of(individual_metrics)) %>%
  pivot_wider(names_from = gene_selection, values_from = all_of(individual_metrics))

stat_individual <- diff_individual %>%
  group_by(integration_method) %>%
  summarise(
    across(
      all_of(paste0(individual_metrics, "_angl")),
      .fns = list(
        median_delta = ~ median(. - get(sub("_angl$", "_hvg", cur_column())), na.rm = TRUE),
        p_value      = ~ tryCatch(
          wilcox.test(. - get(sub("_angl$", "_hvg", cur_column())),
                      mu = 0, alternative = "greater")$p.value,
          error = function(e) NA_real_
        )
      ),
      .names = "{sub('_angl$', '', .col)}_{.fn}"
    ),
    .groups = "drop"
  )

stat_individual
```

# Session Info
```{r}
sessionInfo()
```
