---
title: "bm_simulated_visualization"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../output/")})
output:
    html_document:
      toc: yes
      toc_float:
        collapsed: no
        smooth_scroll: no
      df_print: paged

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.height = 10)
knitr::opts_chunk$set(fig.width = 14)
```

```{r}
library(tidyverse)
library(Seurat)
library(anndataR)
library(bigstatsr)
# devtools::load_all("../../anglemania")
library(matrixStats)
library(pbapply)
library(glue)
source("/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/scripts/ggplot_utils.R")
today <- function() {
  return(format(Sys.Date(), "%Y%m%d"))
}
```


# dropout
* sparsity was not calculated yet, so I'll add this as a column to the samplesheet
```{r}
load_and_calc_sparsity <- function(file_path) {
  adata <- read_h5ad(file_path)
  counts <- adata$X
  total_elements <- nrow(counts) * ncol(counts)
  sparsity <- sum(counts == 0) / total_elements
  return(round(sparsity, 4) * 100)
}

```

## load metrics
```{r}
config <- yaml::read_yaml("/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/scripts/config_files/config_varying_batch_biovar_dropout_sample.yml")
metrics <- read.csv(config$visualization$metrics_file, sep = "\t", stringsAsFactors = FALSE)
head(metrics)
colnames(metrics)
samplesheet <- read.csv(config$samplesheet, sep = "\t", stringsAsFactors = FALSE)

if (!("sparsity" %in% colnames(samplesheet))) {
  samplesheet$sparsity <- pblapply(samplesheet$file_path, load_and_calc_sparsity, cl = 10) %>% unlist()
  samplesheet <- samplesheet %>% # for each dropout.mid level, calculate the mean sparsity because they are like 74 and 75 % for the same dropout, but I want to use it as a factor later
    mutate(sparsity = round(mean(sparsity)), .by = dropout.mid)
  write.table(samplesheet, file = config$samplesheet, sep = "\t", row.names = FALSE, quote = FALSE)
}

metrics <- metrics %>%
  left_join(samplesheet %>% select(-original_h5ad), by = join_by(sample == sample_name)) %>%
    mutate(
      de.facLoc = case_when(
        de.facLoc == 0.05 ~ "mild",
        de.facLoc == 0.1 ~ "moderate",
        de.facLoc == 0.3 ~ "strong",
      ),
      de.facLoc = factor(de.facLoc, levels = c("mild", "moderate", "strong")),
      batch.facLoc = case_when(
        batch.facLoc == 0.05 ~ "mild",
        batch.facLoc == 0.1 ~ "moderate",
        batch.facLoc == 0.3 ~ "strong",
      ),
      batch.facLoc = factor(batch.facLoc, levels = c("mild", "moderate", "strong")),
      dropout.mid = factor(dropout.mid),
      gene_selection = factor(gene_selection),
      integration_method = factor(integration_method),
      sparsity = factor(sparsity)
  ) 

head(metrics)

```

```{r}
outplot <- config$visualization$outplot_dir_bm_visualization
if (!dir.exists(outplot)) {
  dir.create(outplot, recursive = TRUE)
}
```

## Plot
### Heatmap integration score as a function of biol. variab. and batch effects
```{r}
for (m in unique(metrics$integration_method)) {
  for (d in unique(metrics$sparsity)) {
    for (g in unique(metrics$gene_selection)) {
      p <- metrics %>%
        filter(integration_method == m, sparsity == d, gene_selection == g) %>%
        ggplot(aes(x = batch.facLoc, y = de.facLoc)) +
        geom_tile(aes(fill = Total, color = Total)) +
        geom_text(aes(label = round(Total, 2)), color = "black", size = 4) +
        coord_equal() +
        labs(x = "Batch Effects", y = "Biological Variability", title = paste0(m, "-sparsity", d, "perc-", g, "-TotalScore")) +
        scale_fill_gradient(low = "grey", high = "#64487A", limits = c(0, 1)) +
        scale_color_gradient(low = "grey", high = "#64487A", limits = c(0, 1))
      print(p)
    }
  }
}
```

### violin plots
#### Total Score
* Here, I plot the sparsity 
```{r}
dodge <- position_dodge(width = 0.7)
for (m in unique(metrics$integration_method)) {
  for (de in unique(metrics$de.facLoc)) {
      p <- metrics %>%
        filter(de.facLoc == de, integration_method == m) %>%
        ggplot(aes(x = as.factor(sparsity), y = Total)) +
        # geom_violin(aes(fill = gene_selection), position = dodge, width = 2)+
        geom_boxplot(width = 0.5, aes(fill = gene_selection), position = dodge) +
        geom_jitter(
          aes(fill = gene_selection), 
          alpha = 0.7, 
          position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7)
          ) +
        # make manual color
        scale_fill_manual(values = c("angl" = "#64487A", "hvg" = "grey"))+
        theme_bw() +
        labs(
          x = "Sparsity [%]", 
          y = "Integration Score", 
          title = glue("{m}\nTotalScore by sparsity\nbiol. variability: {de}")
          ) +
        ylim(0, 1) +
        # increase font size of all text
        theme(
          aspect.ratio = 1,
          legend.position = c(0.8, 0.1)
        ) + 
        theme_big_text()

        print(p)
        ggsave(
            glue("{outplot}/{today()}_{m}_TotalScore_violin.pdf"),
            width = 8, 
            height = 6,
            plot = p
        )
    
      
    }
  }

```

### scatter plot
#### total score
```{r}
for (m in unique(metrics$integration_method)) {
    p <- metrics %>%
      filter(
        integration_method == m
      ) %>%
      group_by(de.facLoc, sparsity, integration_method, gene_selection) %>%
        summarise(mean_total = mean(Total), .groups = "drop") %>%
        ggplot(aes(x = sparsity, y = mean_total, color = de.facLoc)) +
        geom_point(size = 6, alpha = 0.8, aes(shape = gene_selection)) +
        ylim(0, 1) +
        guides(color = guide_legend(ncol = 2)) +
        scale_color_brewer(palette = "Dark2") +
        labs(
          x = "Sparsity [%]",
          y = "Integration Score",
          color = "Biological Variability",
          title = glue("{m}\nTotal Score by \nsparsity and de.facLoc")
        ) +
        big_text_theme()

      print(p)
      ggsave(
        glue("{outplot}/{today()}_{m}_TotalScore_by_sparsity_and_de.facLoc_ScatterPlot.pdf"),
        width = 10, 
        height = 10,
        plot = p
      )
  }

```

#### Bio score
```{r}
for (m in unique(metrics$integration_method)) {
  p <- metrics %>%
    filter(
      integration_method == m
    ) %>%
    group_by(de.facLoc, sparsity, integration_method, gene_selection) %>%
    summarise(mean_total = mean(Total), .groups = "drop") %>%
    ggplot(aes(x = sparsity, y = mean_total, color = de.facLoc)) +
    geom_point(size = 6, alpha = 0.8, aes(shape = gene_selection)) +
    ylim(0, 1) +
    guides(color = guide_legend(ncol = 2)) +
    scale_color_brewer(palette = "Dark2") +
    labs(
      x = "Sparsity [%]",
      y = "Bio Conservation Score",
      color = "Biological Variability",
      title = glue("{m}\nBio Conservation Score by \nsparsity and de.facLoc")
    ) +
    big_text_theme()

  print(p)
  ggsave(
    glue("{outplot}/{today()}_{m}_BioConservationScore_by_sparsity_and_de.facLoc_ScatterPlot.pdf"),
    width = 10, 
    height = 10,
    plot = p
  )
}
```

#### Batch correction score
```{r}
for (m in unique(metrics$integration_method)) {
  p <- metrics %>%
    filter(
      integration_method == m
    ) %>%
    group_by(de.facLoc, sparsity, integration_method, gene_selection) %>%
    summarise(mean_total = mean(Batch.correction), .groups = "drop") %>%
    ggplot(aes(x = sparsity, y = mean_total, color = de.facLoc)) +
    geom_point(size = 6, alpha = 0.8, aes(shape = gene_selection)) +
    ylim(0, 1) +
    guides(color = guide_legend(ncol = 2)) +
    scale_color_brewer(palette = "Dark2") +
    labs(
      x = "Sparsity [%]",
      y = "Batch Correction Score",
      color = "Biological Variability",
      title = glue("{m}\nBatch Correction Score by \nsparsity and de.facLoc")
    ) +
    big_text_theme()

  print(p)
  ggsave(
    glue("{outplot}/{today()}_{m}_BatchCorrectionScore_by_sparsity_and_de.facLoc_ScatterPlot.pdf"),
    width = 10, 
    height = 10,
    plot = p
  )
}
```



# Metacells
## load metrics
* because we only kept the samples that had reliable SEACell cell type purity (purity >= 08)
```{r}
# /data/akalin/akollot/anglemania_benchmark/simulated/with_dropout/SEACells/metrics/lower_batch_biovar_combined_metrics.tsv

metrics <- read.csv("/data/akalin/akollot/anglemania_benchmark/simulated/with_dropout/SEACells/metrics/lower_batch_biovar_combined_metrics.tsv", sep = "\t", stringsAsFactors = FALSE)
head(metrics)
colnames(metrics)
samplesheet <- read.csv("/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/h5ad/with_dropout/own_sim/SEACells/re_merged/filtered/samplesheet.tsv", sep = "\t", stringsAsFactors = FALSE)
colnames(samplesheet)
if (!("sparsity" %in% colnames(samplesheet))) {
    samplesheet$sparsity <- pblapply(samplesheet$file_path, load_and_calc_sparsity, cl = 40) %>% unlist()
    samplesheet <- samplesheet %>% # for each dropout.mid level, calculate the mean sparsity because they are like 74 and 75 % for the same dropout, but I want to use it as a factor later
      mutate(sparsity = round(mean(sparsity)), .by = dropout.mid)
  write.table(samplesheet, file = "/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/h5ad/with_dropout/own_sim/SEACells/re_merged/filtered/samplesheet.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
}

metrics <- metrics %>%
  left_join(samplesheet, by = join_by(sample == sample_name)) %>%
  mutate(
    de.facLoc = factor(de.facLoc),
    batch.facLoc = factor(batch.facLoc),
    dropout.mid = factor(dropout.mid),
    gene_selection = factor(gene_selection),
    integration_method = factor(integration_method),
    sparsity = factor(sparsity)
  )

View(metrics)
```

```{r}
outplot <- "/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/output/plots/bm_simulated_visualization_dropout/"
```

## Plot
### Heatmap integration score as a function of biol. variab. and batch effects
```{r}
View(metrics %>% filter(integration_method == "scvi", gene_selection == "angl"))
for (m in unique(metrics$integration_method)) {
  for (d in unique(metrics$sparsity)) {
    for (g in unique(metrics$gene_selection)) {
      p <- metrics %>%
        filter(integration_method == m, sparsity == d, gene_selection == g) %>%
        ggplot(aes(x = batch.facLoc, y = de.facLoc)) +
        geom_tile(aes(fill = Total, color = Total)) +
        geom_text(aes(label = round(Total, 2)), color = "black", size = 4) +
        coord_equal() +
        labs(x = "Batch Effects", y = "Biological Variability", title = paste0(m, "-sparsity", d, "perc-", g, "-TotalScore")) +
        scale_fill_gradient(low = "grey", high = "#64487A", limits = c(0, 1)) +
        scale_color_gradient(low = "grey", high = "#64487A", limits = c(0, 1))
      print(p)
    }
  }
}
```

### violin plots
#### Total Score
* Here, I plot the sparsity 
```{r}

dodge <- position_dodge(width = 0.7)
for (m in unique(metrics$integration_method)) {
  for (de in unique(metrics$de.facLoc)) {
      p <- metrics %>%
        filter(de.facLoc == de, integration_method == m) %>%
        ggplot(aes(x = as.factor(sparsity), y = Total)) +
        # geom_violin(aes(fill = gene_selection), position = dodge, width = 2)+
        geom_boxplot(width = 0.5, aes(fill = gene_selection), position = dodge) +
        geom_jitter(
          aes(fill = gene_selection), 
          alpha = 0.7, 
          position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7)
          ) +
        # make manual color
        scale_fill_manual(values = c("angl" = "#64487A", "hvg" = "grey"))+
        theme_bw() +
        labs(
          x = "Sparsity [%]", 
          y = "Integration Score", 
          title = paste0(m, "-de.facLoc", de, "-TotalScore")
          ) +
        ylim(0, 1) +
        # increase font size of all text
        theme(
          text = element_text(size = 30, face = "bold"),
          aspect.ratio = 1,
          legend.position = c(0.8, 0.1)
        )

        print(p)
        # ggsave(
        #     file.path(outplot, paste0(m, "_ngroup", i, "_nbatch", j, "_TotalScore_violin.pdf")),
        #     width = 10, 
        #     height = 10,
        #     plot = p
        # )
    
      
    }
  }

```

### scatter plot
#### total score
```{r}
for (m in unique(metrics$integration_method)) {
    p <- metrics %>%
      filter(
        integration_method == m
      ) %>%
      group_by(de.facLoc, sparsity, integration_method, gene_selection) %>%
        summarise(mean_total = mean(Total), .groups = "drop") %>%
        ggplot(aes(x = sparsity, y = mean_total, color = de.facLoc)) +
        geom_point(size = 6, alpha = 0.8, aes(shape = gene_selection)) +
        ylim(0, 1) +
        guides(color = guide_legend(ncol = 2)) +
        scale_color_brewer(palette = "Dark2") +
        labs(
          x = "Sparsity [%]",
          y = "Integration Score",
          title = paste0(m, " - Total Score by \nsparsity and de.facLoc")
        ) +
        big_text_theme()

      print(p)
    ggsave(file.path(outplot, paste0(m, "_TotalScore_by_sparsity_and_de.facLoc_ScatterPlo_Metacells.pdf")), width = 10, height = 10, plot = p)
  }

```

#### Bio score
```{r}
for (m in unique(metrics$integration_method)) {
  p <- metrics %>%
    filter(
      integration_method == m
    ) %>%
    group_by(de.facLoc, sparsity, integration_method, gene_selection) %>%
    summarise(mean_total = mean(Total), .groups = "drop") %>%
    ggplot(aes(x = sparsity, y = mean_total, color = de.facLoc)) +
    geom_point(size = 6, alpha = 0.8, aes(shape = gene_selection)) +
    ylim(0, 1) +
    guides(color = guide_legend(ncol = 2)) +
    scale_color_brewer(palette = "Dark2") +
    labs(
      x = "Sparsity [%]",
      y = "Bio Conservation Score",
      title = paste0(m, " - Bio Conservation Score by \nsparsity and de.facLoc")
    ) +
    big_text_theme()

  print(p)
  ggsave(file.path(outplot, paste0(m, "_BioConservationScore_by_sparsity_and_de.facLoc_ScatterPlot_Metacells.pdf")), width = 10, height = 10, plot = p)
}
```

# Comparison Metacells vs normalCells
```{r}
metrics_metacells <- read.csv(
  "/data/akalin/akollot/anglemania_benchmark/simulated/with_dropout/SEACells/metrics/lower_batch_biovar_combined_metrics.tsv",
  sep = "\t", 
  stringsAsFactors = FALSE
)
samplesheet_metacells <- read.csv(
  "/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/h5ad/with_dropout/own_sim/SEACells/re_merged/filtered/samplesheet.tsv",
  sep = "\t",
  stringsAsFactors = FALSE
)
metrics_normalcells <- read.csv(
  "/data/akalin/akollot/anglemania_benchmark/simulated/with_dropout/metrics/lower_batch_biovar_combined_metrics.tsv",
  sep = "\t", 
  stringsAsFactors = FALSE
)
samplesheet_normalcells <- read.csv(
  "/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/h5ad/with_dropout/own_sim/samplesheet_low_facLoc.tsv",
  sep = "\t", 
  stringsAsFactors = FALSE
)

metrics_metacells <- metrics_metacells %>%
  mutate(
    metacells = TRUE,
    parent_sample = gsub("_filtered", "", sample)
  )
metrics_normalcells <- metrics_normalcells %>%
  mutate(
    metacells = FALSE,
    parent_sample = sample
  )
samplesheet_metacells <- samplesheet_metacells %>%
  mutate(
    sample_name = gsub("_filtered", "", sample_name)
  )
samplesheet <- samplesheet_metacells %>%
  select(sample_name) %>%
  left_join(samplesheet_normalcells)
metrics <- bind_rows(metrics_metacells, metrics_normalcells)

metrics <- metrics %>%
  filter(parent_sample %in% samplesheet$sample_name)
head(samplesheet_metacells)

metrics <- metrics %>%
  left_join(samplesheet, by = join_by(parent_sample == sample_name)) %>%
  mutate(
    de.facLoc = factor(de.facLoc),
    batch.facLoc = factor(batch.facLoc),
    dropout.mid = factor(dropout.mid),
    gene_selection = factor(gene_selection),
    integration_method = factor(integration_method),
    sparsity = factor(sparsity)
  )
View(metrics %>% arrange(batch.facLoc, de.facLoc, parent_sample))
```

```{r}
for (m in unique(metrics$integration_method)) {
  for (d in unique(metrics$sparsity)) {
    for (g in unique(metrics$gene_selection)) {
      for (me in unique(metrics$metacells)) {
        p <- metrics %>%
          filter(integration_method == m, sparsity == d, gene_selection == g, metacells == me) %>%
          ggplot(aes(x = batch.facLoc, y = de.facLoc)) +
          geom_tile(aes(fill = Total, color = Total)) +
          geom_text(aes(label = round(Total, 2)), color = "black", size = 4) +
          coord_equal() +
          labs(x = "Batch Effects", y = "Biological Variability", title = paste0("metacells:", me, "-", m, "-sparsity", d, "perc-", g, "-TotalScore")) +
          scale_fill_gradient(low = "grey", high = "#64487A", limits = c(0, 1)) +
          scale_color_gradient(low = "grey", high = "#64487A", limits = c(0, 1))
        print(p)
      }
    }
  }
}
```

# make complex heatmap

```{r}
samplesheet %>%
  summarise(n = n(), .by = c(batch.facLoc, de.facLoc))
metrics %>%
  summarise(n = n(), .by = c(sparsity))
```

```{r}
dodge <- position_dodge(width = 0.7)
for (m in unique(metrics$integration_method)) {
  for (g in unique(metrics$gene_selection)) {
    p <- metrics %>%
      filter(integration_method == m, gene_selection == g) %>%
      ggplot(aes(x = sparsity, y = Total, fill = metacells)) +
      geom_boxplot(width = 0.5, position = dodge) +
      geom_jitter(
        alpha = 0.7,
        position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7)
      ) +
      labs(
        x = "Sparsity [%]",
        y = "Integration Score",
        title = paste0(m, "-", g, " - Total Score by \nsparsity and de.facLoc")
      ) +
      facet_wrap(~de.facLoc) 

    print(p)
  }
}
```

```{r}
for (m in unique(metrics$integration_method)) {
  for (g in unique(metrics$gene_selection)) {
    p <- metrics %>%
      filter(integration_method == m, gene_selection == g) %>%
      group_by(de.facLoc, sparsity, metacells) %>%
      summarise(mean_total = mean(Total), .groups = "drop") %>%
      ggplot(aes(x = sparsity, y = mean_total, color = de.facLoc)) +
      geom_point(size = 6, alpha = 0.8, aes(shape = metacells)) +
      ylim(0, 1) +
      guides(color = guide_legend(ncol = 2)) +
      scale_color_brewer(palette = "Dark2") +
      labs(
        x = "Original sparsity [%]",
        y = "Integration Score",
        title = paste0(m, "-", g, " - Total Score by \nsparsity and de.facLoc")
      ) +
      big_text_theme()

    print(p)
  }
}

```

```{r}
for (m in unique(metrics$integration_method)) {
  for (g in unique(metrics$gene_selection)) {
    p <- metrics %>%
      filter(integration_method == m, gene_selection == g) %>%
      group_by(de.facLoc, sparsity, metacells) %>%
      summarise(mean_total = mean(Bio.conservation), .groups = "drop") %>%
      ggplot(aes(x = sparsity, y = mean_total, color = de.facLoc)) +
      geom_point(size = 6, alpha = 0.8, aes(shape = metacells)) +
      ylim(0, 1) +
      guides(color = guide_legend(ncol = 2)) +
      scale_color_brewer(palette = "Dark2") +
      labs(
        x = "Original sparsity [%]",
        y = "Bio Conservation Score",
        title = paste0(m, "-", g, " - Bio Conservation Score by \nsparsity and de.facLoc")
      ) +
      big_text_theme()

    print(p)
  }
}

```


```{r}
for (m in unique(metrics$integration_method)) {
  for (g in unique(metrics$gene_selection)) {
    p <- metrics %>%
      filter(integration_method == m, gene_selection == g) %>%
      group_by(de.facLoc, sparsity, metacells) %>%
      summarise(mean_total = mean(Batch.correction), .groups = "drop") %>%
      ggplot(aes(x = sparsity, y = mean_total, color = de.facLoc)) +
      geom_point(size = 6, alpha = 0.8, aes(shape = metacells)) +
      ylim(0, 1) +
      guides(color = guide_legend(ncol = 2)) +
      scale_color_brewer(palette = "Dark2") +
      labs(
        x = "Original sparsity [%]",
        y = "Batch Correction Score",
        title = paste0(m, "-", g, " - Batch Correction Score by \nsparsity and de.facLoc")
      ) +
      big_text_theme()

    print(p)
  }
}

```

```{r}
sessionInfo()
```