---
title: "bm_simulated_visualization"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../output/")})
output:
    html_document:
      toc: yes
      toc_float:
        collapsed: no
        smooth_scroll: no
      df_print: paged

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.height = 10)
knitr::opts_chunk$set(fig.width = 14)
```

```{r}
# .libPaths(.Library)
library(tidyverse)
library(Seurat)
library(anndataR)
library(bigstatsr)
# devtools::load_all("../../anglemania")
library(matrixStats)
library(yaml)
```

# Load config
```{r}
cfg <- read_yaml("/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/scripts/config_files/config_varying_batch_biovar_facScale0.1_pyanglemania.yml")
```

# Load metrics
```{r}
# Extract relevant paths
metrics_path <- cfg$visualization$metrics_file
samplesheet_path <- cfg$samplesheet
outplot_bm <- cfg$visualization$outplot_dir_bm_visualization
# Ensure plot directories exist
if (!dir.exists(outplot_bm)) {
  dir.create(outplot_bm, recursive = TRUE, showWarnings = FALSE)
}

# Load data
metrics <- read.delim(metrics_path)
samplesheet <- read.delim(samplesheet_path)

metrics <- metrics %>%
  left_join(samplesheet %>% select(-original_h5ad), by = join_by(sample == sample_name)) %>%
  mutate(
    de.facLoc = factor(de.facLoc),
    batch.facLoc = factor(batch.facLoc),
    gene_selection = factor(gene_selection),
    integration_method = factor(integration_method)
  )  %>%
    mutate(
      de.facLoc = case_when(
        de.facLoc == 0 ~ "none",
        de.facLoc == 0.01 ~ "small",
        de.facLoc == 0.05 ~ "mild",
        de.facLoc == 0.1 ~ "moderate",
        de.facLoc == 0.3 ~ "strong",
        de.facLoc == 0.6 ~ "very strong",
      ),
      de.facLoc = factor(de.facLoc, levels = c("none", "small", "mild", "moderate", "strong", "very strong")),
      batch.facLoc = case_when(
        batch.facLoc == 0 ~ "none",
        batch.facLoc == 0.01 ~ "small",
        batch.facLoc == 0.05 ~ "mild",
        batch.facLoc == 0.1 ~ "moderate",
        batch.facLoc == 0.3 ~ "strong",
        batch.facLoc == 0.6 ~ "very strong",
      ),
      batch.facLoc = factor(batch.facLoc, levels = c("none", "small", "mild", "moderate", "strong", "very strong")),
      gene_selection = factor(gene_selection),
      integration_method = factor(integration_method)
    ) |>
    as_tibble()

head(metrics)
```



# Plot
## Heatmap integration score as a function of biol. variab. and batch effects
```{r}
for (m in unique(metrics$integration_method)) {
  for (i in unique(metrics$ngroup)) {
    for (j in unique(metrics$nbatch)) {
      p <- metrics %>%
        filter(ngroup == i, nbatch == j, integration_method == m) %>%
        ggplot(aes(x = batch.facLoc, y = de.facLoc)) +
        geom_tile(aes(fill = Total, color = Total)) +
        coord_equal() +
        labs(x = "Batch Effects", y = "Biological Variability", title = paste0(m, "-ngroup", i, "-nbatch", j, "-TotalScore"))
      print(p)
    }
  }
}
p <- metrics %>%
  ggplot(aes(x = batch.facLoc, y = de.facLoc)) +
  geom_tile(aes(fill = Total, color = Total)) +
  coord_equal() +
  scale_fill_viridis_c() +
  facet_wrap(~gene_selection+integration_method)
p

# p <- metrics %>% 
#   ggplot(aes(x = de.facLoc, y = Total))+
#   geom_point(aes(color = batch.facLoc, shape = gene_selection))+
#   facet_wrap(~gene_selection+integration_method)
# p

```

```{r}
for (m in unique(metrics$integration_method)) {
  for (g in unique(metrics$gene_selection)) {
    p <- metrics %>%
      filter(integration_method == m, gene_selection == g) %>%
      summarise(
        mean_score = mean(Total),
        median_score = median(Total),
        integration_method = m,
        de.facLoc = de.facLoc,
        batch.facLoc = batch.facLoc,
        .by = c(de.facLoc, batch.facLoc)
      ) %>%
      distinct() %>% 
      ggplot(aes(x = batch.facLoc, y = de.facLoc)) +
        geom_tile(aes(fill = mean_score)) +
        geom_text(aes(label = round(mean_score, 2)), color = "black", size = 4) +  # Add text to tiles
        scale_fill_gradient(low ="grey", high = "#64487A", limits=c(0,1))+
        coord_equal() +
        labs(x = "Batch Effects", y = "Biological Variability", title = paste0(m, " - ", g, " - mean Total Score"))
    print(p)
  }
}
```

```{r}
for (m in unique(metrics$integration_method)) {
  for (g in unique(metrics$gene_selection)) {
    p <- metrics %>%
      filter(integration_method == m, gene_selection == g) %>%
      summarise(
        mean_score = mean(Total),
        median_score = median(Total),
        integration_method = m,
        de.facLoc = de.facLoc,
        batch.facLoc = batch.facLoc,
        .by = c(de.facLoc, batch.facLoc)
      ) %>%
      distinct() %>% 
      ggplot(aes(x = batch.facLoc, y = de.facLoc)) +
        geom_tile(aes(fill = median_score)) +
        geom_text(aes(label = round(median_score, 2)), color = "black", size = 4) +  # Add text to tiles
        scale_fill_gradient(low ="grey", high = "#64487A", limits=c(0,1))+
        coord_equal() +
        labs(x = "Batch Effects", y = "Biological Variability", title = paste0(m, " - ", g, " - median Total Score"))
    print(p)
  }
}
```

## violin plots
### Total Score
```{r}
dodge <- position_dodge(width = 0.7)
for (m in unique(metrics$integration_method)) {
  for (i in unique(metrics$ngroup)) {
    for (j in unique(metrics$nbatch)) {
      p <- metrics %>%
        filter(ngroup == i, nbatch == j, integration_method == m) %>%
        ggplot(aes(x = as.factor(de.facLoc), y = Total)) +
        # geom_violin(aes(fill = gene_selection), position = dodge, width = 2)+
        geom_boxplot(width = 0.5, aes(fill = gene_selection), position = dodge) +
        geom_jitter(aes(fill = gene_selection), alpha = 0.7, position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7)) +
        # make manual color
        scale_fill_manual(values = c("angl" = "#64487A", "hvg" = "grey"))+
        theme_bw() +
        labs(x = "Biological Variability", y = "Integration Score", title = paste0(m, "-ngroup", i, "-nbatch", j )) +
        ylim(0, 1) +
        # increase font size of all text
        theme(
          text = element_text(size = 20, face = "bold"),
          aspect.ratio = 1,
          legend.position = c(0.8, 0.2)
        )

        print(p)
        ggsave(
            file.path(outplot_bm, paste0(m, "_ngroup", i, "_nbatch", j, "_TotalScore_violin.pdf")),
            width = 7,
            height = 7,
            plot = p
        )
    }
  }
}

```

```{r}
metrics %>%
  distinct(ngroup, nbatch)
dodge <- position_dodge(width = 0.7)

p <- metrics %>%
  filter(integration_method == "scvi") %>%
  ggplot(aes(x = as.factor(de.facLoc), y = Total)) +
  # geom_violin(aes(fill = gene_selection), position = dodge, width = 2)+
  geom_boxplot(width = 0.5, aes(fill = gene_selection), position = dodge) +
  geom_jitter(aes(fill = gene_selection), position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7)) +
  # make manual color
  # scale_color_brewer(palette = "Paired") +  # "Paired" is a good palette for qualitative factors
  theme_bw() +
  scale_fill_manual(values = c("angl" = "#64487A", "hvg" = "grey"))+
  labs(x = "Biological Variability", y = "Integration Score") +
  # increase font size of all text
  theme(
    text = element_text(size = 30, face = "bold"),
    aspect.ratio = 1,
    legend.position = c(0.9, 0.1)
  ) +
  facet_wrap(~ ngroup + nbatch, ncol = 5, labeller = "label_both")
p

metrics %>%
  distinct(ngroup, nbatch)
dodge <- position_dodge(width = 0.7)
p <- metrics %>%
  filter(integration_method == "scanorama") %>%
  ggplot(aes(x = as.factor(de.facLoc), y = Total)) +
  # geom_violin(aes(fill = gene_selection), position = dodge, width = 2)+
  geom_boxplot(width = 0.5, aes(fill = gene_selection), position = dodge) +
  geom_jitter(aes(fill = gene_selection), position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7)) +
  # make manual color
  # scale_color_brewer(palette = "Paired") +  # "Paired" is a good palette for qualitative factors
  theme_bw() +
  scale_fill_manual(values = c("angl" = "#64487A", "hvg" = "grey"))+
  labs(x = "Biological Variability", y = "Integration Score") +
  # increase font size of all text
  theme(
    text = element_text(size = 30, face = "bold"),
    aspect.ratio = 1,
    legend.position = c(0.9, 0.1)
  ) +
  facet_wrap(~ ngroup + nbatch, ncol = 5, labeller = "label_both")
p
```

### Bio.conservation
```{r}
dodge <- position_dodge(width = 0.7)
for (m in unique(metrics$integration_method)) {
  for (i in unique(metrics$ngroup)) {
    for (j in unique(metrics$nbatch)) {
      p <- metrics %>%
        filter(ngroup == i, nbatch == j, integration_method == m) %>%
        ggplot(aes(x = as.factor(de.facLoc), y = Bio.conservation)) +
        # geom_violin(aes(fill = gene_selection), position = dodge, width = 2)+
        geom_boxplot(width = 0.5, aes(fill = gene_selection), position = dodge) +
        geom_jitter(aes(fill = gene_selection), alpha = 0.7, position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7)) +
        # make manual color
        scale_fill_manual(values = c("angl" = "#64487A", "hvg" = "grey"))+
        theme_bw() +
        labs(x = "Biological Variability", y = "BioConservation Score", title = paste0(m, "-ngroup", i, "-nbatch", j )) +
        ylim(0, 1) +
        # increase font size of all text
        theme(
          text = element_text(size = 20, face = "bold"),
          aspect.ratio = 1,
          legend.position = c(0.8, 0.2)
        )

        print(p)
        ggsave(
            file.path(outplot_bm, paste0(m, "_ngroup", i, "_nbatch", j, "_BioConservation_violin.pdf")),
            width = 7,
            height = 7,
            plot = p
        )
    }
  }
}
```

### Batch.correction
```{r}
dodge <- position_dodge(width = 0.7)
for (m in unique(metrics$integration_method)) {
  for (i in unique(metrics$ngroup)) {
    for (j in unique(metrics$nbatch)) {
      p <- metrics %>%
        filter(ngroup == i, nbatch == j, integration_method == m) %>%
        ggplot(aes(x = as.factor(de.facLoc), y = Batch.correction)) +
        # geom_violin(aes(fill = gene_selection), position = dodge, width = 2)+
        geom_boxplot(width = 0.5, aes(fill = gene_selection), position = dodge) +
        geom_jitter(aes(fill = gene_selection), alpha = 0.7, position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7)) +
        # make manual color
        scale_fill_manual(values = c("angl" = "#64487A", "hvg" = "grey"))+
        theme_bw() +
        labs(x = "Biological Variability", y = "BatchCorrection score", title = paste0(m, "-ngroup", i, "-nbatch", j )) +
        ylim(0, 1) +
        # increase font size of all text
        theme(
          text = element_text(size = 20, face = "bold"),
          aspect.ratio = 1,
          legend.position = c(0.8, 0.2)
        )
      print(p)
      ggsave(
        file.path(outplot_bm, paste0(m, "_ngroup", i, "_nbatch", j, "_BatchCorrection_violin.pdf")),
        width = 7,
        height = 7,
        plot = p
      )
    }
  }
}
```

## low DE total integration score across increasing batch effect
```{r}
lowde <- metrics |>
    filter(de.facLoc == "moderate")

# scatter plot
lowde |>
    ggplot(aes(x = batch.facLoc, y = Total, fill = gene_selection)) +
    geom_boxplot() +
    theme_minimal() +
    facet_wrap(~integration_method)

lowde |>
    filter(nbatch == 5, ngroup == 2) |>
    ggplot(aes(x = batch.facLoc, y = Total, color = integration_method, shape = gene_selection)) +
    geom_point() +
    theme_minimal()
lowde |>
    filter(nbatch == 2, ngroup == 5) |>
    ggplot(aes(x = batch.facLoc, y = Total, color = integration_method, shape = gene_selection)) +
    geom_point() +
    theme_minimal()
```

# Statistical Testing
## angl vs hvg: paired Wilcoxon test on integration scores

One-sample Wilcoxon test on per-sample deltas (angl − hvg).
H0: median Δ = 0 vs H1: median Δ > 0 (anglemania outperforms HVG).
p-values are FDR-adjusted across integration methods.

```{r stat_overall}
diff_df <- metrics %>%
  filter(gene_selection %in% c("angl", "hvg")) %>%
  select(sample, integration_method, gene_selection, Total, Bio.conservation, Batch.correction) %>%
  pivot_wider(
    names_from  = gene_selection,
    values_from = c(Total, Bio.conservation, Batch.correction)
  ) %>%
  mutate(
    delta_Total = Total_angl             - Total_hvg,
    delta_Bio   = Bio.conservation_angl  - Bio.conservation_hvg,
    delta_Batch = Batch.correction_angl  - Batch.correction_hvg
  )

stat_results <- diff_df %>%
  group_by(integration_method) %>%
  summarise(
    n                  = n(),
    median_delta_Total = median(delta_Total, na.rm = TRUE),
    p_Total            = wilcox.test(delta_Total, mu = 0, alternative = "greater")$p.value,
    median_delta_Bio   = median(delta_Bio,   na.rm = TRUE),
    p_Bio              = wilcox.test(delta_Bio,   mu = 0, alternative = "greater")$p.value,
    median_delta_Batch = median(delta_Batch, na.rm = TRUE),
    p_Batch            = wilcox.test(delta_Batch, mu = 0, alternative = "greater")$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    p_Total_fdr  = p.adjust(p_Total,  method = "BH"),
    p_Bio_fdr    = p.adjust(p_Bio,    method = "BH"),
    p_Batch_fdr  = p.adjust(p_Batch,  method = "BH")
  )

stat_results
```

```{r stat_delta_boxplot}
diff_long <- diff_df %>%
  pivot_longer(
    cols      = starts_with("delta_"),
    names_to  = "score_type",
    values_to = "delta"
  ) %>%
  mutate(score_type = factor(
    dplyr::recode(score_type,
      delta_Total = "Total",
      delta_Bio   = "Bio conservation",
      delta_Batch = "Batch correction"
    ),
    levels = c("Total", "Bio conservation", "Batch correction")
  ))

p <- ggplot(diff_long, aes(integration_method, delta, fill = integration_method)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_boxplot(outlier.size = 0.5, alpha = 0.8) +
  stat_summary(fun = median, geom = "point", shape = 18, size = 3) +
  scale_fill_manual(values = c("scvi" = "#4E79A7", "scanorama" = "#F28E2B")) +
  facet_wrap(~score_type) +
  labs(
    title    = "Score difference: anglemania − HVG (paired by sample)",
    subtitle = "Positive = anglemania outperforms HVG; dashed line at 0",
    x = "Integration method", y = "Δ score (angl − hvg)", fill = NULL
  ) +
  theme_bw() +
  theme(text = element_text(size = 14), legend.position = "none")
p
ggsave(file.path(outplot_bm, "stat_delta_boxplot.pdf"), p, width = 10, height = 5)
```

## Per-condition significance (batch × bio effect heatmap)

Tests whether angl − hvg Total score delta is significantly > 0 within each
(batch.facLoc × de.facLoc) condition, across the ngroup × nbatch replicates.

```{r stat_per_condition}
stat_by_condition <- diff_df %>%
  left_join(
    metrics %>% filter(gene_selection == "angl") %>%
      select(sample, integration_method, batch.facLoc, de.facLoc),
    by = c("sample", "integration_method")
  ) %>%
  group_by(integration_method, batch.facLoc, de.facLoc) %>%
  summarise(
    n            = n(),
    median_delta = median(delta_Total, na.rm = TRUE),
    p_value      = tryCatch(
      wilcox.test(delta_Total, mu = 0, alternative = "greater")$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) %>%
  group_by(integration_method) %>%
  mutate(
    p_fdr = p.adjust(p_value, method = "BH"),
    sig   = case_when(
      p_fdr < 0.001 ~ "***", p_fdr < 0.01 ~ "**",
      p_fdr < 0.05  ~ "*",   TRUE ~ "ns"
    )
  ) %>%
  ungroup()

for (m in unique(stat_by_condition$integration_method)) {
  p <- stat_by_condition %>%
    filter(integration_method == m) %>%
    ggplot(aes(x = batch.facLoc, y = de.facLoc, fill = median_delta)) +
    geom_tile(color = "white") +
    geom_text(aes(label = sig), size = 5) +
    scale_fill_gradient2(
      low = "blue", mid = "white", high = "#64487A",
      midpoint = 0, limits = c(-0.2, 0.2), oob = scales::squish
    ) +
    coord_equal() +
    labs(
      title    = paste0(m, ": angl vs hvg Total score Δ"),
      subtitle = "Fill = median Δ (angl − hvg); label = BH-corrected significance",
      x = "Batch effect", y = "Biological variability", fill = "Δ Total"
    ) +
    theme_bw() +
    theme(text = element_text(size = 14))
  print(p)
  ggsave(
    file.path(outplot_bm, paste0(m, "_stat_significance_heatmap.pdf")),
    p, width = 7, height = 6
  )
}
```

# session Info
```{r}
sessionInfo()
```