---
title: "BTVR Sanity Check: Does high BTVR → better integration?"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../output/")})
output:
    html_document:
      toc: yes
      toc_float:
        collapsed: no
        smooth_scroll: no
      df_print: paged
---

**Hypothesis**: Genes with high ground-truth BTVR (Between-to-Total Variance
Ratio) — high biological signal, low batch noise — should produce better batch
integration than genes with low BTVR. This notebook validates that hypothesis
by comparing integration metrics for `top_btvr` vs `bottom_btvr` gene
selections against the standard `hvg` and `angl` selections.

**Pipeline**: Results produced by running Snakemake with
`config_files/config_btvr_sanity_check.yml` on a subset of the facScale0.1
simulations (nbatch=4, ngroup=4; 36 samples covering a 6×6 grid of
batch.facLoc × de.facLoc combinations).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_chunk$set(fig.height = 10)
knitr::opts_chunk$set(fig.width = 14)
options(Seurat.object.assay.version = "v5")
```

# Setup

```{r libraries}
.libPaths(.Library)
suppressPackageStartupMessages({
    library(tidyverse)
    library(scales)
    library(rhdf5)
    library(glue)
    library(patchwork)
    library(yaml)
})
source("/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/scripts/ggplot_utils.R")
```

```{r paths}
base_dir        <- "/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark"
cfg             <- read_yaml(file.path(base_dir, "scripts/config_files/config_btvr_sanity_check.yml"))
samplesheet_src <- "/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/samplesheet_facScale0.1.tsv"
samplesheet_out <- "/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/samplesheet_btvr_sanity_check.tsv"
combined_metrics_path <- file.path(cfg$output_dir, cfg$out_metrics,
                                   paste0(cfg$name, "_combined_metrics.tsv"))
outplot <- file.path(base_dir, "output/plots/btvr_sanity_check/")
if (!dir.exists(outplot)) dir.create(outplot, recursive = TRUE)
```

# Step 0: Create filtered samplesheet (if not yet done)

This chunk creates the filtered samplesheet needed to run the Snakemake
pipeline. Run this **before** launching Snakemake.

```{r create_samplesheet}
if (!file.exists(samplesheet_out)) {
    message("Creating filtered samplesheet: nbatch=4, ngroup=4 ...")
    ss <- read.delim(samplesheet_src)
    sub <- ss[ss$nbatch == 4 & ss$ngroup == 4, ]
    message(nrow(sub), " samples selected")
    write.table(sub, samplesheet_out, sep = "\t", row.names = FALSE, quote = FALSE)
    message("Written to: ", samplesheet_out)
} else {
    message("Samplesheet already exists: ", samplesheet_out)
}
samplesheet <- read.delim(samplesheet_out)
message(nrow(samplesheet), " samples in samplesheet")
head(samplesheet)
```

After running this chunk, launch the pipeline from `scripts/`:

```bash
cd scripts/
snakemake -s Snakefile.smk --configfile config_files/config_btvr_sanity_check.yml -n
snakemake -s Snakefile.smk --configfile config_files/config_btvr_sanity_check.yml --cores 8
```

---

# Part 1: BTVR property of selected gene sets

Before looking at integration results, we verify that the gene sets actually
differ in their BTVR distributions as expected. We load one representative
h5ad, compute BTVR from the ground-truth splatter columns, and compare the
distributions of the four gene sets.

```{r btvr_functions}
get_BTVR <- function(var_df) {
    bio_cols  <- select(var_df, starts_with("DEFacGroup"))
    tech_cols <- select(var_df, starts_with("BatchFacBatch"))
    raw_BioVar  <- apply(bio_cols,  1, var)
    raw_TechVar <- apply(tech_cols, 1, var)
    # rescaled log-ratio (from compare_gene_selections.Rmd)
    BTVR <- rescale(log((raw_BioVar + 1e-15) / (raw_TechVar + 1e-15)))
    tibble(
        gene        = rownames(var_df),
        BTVR        = BTVR,
        raw_BioVar  = raw_BioVar,
        raw_TechVar = raw_TechVar
    )
}

get_var_from_h5ad <- function(h5ad_path) {
    df <- h5read(h5ad_path, "/var") |> as.data.frame() |> column_to_rownames("_index")
    h5closeAll()
    df
}
```

```{r pick_example_sample}
# Use a sample with moderate batch and bio effects as the example
samplesheet <- read.delim(samplesheet_out)
example_row <- samplesheet |>
    filter(batch.facLoc == 0.3, de.facLoc == 0.3) |>
    slice(1)

if (nrow(example_row) == 0) {
    example_row <- samplesheet |> slice(1)
    message("Fallback: using first row of samplesheet")
}

example_h5ad    <- example_row$file_path
example_sample  <- example_row$sample_name
message("Example sample: ", example_sample)
message("h5ad path: ", example_h5ad)
```

```{r compute_btvr_example}
var_df   <- get_var_from_h5ad(example_h5ad)
btvr_all <- get_BTVR(var_df)

# Retrieve the gene lists produced by the pipeline for this sample
preprocessed_dir <- file.path(cfg$output_dir, cfg$out_preprocessed)
gene_lists <- list()
for (gs in cfg$gene_selection) {
    tsv_path <- file.path(preprocessed_dir, glue("{example_sample}_{gs}.tsv"))
    if (file.exists(tsv_path)) {
        gene_lists[[gs]] <- read.delim(tsv_path)$hgnc_symbol
    } else {
        message("Gene list not found (pipeline not yet run?): ", tsv_path)
        gene_lists[[gs]] <- character(0)
    }
}
```

```{r btvr_violin, fig.height=6}
btvr_labeled <- map_dfr(names(gene_lists), function(gs) {
    btvr_all |>
        filter(gene %in% gene_lists[[gs]]) |>
        mutate(gene_selection = gs)
}) |>
    mutate(gene_selection = factor(gene_selection,
           levels = c("top_btvr", "bottom_btvr", "hvg", "angl")))

gs_labels <- c(
    top_btvr    = "Top 2000 BTVR",
    bottom_btvr = "Bottom 2000 BTVR",
    hvg         = "HVG",
    angl        = "anglemania"
)
gs_colors <- c(
    top_btvr    = "#2196F3",
    bottom_btvr = "#F44336",
    hvg         = "#FF9800",
    angl        = "#64487A"
)

p_violin <- ggplot(btvr_labeled, aes(gene_selection, BTVR, fill = gene_selection)) +
    geom_violin(trim = FALSE, alpha = 0.8) +
    geom_boxplot(width = 0.12, fill = "white", outlier.size = 0.3) +
    scale_fill_manual(values = gs_colors, labels = gs_labels) +
    scale_x_discrete(labels = gs_labels) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
        title = glue("BTVR distribution of selected genes\n({example_sample})"),
        x     = "Gene selection method",
        y     = "BTVR (rescaled, higher = more biological signal)",
        fill  = NULL
    ) +
    theme_big_text() +
    theme(legend.position = "none")

p_violin
ggsave(file.path(outplot, glue("{today()}_btvr_violin_{example_sample}.pdf")),
       p_violin, width = 8, height = 5)
```

```{r btvr_density, fig.height=5}
p_density <- ggplot(btvr_labeled, aes(BTVR, color = gene_selection, fill = gene_selection)) +
    geom_density(alpha = 0.15, linewidth = 0.8) +
    scale_color_manual(values = gs_colors, labels = gs_labels) +
    scale_fill_manual(values  = gs_colors, labels = gs_labels) +
    labs(
        title = "BTVR density per gene selection",
        x     = "BTVR (rescaled)",
        y     = "Density",
        color = NULL, fill = NULL
    ) +
    theme_big_text()

p_density
ggsave(file.path(outplot, glue("{today()}_btvr_density_{example_sample}.pdf")),
       p_density, width = 8, height = 4)
```

---

# Part 2: Integration metrics

```{r check_metrics_file}
if (!file.exists(combined_metrics_path)) {
    stop(
        "Combined metrics file not found:\n  ", combined_metrics_path,
        "\n\nRun the Snakemake pipeline first:\n",
        "  cd scripts/\n",
        "  snakemake -s Snakefile.smk --configfile config_files/config_btvr_sanity_check.yml --cores 8"
    )
}
message("Loading metrics from: ", combined_metrics_path)
```

```{r load_metrics}
metrics_raw <- read.delim(combined_metrics_path)

metrics <- metrics_raw |>
    left_join(
        samplesheet |> select(sample_name, batch.facLoc, de.facLoc, nbatch, ngroup),
        by = join_by(sample == sample_name)
    ) |>
    mutate(
        gene_selection     = factor(gene_selection,
            levels = c("top_btvr", "bottom_btvr", "hvg", "angl")),
        integration_method = factor(integration_method,
            levels = c("scvi", "scanorama")),
        de.facLoc    = factor(de.facLoc),
        batch.facLoc = factor(batch.facLoc)
    )

message(nrow(metrics), " rows loaded")
head(metrics)
```

## 2.1 Overall summary: boxplot across all conditions

The key sanity check: `top_btvr` should score highest, `bottom_btvr` lowest.
`angl` and `hvg` should fall somewhere in between (ideally `angl` ≥ `hvg`).

```{r boxplot_overall, fig.height=6}
p_box <- metrics |>
    ggplot(aes(gene_selection, Total, fill = gene_selection)) +
    geom_boxplot(outlier.size = 0.5, alpha = 0.8) +
    stat_summary(fun = mean, geom = "point", shape = 4, size = 3, color = "black") +
    scale_fill_manual(values = gs_colors, labels = gs_labels) +
    scale_x_discrete(labels = gs_labels) +
    scale_y_continuous(limits = c(0, 1)) +
    facet_wrap(~integration_method) +
    labs(
        title = "Integration Total score by gene selection (all conditions)",
        x     = NULL,
        y     = "Total score (higher = better)",
        fill  = NULL
    ) +
    theme_big_text() +
    theme(
        legend.position  = "none",
        axis.text.x      = element_text(angle = 25, hjust = 1)
    )

p_box
ggsave(file.path(outplot, glue("{today()}_total_score_boxplot.pdf")),
       p_box, width = 10, height = 6)
```

```{r boxplot_bio_batch, fig.height=8}
metrics_long <- metrics |>
    select(gene_selection, integration_method, `Bio conservation`, `Batch correction`, Total) |>
    pivot_longer(c(`Bio conservation`, `Batch correction`, Total),
                 names_to = "metric_type", values_to = "score") |>
    mutate(metric_type = factor(metric_type,
           levels = c("Bio conservation", "Batch correction", "Total")))

p_box2 <- ggplot(metrics_long, aes(gene_selection, score, fill = gene_selection)) +
    geom_boxplot(outlier.size = 0.4, alpha = 0.8) +
    stat_summary(fun = mean, geom = "point", shape = 4, size = 2) +
    scale_fill_manual(values = gs_colors, labels = gs_labels) +
    scale_x_discrete(labels = gs_labels) +
    scale_y_continuous(limits = c(0, 1)) +
    facet_grid(metric_type ~ integration_method) +
    labs(
        title = "Integration scores by gene selection",
        x = NULL, y = "Score", fill = NULL
    ) +
    theme_big_text() +
    theme(
        legend.position = "none",
        axis.text.x     = element_text(angle = 25, hjust = 1)
    )

p_box2
ggsave(file.path(outplot, glue("{today()}_scores_by_type_boxplot.pdf")),
       p_box2, width = 12, height = 9)
```

## 2.2 Heatmaps: Total score as function of batch × bio effects

Using the same heatmap style as `bm_simulated_visualization.Rmd`.

```{r heatmap_total, fig.height=8}
heatmap_labels <- c(
    top_btvr    = "Top BTVR",
    bottom_btvr = "Bottom BTVR",
    hvg         = "HVG",
    angl        = "anglemania"
)

for (m in levels(metrics$integration_method)) {
    p <- metrics |>
        filter(integration_method == m) |>
        draw_heatmap2(
            df     = _,
            score  = "Total",
            fill   = "mean",
            labels = heatmap_labels,
            scale_fill_limits = c(0, 1)
        ) +
        labs(title = glue("{m} — mean Total score by batch × bio effect"))
    print(p)
    ggsave(
        file.path(outplot, glue("{today()}_heatmap_total_{m}.pdf")),
        p, width = 16, height = 6
    )
}
```

```{r heatmap_bio, fig.height=8}
for (m in levels(metrics$integration_method)) {
    p <- metrics |>
        filter(integration_method == m) |>
        draw_heatmap2(
            df     = _,
            score  = "Bio conservation",
            fill   = "mean",
            labels = heatmap_labels,
            scale_fill_limits = c(0, 1)
        ) +
        labs(title = glue("{m} — mean Bio conservation by batch × bio effect"))
    print(p)
    ggsave(
        file.path(outplot, glue("{today()}_heatmap_biocons_{m}.pdf")),
        p, width = 16, height = 6
    )
}
```

```{r heatmap_batch, fig.height=8}
for (m in levels(metrics$integration_method)) {
    p <- metrics |>
        filter(integration_method == m) |>
        draw_heatmap2(
            df     = _,
            score  = "Batch correction",
            fill   = "mean",
            labels = heatmap_labels,
            scale_fill_limits = c(0, 1)
        ) +
        labs(title = glue("{m} — mean Batch correction by batch × bio effect"))
    print(p)
    ggsave(
        file.path(outplot, glue("{today()}_heatmap_batchcorr_{m}.pdf")),
        p, width = 16, height = 6
    )
}
```

## 2.3 Top BTVR vs Bottom BTVR: direct comparison

The core sanity check — how much does top_btvr beat bottom_btvr?

```{r delta_btvr, fig.height=6}
delta_df <- metrics |>
    filter(gene_selection %in% c("top_btvr", "bottom_btvr")) |>
    select(sample, integration_method, gene_selection, Total,
           `Bio conservation`, `Batch correction`,
           batch.facLoc, de.facLoc) |>
    pivot_longer(c(Total, `Bio conservation`, `Batch correction`),
                 names_to = "metric", values_to = "score") |>
    pivot_wider(names_from = gene_selection, values_from = score) |>
    mutate(delta = top_btvr - bottom_btvr)

p_delta <- ggplot(delta_df, aes(metric, delta, fill = metric)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    geom_boxplot(outlier.size = 0.4, alpha = 0.8) +
    stat_summary(fun = mean, geom = "point", shape = 4, size = 3) +
    facet_wrap(~integration_method) +
    labs(
        title = "Score difference: Top BTVR minus Bottom BTVR",
        subtitle = "Positive values confirm the hypothesis",
        x = NULL, y = "Δ score (top − bottom BTVR)",
        fill = NULL
    ) +
    theme_big_text() +
    theme(legend.position = "none")

p_delta
ggsave(file.path(outplot, glue("{today()}_delta_top_vs_bottom_btvr.pdf")),
       p_delta, width = 10, height = 6)
```

## 2.4 Where does anglemania fall relative to BTVR bounds?

```{r rank_comparison, fig.height=6}
rank_df <- metrics |>
    select(sample, integration_method, gene_selection, Total) |>
    mutate(gene_selection = factor(gene_selection,
           levels = c("top_btvr", "angl", "hvg", "bottom_btvr"))) |>
    group_by(sample, integration_method) |>
    mutate(rank = rank(-Total, ties.method = "average")) |>
    ungroup()

p_rank <- ggplot(rank_df, aes(gene_selection, rank, fill = gene_selection)) +
    geom_boxplot(outlier.size = 0.4, alpha = 0.8) +
    stat_summary(fun = mean, geom = "point", shape = 4, size = 3) +
    scale_fill_manual(values = gs_colors) +
    scale_x_discrete(labels = heatmap_labels) +
    scale_y_reverse(breaks = 1:4) +
    facet_wrap(~integration_method) +
    labs(
        title = "Rank of Total score per sample (1 = best)",
        x = NULL, y = "Rank (lower = better)", fill = NULL
    ) +
    theme_big_text() +
    theme(
        legend.position = "none",
        axis.text.x     = element_text(angle = 20, hjust = 1)
    )

p_rank
ggsave(file.path(outplot, glue("{today()}_rank_comparison.pdf")),
       p_rank, width = 10, height = 6)
```

# Session info

```{r session_info}
sessionInfo()
```
