---
title: "BTVR Sanity Check: Does high BTVR → better integration?"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../output/")})
output:
    html_document:
      toc: yes
      toc_float:
        collapsed: no
        smooth_scroll: no
      df_print: paged
---

**Hypothesis**: Genes with high ground-truth BTVR (Between-to-Total Variance
Ratio) — high biological signal, low batch noise — should produce better batch
integration than genes with low BTVR. This notebook validates that hypothesis
by comparing integration metrics for `topbtvr` vs `bottombtvr` gene
selections against the standard `hvg` and `angl` selections.

**Pipeline**: Results produced by running Snakemake with
`config_files/config_btvr_sanity_check.yml` on a subset of the facScale0.1
simulations (nbatch=4, ngroup=4; 36 samples covering a 6×6 grid of
batch.facLoc × de.facLoc combinations).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_chunk$set(fig.height = 10)
knitr::opts_chunk$set(fig.width = 14)
options(Seurat.object.assay.version = "v5")
```

# Setup

```{r libraries}
.libPaths(.Library)
suppressPackageStartupMessages({
    library(tidyverse)
    library(scales)
    library(rhdf5)
    library(glue)
    library(patchwork)
    library(yaml)
})
source("/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/scripts/ggplot_utils.R")
```

```{r paths}
base_dir        <- "/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark"
cfg             <- read_yaml(file.path(base_dir, "scripts/config_files/config_btvr_sanity_check.yml"))
samplesheet_src <- "/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/samplesheet_facScale0.1.tsv"
samplesheet_out <- "/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/samplesheet_btvr_sanity_check.tsv"
combined_metrics_path <- file.path(cfg$output_dir, cfg$out_metrics,
                                   paste0(cfg$name, "_combined_metrics.tsv"))
outplot <- file.path(base_dir, "output/plots/btvr_sanity_check/")
if (!dir.exists(outplot)) dir.create(outplot, recursive = TRUE)
```


---

# Part 1: BTVR property of selected gene sets

Before looking at integration results, we verify that the gene sets actually
differ in their BTVR distributions as expected. We load one representative
h5ad, compute BTVR from the ground-truth splatter columns, and compare the
distributions of the four gene sets.

```{r btvr_functions}
get_BTVR <- function(var_df) {
    bio_cols  <- select(var_df, starts_with("DEFacGroup"))
    tech_cols <- select(var_df, starts_with("BatchFacBatch"))
    raw_BioVar  <- apply(bio_cols,  1, var)
    raw_TechVar <- apply(tech_cols, 1, var)
    # rescaled log-ratio (from compare_gene_selections.Rmd)
    BTVR <- rescale(log((raw_BioVar + 1e-15) / (raw_TechVar + 1e-15)))
    tibble(
        gene        = rownames(var_df),
        BTVR        = BTVR,
        raw_BioVar  = raw_BioVar,
        raw_TechVar = raw_TechVar
    )
}

get_var_from_h5ad <- function(h5ad_path) {
    df <- h5read(h5ad_path, "/var") |>
        as.data.frame() |>
        column_to_rownames("X_index")
    h5closeAll()
    df
}
```

```{r pick_example_sample}
# Use a sample with moderate batch and bio effects as the example
samplesheet <- read.delim(samplesheet_out)
example_row <- samplesheet |>
    slice_head(n = 1)

if (nrow(example_row) == 0) {
    example_row <- samplesheet |> slice_head(n = 1)
    message("Fallback: using first row of samplesheet")
}

example_h5ad    <- example_row$file_path
example_sample  <- example_row$sample_name
message("Example sample: ", example_sample)
message("h5ad path: ", example_h5ad)
```

```{r compute_btvr_example}
var_df   <- get_var_from_h5ad(example_h5ad)
btvr_all <- get_BTVR(var_df)

# Retrieve the gene lists produced by the pipeline for this sample
preprocessed_dir <- file.path(cfg$output_dir, cfg$out_preprocessed)
gene_lists <- list()
for (gs in cfg$gene_selection) {
    tsv_path <- file.path(preprocessed_dir, glue("{example_sample}_{gs}.tsv"))
    if (file.exists(tsv_path)) {
        gene_lists[[gs]] <- read.delim(tsv_path)$hgnc_symbol
    } else {
        message("Gene list not found (pipeline not yet run?): ", tsv_path)
        gene_lists[[gs]] <- character(0)
    }
}
```

```{r btvr_violin, fig.height=6}
btvr_labeled <- map_dfr(names(gene_lists), function(gs) {
    btvr_all |>
        filter(gene %in% gene_lists[[gs]]) |>
        mutate(gene_selection = gs)
}) |>
    mutate(gene_selection = factor(gene_selection,
           levels = c("topbtvr", "bottombtvr", "hvg", "angl")))

gs_labels <- c(
    topbtvr    = "Top 2000 BTVR",
    bottombtvr = "Bottom 2000 BTVR",
    hvg         = "HVG",
    angl        = "anglemania"
)
gs_colors <- c(
    topbtvr    = "#2196F3",
    bottombtvr = "#F44336",
    hvg         = "#FF9800",
    angl        = "#64487A"
)

p_violin <- ggplot(btvr_labeled, aes(gene_selection, BTVR, fill = gene_selection)) +
    geom_violin(trim = FALSE, alpha = 0.8) +
    geom_boxplot(width = 0.12, fill = "white", outlier.size = 0.3) +
    scale_fill_manual(values = gs_colors, labels = gs_labels) +
    scale_x_discrete(labels = gs_labels) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
        title = glue("BTVR distribution of selected genes\n({example_sample})"),
        x     = "Gene selection method",
        y     = "BTVR (rescaled, higher = more biological signal)",
        fill  = NULL
    ) +
    theme_big_text() +
    theme(legend.position = "none")

p_violin
ggsave(file.path(outplot, glue("{today()}_btvr_violin_{example_sample}.pdf")),
       p_violin, width = 8, height = 5)
```

```{r btvr_density, fig.height=5}
p_density <- ggplot(btvr_labeled, aes(BTVR, color = gene_selection, fill = gene_selection)) +
    geom_density(alpha = 0.15, linewidth = 0.8) +
    scale_color_manual(values = gs_colors, labels = gs_labels) +
    scale_fill_manual(values  = gs_colors, labels = gs_labels) +
    labs(
        title = "BTVR density per gene selection",
        x     = "BTVR (rescaled)",
        y     = "Density",
        color = NULL, fill = NULL
    ) +
    theme_big_text()

p_density
ggsave(file.path(outplot, glue("{today()}_btvr_density_{example_sample}.pdf")),
       p_density, width = 8, height = 4)
```

---

# Part 2: Integration metrics

```{r check_metrics_file}
if (!file.exists(combined_metrics_path)) {
    stop(
        "Combined metrics file not found:\n  ", combined_metrics_path,
        "\n\nRun the Snakemake pipeline first:\n",
        "  cd scripts/\n",
        "  snakemake -s Snakefile.smk --configfile config_files/config_btvr_sanity_check.yml --cores 8"
    )
}
message("Loading metrics from: ", combined_metrics_path)
```

```{r load_metrics}
metrics_raw <- read.delim(combined_metrics_path)

metrics <- metrics_raw |>
    left_join(
        samplesheet |> select(sample_name, batch.facLoc, de.facLoc, nbatch, ngroup),
        by = join_by(sample == sample_name)
    ) |>
    mutate(
        gene_selection     = factor(gene_selection,
            levels = c("topbtvr", "bottombtvr", "hvg", "angl")),
        integration_method = factor(integration_method,
            levels = c("scvi", "scanorama")),
        de.facLoc    = factor(de.facLoc),
        batch.facLoc = factor(batch.facLoc)
    )

message(nrow(metrics), " rows loaded")
head(metrics)
```

## 2.1 Overall summary: boxplot across all conditions

The key sanity check: `topbtvr` should score highest, `bottombtvr` lowest.
`angl` and `hvg` should fall somewhere in between (ideally `angl` ≥ `hvg`).

```{r boxplot_overall, fig.height=6}
p_box <- metrics |>
    ggplot(aes(gene_selection, Total, fill = gene_selection)) +
    geom_boxplot(outlier.size = 0.5, alpha = 0.8) +
    stat_summary(fun = mean, geom = "point", shape = 4, size = 3, color = "black") +
    scale_fill_manual(values = gs_colors, labels = gs_labels) +
    scale_x_discrete(labels = gs_labels) +
    scale_y_continuous(limits = c(0, 1)) +
    facet_wrap(~integration_method) +
    labs(
        title = "Integration Total score by gene selection (all conditions)",
        x     = NULL,
        y     = "Total score (higher = better)",
        fill  = NULL
    ) +
    theme_big_text() +
    theme(
        legend.position  = "none",
        axis.text.x      = element_text(angle = 25, hjust = 1)
    )

p_box
ggsave(file.path(outplot, glue("{today()}_total_score_boxplot.pdf")),
       p_box, width = 10, height = 6)
```

```{r boxplot_bio_batch, fig.height=8}
metrics_long <- metrics |>
    select(gene_selection, integration_method, Bio.conservation, Batch.correction, Total) |>
    pivot_longer(c(Bio.conservation, Batch.correction, Total),
                 names_to = "metric_type", values_to = "score") |>
    mutate(metric_type = factor(metric_type,
           levels = c("Bio.conservation", "Batch.correction", "Total")))

p_box2 <- ggplot(metrics_long, aes(gene_selection, score, fill = gene_selection)) +
    geom_boxplot(outlier.size = 0.4, alpha = 0.8) +
    stat_summary(fun = mean, geom = "point", shape = 4, size = 2) +
    scale_fill_manual(values = gs_colors, labels = gs_labels) +
    scale_x_discrete(labels = gs_labels) +
    scale_y_continuous(limits = c(0, 1)) +
    facet_grid(metric_type ~ integration_method) +
    labs(
        title = "Integration scores by gene selection",
        x = NULL, y = "Score", fill = NULL
    ) +
    theme_big_text() +
    theme(
        legend.position = "none",
        axis.text.x     = element_text(angle = 25, hjust = 1)
    )

p_box2
ggsave(file.path(outplot, glue("{today()}_scores_by_type_boxplot.pdf")),
       p_box2, width = 12, height = 9)
```

## 2.2 Heatmaps: Total score as function of batch × bio effects

Using the same heatmap style as `bm_simulated_visualization.Rmd`.

```{r heatmap_total, fig.height=8}
heatmap_labels <- c(
    topbtvr    = "Top BTVR",
    bottombtvr = "Bottom BTVR",
    hvg         = "HVG",
    angl        = "anglemania"
)

for (m in levels(metrics$integration_method)) {
    p <- metrics |>
        filter(integration_method == m) |>
        draw_heatmap2(
            df     = _,
            score  = "Total",
            fill   = "mean",
            labels = heatmap_labels,
            scale_fill_limits = c(0, 1)
        ) +
        labs(title = glue("{m} — mean Total score by batch × bio effect"))
    print(p)
    ggsave(
        file.path(outplot, glue("{today()}_heatmap_total_{m}.pdf")),
        p, width = 16, height = 6
    )
}
```

```{r heatmap_bio, fig.height=8}
for (m in levels(metrics$integration_method)) {
    p <- metrics |>
        filter(integration_method == m) |>
        draw_heatmap2(
            df     = _,
            score  = "Bio.conservation",
            fill   = "mean",
            labels = heatmap_labels,
            scale_fill_limits = c(0, 1)
        ) +
        labs(title = glue("{m} — mean Bio conservation by batch × bio effect"))
    print(p)
    ggsave(
        file.path(outplot, glue("{today()}_heatmap_biocons_{m}.pdf")),
        p, width = 16, height = 6
    )
}
```

```{r heatmap_batch, fig.height=8}
for (m in levels(metrics$integration_method)) {
    p <- metrics |>
        filter(integration_method == m) |>
        draw_heatmap2(
            df     = _,
            score  = "Batch.correction",
            fill   = "mean",
            labels = heatmap_labels,
            scale_fill_limits = c(0, 1)
        ) +
        labs(title = glue("{m} — mean Batch correction by batch × bio effect"))
    print(p)
    ggsave(
        file.path(outplot, glue("{today()}_heatmap_batchcorr_{m}.pdf")),
        p, width = 16, height = 6
    )
}
```

## 2.3 Top BTVR vs Bottom BTVR: direct comparison

The core sanity check — how much does topbtvr beat bottombtvr?

```{r delta_btvr, fig.height=6}
delta_df <- metrics |>
    filter(gene_selection %in% c("topbtvr", "bottombtvr")) |>
    select(sample, integration_method, gene_selection, Total,
           Bio.conservation, Batch.correction,
           batch.facLoc, de.facLoc) |>
    pivot_longer(c(Total, Bio.conservation, Batch.correction),
                 names_to = "metric", values_to = "score") |>
    pivot_wider(names_from = gene_selection, values_from = score) |>
    mutate(delta = topbtvr - bottombtvr)

p_delta <- ggplot(delta_df, aes(metric, delta, fill = metric)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    geom_boxplot(outlier.size = 0.4, alpha = 0.8) +
    stat_summary(fun = mean, geom = "point", shape = 4, size = 3) +
    facet_wrap(~integration_method) +
    labs(
        title = "Score difference: Top BTVR minus Bottom BTVR",
        subtitle = "Positive values confirm the hypothesis",
        x = NULL, y = "Δ score (top − bottom BTVR)",
        fill = NULL
    ) +
    theme_big_text() +
    theme(legend.position = "none")

p_delta
ggsave(file.path(outplot, glue("{today()}_delta_top_vs_bottombtvr.pdf")),
       p_delta, width = 10, height = 6)
```

## 2.4 Where does anglemania fall relative to BTVR bounds?

```{r rank_comparison, fig.height=6}
rank_df <- metrics |>
    select(sample, integration_method, gene_selection, Total) |>
    mutate(gene_selection = factor(gene_selection,
           levels = c("topbtvr", "angl", "hvg", "bottombtvr"))) |>
    group_by(sample, integration_method) |>
    mutate(rank = rank(-Total, ties.method = "average")) |>
    ungroup()

p_rank <- ggplot(rank_df, aes(gene_selection, rank, fill = gene_selection)) +
    geom_boxplot(outlier.size = 0.4, alpha = 0.8) +
    stat_summary(fun = mean, geom = "point", shape = 4, size = 3) +
    scale_fill_manual(values = gs_colors) +
    scale_x_discrete(labels = heatmap_labels) +
    scale_y_reverse(breaks = 1:4) +
    facet_wrap(~integration_method) +
    labs(
        title = "Rank of Total score per sample (1 = best)",
        x = NULL, y = "Rank (lower = better)", fill = NULL
    ) +
    theme_big_text() +
    theme(
        legend.position = "none",
        axis.text.x     = element_text(angle = 20, hjust = 1)
    )

p_rank
ggsave(file.path(outplot, glue("{today()}_rank_comparison.pdf")),
       p_rank, width = 10, height = 6)
```

# Statistical Testing

## 3.1 BTVR: angl vs hvg across all samples (paired per-sample test)

Compute the median BTVR of selected genes for each sample, then test whether
anglemania genes have higher median BTVR than HVG genes across all samples.
H0: median Δ = 0 vs H1: median Δ > 0 (angl selects higher-BTVR genes).

```{r stat_btvr_all_samples}
# Load gene lists and BTVR values for all samples
all_btvr_list <- lapply(seq_len(nrow(samplesheet)), function(i) {
  row         <- samplesheet[i, ]
  h5ad_path   <- row$file_path
  sample_name <- row$sample_name

  var_df   <- tryCatch(get_var_from_h5ad(h5ad_path), error = function(e) NULL)
  if (is.null(var_df)) return(NULL)
  btvr_all <- get_BTVR(var_df)

  result <- lapply(c("angl", "hvg"), function(gs) {
    tsv_path <- file.path(preprocessed_dir, glue("{sample_name}_{gs}.tsv"))
    if (!file.exists(tsv_path)) return(NULL)
    genes <- read.delim(tsv_path)$hgnc_symbol
    btvr_all %>%
      filter(gene %in% genes) %>%
      summarise(
        median_BTVR = median(BTVR, na.rm = TRUE),
        mean_BTVR   = mean(BTVR,   na.rm = TRUE)
      ) %>%
      mutate(
        gene_selection = gs,
        sample         = sample_name,
        batch.facLoc   = row$batch.facLoc,
        de.facLoc      = row$de.facLoc
      )
  })
  bind_rows(result)
})

btvr_per_sample <- bind_rows(all_btvr_list)

diff_btvr <- btvr_per_sample %>%
  filter(gene_selection %in% c("angl", "hvg")) %>%
  pivot_wider(names_from = gene_selection, values_from = c(median_BTVR, mean_BTVR)) %>%
  mutate(delta_median = median_BTVR_angl - median_BTVR_hvg)

# Overall Wilcoxon (all conditions pooled)
wt_btvr_overall <- wilcox.test(diff_btvr$delta_median, mu = 0, alternative = "greater")
cat(sprintf(
  "Overall BTVR: median Δ = %.4f, p = %.4g, n = %d samples\n",
  median(diff_btvr$delta_median, na.rm = TRUE),
  wt_btvr_overall$p.value,
  nrow(diff_btvr)
))
```

```{r stat_btvr_delta_boxplot, fig.height=5, fig.width=5}
p_btvr_delta <- ggplot(diff_btvr, aes(x = "angl − hvg", y = delta_median)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_boxplot(width = 0.4, fill = "#64487A", alpha = 0.7, outlier.size = 0.5) +
  stat_summary(fun = median, geom = "point", shape = 18, size = 4, color = "white") +
  labs(
    title    = "BTVR difference: anglemania − HVG (per sample)",
    subtitle = sprintf("Wilcoxon p = %.4g", wt_btvr_overall$p.value),
    x = NULL, y = "Δ median BTVR (angl − hvg)"
  ) +
  theme_bw() +
  theme(text = element_text(size = 14), axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p_btvr_delta
ggsave(file.path(outplot, glue("{today()}_stat_btvr_delta_boxplot.pdf")),
       p_btvr_delta, width = 5, height = 5)
```

## 3.2 Integration scores: angl vs hvg (paired Wilcoxon)

```{r stat_integration_angl_hvg}
diff_scores <- metrics %>%
  filter(gene_selection %in% c("angl", "hvg")) %>%
  select(sample, integration_method, gene_selection, Total, Bio.conservation, Batch.correction) %>%
  pivot_wider(
    names_from  = gene_selection,
    values_from = c(Total, Bio.conservation, Batch.correction)
  ) %>%
  mutate(
    delta_Total = Total_angl             - Total_hvg,
    delta_Bio   = Bio.conservation_angl  - Bio.conservation_hvg,
    delta_Batch = Batch.correction_angl  - Batch.correction_hvg
  )

stat_scores <- diff_scores %>%
  group_by(integration_method) %>%
  summarise(
    n                  = n(),
    median_delta_Total = median(delta_Total, na.rm = TRUE),
    p_Total            = wilcox.test(delta_Total, mu = 0, alternative = "greater")$p.value,
    median_delta_Bio   = median(delta_Bio,   na.rm = TRUE),
    p_Bio              = wilcox.test(delta_Bio,   mu = 0, alternative = "greater")$p.value,
    median_delta_Batch = median(delta_Batch, na.rm = TRUE),
    p_Batch            = wilcox.test(delta_Batch, mu = 0, alternative = "greater")$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    p_Total_fdr  = p.adjust(p_Total,  method = "BH"),
    p_Bio_fdr    = p.adjust(p_Bio,    method = "BH"),
    p_Batch_fdr  = p.adjust(p_Batch,  method = "BH")
  )

stat_scores
```

## 3.3 Sanity check: topbtvr vs bottombtvr (paired Wilcoxon)

This is the core sanity check — the BTVR oracle should show a highly significant
improvement. If this test is not significant, the simulation parameters are
not informative for integration quality.

```{r stat_topbtvr_vs_bottombtvr}
diff_oracle <- metrics %>%
  filter(gene_selection %in% c("topbtvr", "bottombtvr")) %>%
  select(sample, integration_method, gene_selection, Total, Bio.conservation, Batch.correction) %>%
  pivot_wider(
    names_from  = gene_selection,
    values_from = c(Total, Bio.conservation, Batch.correction)
  ) %>%
  mutate(
    delta_Total = Total_topbtvr             - Total_bottombtvr,
    delta_Bio   = Bio.conservation_topbtvr  - Bio.conservation_bottombtvr,
    delta_Batch = Batch.correction_topbtvr  - Batch.correction_bottombtvr
  )

stat_oracle <- diff_oracle %>%
  group_by(integration_method) %>%
  summarise(
    n                  = n(),
    median_delta_Total = median(delta_Total, na.rm = TRUE),
    p_Total            = wilcox.test(delta_Total, mu = 0, alternative = "greater")$p.value,
    median_delta_Bio   = median(delta_Bio,   na.rm = TRUE),
    p_Bio              = wilcox.test(delta_Bio,   mu = 0, alternative = "greater")$p.value,
    median_delta_Batch = median(delta_Batch, na.rm = TRUE),
    p_Batch            = wilcox.test(delta_Batch, mu = 0, alternative = "greater")$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    p_Total_fdr  = p.adjust(p_Total,  method = "BH"),
    p_Bio_fdr    = p.adjust(p_Bio,    method = "BH"),
    p_Batch_fdr  = p.adjust(p_Batch,  method = "BH")
  )

cat("=== topbtvr vs bottombtvr (oracle sanity check) ===\n")
print(stat_oracle)
cat("\n=== angl vs hvg ===\n")
print(stat_scores)
```

# Session info

```{r session_info}
sessionInfo()
```
