---
title: "compare_genes"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/output/")})
output:
    html_document:
      toc: yes
      toc_float:
        collapsed: no
        smooth_scroll: no
      df_print: paged
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.height = 10)
knitr::opts_chunk$set(fig.width = 14)
options(Seurat.object.assay.version = "v5")
```



```{r, load libraries}
.libPaths(.Library)
.libPaths()
library(tidyverse)
library(ComplexHeatmap)
library(pbapply)
library(rhdf5)
library(httpgd)
library(plotly)
library(jsonlite)
```


```{r}
gene_files <- list.files("/data/akalin/akollot/anglemania_benchmark/simulated/with_dropout/preprocessed", full.names = TRUE)
samplesheet_normalCells <- fromJSON("/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/h5ad/with_dropout/samplesheet.json") %>% as_tibble()
samplesheet_SEACells <- fromJSON("/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/h5ad/with_dropout/SEACells/merged_seacell_data/samplesheet.json") %>% as_tibble()
samplesheet <- samplesheet_SEACells %>%
  bind_rows(samplesheet_normalCells) %>%
  mutate(sim_type = ifelse(grepl("_merged_seacell", sample_name), "metacells", "normalCells"))

find_matching_files <- function(sample_name, files) {
  matching_files <- grep(sample_name, files, value = TRUE)
  # matching_files <- ifelse(grepl("_merged_seacell", sample_name), matching_files, matching_files[!grepl("_merged_seacell", matching_files)])
  return(matching_files)
}

identify_gene_selections <- function(file_name) {
    if (grepl("_full.tsv", file_name)) {
        return("full")
    } else if (grepl("_hvg.tsv", file_name)) {
        return("hvg")
    } else if (grepl("_rand.tsv", file_name)) {
        return("rand")
    } else if (grepl("_angl.tsv", file_name)) {
        return("angl")
    }
}

expanded <- samplesheet %>%
  rowwise() %>%
  mutate(tsv_files = list(find_matching_files(sample_name, gene_files))) %>%
  tidyr::unnest(tsv_files) %>%
  rowwise() %>%
  mutate(
    gene_selection = identify_gene_selections(tsv_files),
    gene_selection = factor(gene_selection, levels = c("full", "hvg", "rand", "angl"))
  ) %>%
 filter(
    !(sim_type == "normalCells" & str_detect(tsv_files, "_merged_seacell"))
  ) %>%
  ungroup()
View(expanded)
```

```{r}
outplot <- "/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/output/plots/compare_gene_selections_dropout"
```


# check overlap of anglemania genes to HVGs as a function of the dropout.mid parameter of splatter
```{r}
result <- expanded %>%
  filter(gene_selection %in% c("hvg", "angl")) %>%
  group_by(sample_name) %>%
  mutate(group_size = n()) %>%
         filter(group_size > 1)

result <- result %>%
  pivot_wider(names_from = gene_selection, values_from = tsv_files) %>%
  rowwise() %>%
  mutate(angl = list(read.csv(angl, sep = "\t", header = T) %>% pull(hgnc_symbol)),
         hvg = list(read.csv(hvg, sep = "\t", header = T) %>% pull(hgnc_symbol)))

unique(sapply(result$angl, length))

for_plot <- result %>%
  rowwise() %>%
  mutate(intersection = list(intersect(angl, hvg)),
    intersection_size = length(intersection),
    intersection_size_perc = intersection_size / length(angl)*100) %>% 
    select(-angl, -hvg) 

```

```{r}
for_plot %>%
  filter(sim == "sim2") %>%
  ggplot(aes(x = as.factor(dropout.mid), y = intersection_size_perc)) +
  geom_point(aes(color = sim_type), size = 5) +
  labs(x = "strength of the dropout", y = "intersection of HVG and anglemania genes [%]", title = "overlap of anglemania genes to HVGs as a function of the dropout.mid parameter of splatter") +
  theme(text = element_text(size = 20, face = "bold"))
```

# check overlap of genes from the same gene selection but comparing metacells to normal cells
```{r}
result <- expanded %>%
  select(-c(parent_sample_path, parent_sample, merged_seacell_sparsity)) %>%
  filter(gene_selection %in% c("hvg", "angl")) %>%
  filter(sim == "sim2") %>%
  group_by(sample_name) %>%
  mutate(sample_name = gsub("_merged_seacell", "", sample_name)) %>%
  pivot_wider(names_from = sim_type, values_from = c(tsv_files, file_path)) %>%
  rowwise() %>%
  mutate(
    tsv_files_metacells = ifelse(is.na(tsv_files_metacells), NA, list(read.csv(tsv_files_metacells, sep = "\t", header = T) %>% pull(hgnc_symbol))),
    tsv_files_normalCells = ifelse(is.na(tsv_files_normalCells), NA, list(read.csv(tsv_files_normalCells, sep = "\t", header = T) %>% pull(hgnc_symbol)))
  ) %>%
  ungroup() %>%
  mutate(tsv_files_metacells = lapply(tsv_files_metacells, function(x) gsub("Gene", "", x)), 
         tsv_files_normalCells = lapply(tsv_files_normalCells, function(x) gsub("Gene", "", x)))

unique(sapply(result$angl, length))

for_plot <- result %>%
  rowwise() %>%
  mutate(
    intersection = list(intersect(tsv_files_metacells, tsv_files_normalCells)),
    intersection_size = length(intersection),
    intersection_size_perc = intersection_size / length(tsv_files_metacells) * 100,
    n_genes_metacells = length(tsv_files_metacells)
  ) %>%
  select(-tsv_files_metacells, -tsv_files_normalCells)
```

```{r}
for_plot %>%
  ggplot(aes(x = as.factor(dropout.mid), y = intersection_size_perc)) +
  geom_bar(aes(fill = n_genes_metacells), stat = "identity") +
  facet_wrap(~gene_selection) +
  labs(x = "strength of the dropout", y = "intersection of genes selected from metacells vs normal cells [%]", title = "overlap of genes selected from metacells vs normal cells") +
  theme(text = element_text(size = 15, face = "bold"), axis.text.x = element_text(size = 10))
```


# make score for each gene that describes the bio to batch variability ratio
```{r}
library(scales)
get_rowData <- function(sim_h5ad_path){
  df <- h5read(sim_h5ad_path, "/var") %>% as.data.frame() %>% column_to_rownames("X_index")
  return(df)
}

get_BTVR <- function(rowdata) {
  # Calculate BioVar and TechVar using vectorized operations
  bio_var_cols <- select(rowdata, starts_with("DEFacGroup"))
  tech_var_cols <- select(rowdata, starts_with("BatchFacBatch"))
  
  # Use apply function to compute variances row-wise
  BioVar <- apply(bio_var_cols, 1, var)
  BioVar[BioVar > 0] <- BioVar[BioVar > 0] %>% log() %>% rescale()
  TechVar <- apply(tech_var_cols, 1, var) %>% log() %>% rescale()
  
  # Compute BTVR directly
  BTVR <- BioVar/TechVar
  
  # Create result tibble
  summary_df <- tibble(
    gene = rownames(rowdata),
    BTVR = BTVR,
    BioVar = BioVar,
    TechVar = TechVar
  )
  
  return(summary_df)
}
```


```{r}
btvr <- expanded %>%
  filter(gene_selection %in% c("hvg", "angl")) %>%
  group_by(sample_name) %>%
  mutate(group_size = n()) %>%
         filter(group_size > 1) %>% # make sure to only include the samples that have both the angl and hvg genes already calculated (snakemake pipeline preprocess)
  pivot_wider(names_from = gene_selection, values_from = tsv_files)

# construct nested tibble but don't use tidyverse cause it's freaking slow
rowdatas <- pblapply(btvr$original_h5ad, get_rowData, cl = 20)
h5closeAll()
BTVR <- pblapply(rowdatas, get_BTVR, cl = 20)
angl <- pblapply(btvr$angl, function(x) {
  genes <- read.csv(x, sep = "\t", header = T) %>% pull(hgnc_symbol)
  }, cl = 20)
hvg <- pblapply(btvr$hvg, function(x) {
  genes <- read.csv(x, sep = "\t", header = T) %>% pull(hgnc_symbol)
  }, cl = 20)

btvr$BTVR <- BTVR
btvr$angl <- angl
btvr$hvg <- hvg

btvr_raw <- btvr 
```

```{r}

btvr <- btvr_raw %>% 
  mutate(btvr_dist = pmap(
list(angl, hvg, BTVR),
      ~ { genes <- unique(c(..1, ..2))
        # Create a tibble for genes with their BTVR values
        merged_tbl <- ..3 %>%
          filter(gene %in% genes) %>%
          mutate(
            angl = gene %in% ..1,
            hvg = gene %in% ..2
          )
        # Return the new tibble
        merged_tbl} )) 

btvr <- btvr %>%
  rowwise() %>%
  mutate(btvr_dist = list(btvr_dist %>% pivot_longer(cols = c(angl, hvg), 
    names_to = "gene_selection", 
    values_to = "is_selected") %>% filter(is_selected)))

for (i in 1:10){
  plot_df <- btvr[i, ]
  print(plot_df)
  # print(plot_df$btvr_dist)
  print(plot_df$btvr_dist %>%
    as.data.frame() %>%
    ggplot(aes(gene_selection, BTVR))+
    geom_violin(aes(fill = gene_selection))+
    ggtitle(paste0("Bio to Technical variability ratio. de.facLoc: ", plot_df$de.facLoc, " batch.facLoc: ", plot_df$batch.facLoc)))
}

```


* anglemania genes have a much higher BTVR (bio to technical variability) => this is great!
```{r}

unnested <- btvr %>%
  select(-BTVR, -angl, -hvg) %>%
  unnest(btvr_dist) %>%
  ungroup() %>%
  filter(!is.infinite(BTVR)) # there was one gene (!) that had 0 TechVar so BTVR was infinite, but really only 1 gene lol.
head(unnested)
# unnested %>%
#   ggplot()
unnested %>%
  distinct(ngroup, nbatch)

# Summarize the data and pivot to wide format
summary_df_BTVR <- unnested %>%
  group_by(gene_selection, batch.facLoc, de.facLoc) %>%
  summarise(
    mean = mean(BTVR, na.rm = TRUE),
    median = median(BTVR, na.rm = TRUE),
    sd = sd(BTVR, na.rm = TRUE),
    n = n(),
    .groups = 'drop'  # Avoid unnecessary grouping
  ) 


```


```{r}
# Plot heatmap
new_labels <- c("angl" = "anglemania genes", "hvg" = "highly variable genes")
size <- 1-6e-4
ggplot(summary_df_BTVR, aes(x = as.factor(batch.facLoc), y = as.factor(de.facLoc), fill = mean)) +
  geom_tile(width = size, height = size) +
  coord_equal() +
  facet_wrap(~gene_selection, labeller = labeller(gene_selection = new_labels)) +  # Facet by gene_selection
  scale_fill_gradient(low = "grey", high = "purple", limits = c(0, 1.1)) +
  # scale_fill_continuous(type = "viridis", limits = c(0, 1.1)) +
  labs(title = "Heatmap of MEAN BTVR by Batch and Biological Variability",
       x = "Strength of Technical Noise (batch.facLoc)",
       y = "Biological Variability (de.facLoc)",
       fill = "Mean BTVR") +
  theme_minimal() +
    theme(
      text = element_text(size = 15, face = "bold"),
      axis.title = element_text(size = 20),
      strip.text = element_text(size = 20)
    )
ggsave(
  "/home/akollot/projects/CRC1588/dataset_integration/anglemania_manuscript/plots/mean_BTVR_heatmap.pdf",
  width = 10,
  height = 8
)
```


```{r}
size <- 1 - 6e-4
ggplot(summary_df_BTVR, aes(x = as.factor(batch.facLoc), y = as.factor(de.facLoc), fill = median)) +
  geom_tile(width = size, height = size) +
  coord_equal() +
  facet_wrap(~gene_selection, labeller = labeller(gene_selection = new_labels)) +  # Facet by gene_selection
  scale_fill_gradient(low = "grey", high = "purple", limits = c(0, 1.1)) +
  labs(
    title = "Heatmap of MEDIAN BTVR by Batch and Biological Variability",
    x = "Strength of Technical Noise (batch.facLoc)",
    y = "Biological Variability (de.facLoc)",
    fill = "Median BTVR"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 15, face = "bold"),
    axis.title = element_text(size = 20),
    # increase facet label size
    strip.text = element_text(size = 20)
  )
ggsave(
  "/home/akollot/projects/CRC1588/dataset_integration/anglemania_manuscript/plots/median_BTVR_heatmap.pdf",
  width = 10,
  height = 8
)

```

* it looks like the higher BTVR is due to the fact that angelmania captures genes with high biological variability!
```{r}
summary_df_BioVar <- unnested %>%
  group_by(gene_selection, batch.facLoc, de.facLoc) %>%
  summarise(
    mean = mean(BioVar, na.rm = TRUE),
    median = median(BioVar, na.rm = TRUE),
    sd = sd(BioVar, na.rm = TRUE),
    n = n(),
    .groups = 'drop'  # Avoid unnecessary grouping
  ) 


ggplot(summary_df_BioVar, aes(x = batch.facLoc, y = de.facLoc, fill = mean)) +
  geom_tile() +
  facet_wrap(~gene_selection) +  # Facet by gene_selection
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Heatmap of MEAN BioVar by Batch and Biological Variability",
       x = "Batch Factor Level (batch.facLoc)",
       y = "Biological Variability (de.facLoc)",
       fill = "Mean BioVar") +
  theme_minimal()


ggplot(summary_df_BioVar, aes(x = batch.facLoc, y = de.facLoc, fill = median)) +
  geom_tile() +
  facet_wrap(~gene_selection) +  # Facet by gene_selection
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Heatmap of MEDIAN BioVar by Batch and Biological Variability",
       x = "Batch Factor Level (batch.facLoc)",
       y = "Biological Variability (de.facLoc)",
       fill = "Median BioVar") +
  theme_minimal()

```

* However it still captures genes with high technical variability, just like HVGs, with increasing Technical Variability / Batch effects
```{r}
summary_df_TechVar <- unnested %>%
  group_by(gene_selection, batch.facLoc, de.facLoc) %>%
  summarise(
    mean = mean(TechVar, na.rm = TRUE),
    median = median(TechVar, na.rm = TRUE),
    sd = sd(TechVar, na.rm = TRUE),
    n = n(),
    .groups = 'drop'  # Avoid unnecessary grouping
  ) 


ggplot(summary_df_TechVar, aes(x = batch.facLoc, y = de.facLoc, fill = mean)) +
  geom_tile() +
  facet_wrap(~gene_selection) +  # Facet by gene_selection
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Heatmap of MEAN TechVar by Batch and Biological Variability",
       x = "Batch Factor Level (batch.facLoc)",
       y = "Biological Variability (de.facLoc)",
       fill = "Mean TechVar") +
  theme_minimal()


ggplot(summary_df_TechVar, aes(x = batch.facLoc, y = de.facLoc, fill = median)) +
  geom_tile() +
  facet_wrap(~gene_selection) +  # Facet by gene_selection
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Heatmap of MEDIAN TechVar by Batch and Biological Variability",
       x = "Batch Factor Level (batch.facLoc)",
       y = "Biological Variability (de.facLoc)",
       fill = "Median TechVar") +
  theme_minimal()

```

```{r}
test <- rowdatas[[1]]
View(test)
```




# Download button
```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
  var images = document.querySelectorAll('img');
  images.forEach(function(img, index) {
    var button = document.createElement('button');
    button.textContent = 'Download as PDF';
    button.style = 'display: block; margin-top: 10px;';

    button.onclick = function() {
      var jsPDF = document.createElement('script');
      jsPDF.src = 'https://cdnjs.cloudflare.com/ajax/libs/jspdf/2.3.1/jspdf.umd.min.js';
      jsPDF.onload = function() {
        var { jsPDF } = window.jspdf;
        const doc = new jsPDF({
          orientation: 'portrait',
          unit: 'pt',
          format: 'a4'
        });

        var imgEl = new Image();
        imgEl.onload = function() {
          // Calculate scale ratio to fit the image within the page
          var scaleWidth = 595 / this.width;
          var scaleHeight = 842 / this.height;
          var scale = Math.min(scaleWidth, scaleHeight);

          // Calculate dimensions to maintain aspect ratio
          var imgWidth = this.width * scale;
          var imgHeight = this.height * scale;

          // Position the image centered horizontally and vertically
          var xPos = (595 - imgWidth) / 2;
          var yPos = (842 - imgHeight) / 2;

          doc.addImage(this, 'PNG', xPos, yPos, imgWidth, imgHeight);
          doc.save('download-' + index + '.pdf');
        };
        imgEl.crossOrigin = "Anonymous";
        imgEl.src = img.src;
      };
      document.body.appendChild(jsPDF);
    };

    img.parentNode.insertBefore(button, img.nextSibling);
  });
});
```