---
title: "compare_genes"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/output/")})
output:
    html_document:
      toc: yes
      toc_float:
        collapsed: no
        smooth_scroll: no
      df_print: paged
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.height = 10)
knitr::opts_chunk$set(fig.width = 14)
options(Seurat.object.assay.version = "v5")
```



```{r, load libraries}
library(tidyverse)
library(ComplexHeatmap)
library(pbapply)
library(rhdf5)
library(httpgd)
library(plotly)
library(jsonlite)
library(glue)
source("/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/scripts/R/ggplot_utils.R")
today <- function() {
  return(format(Sys.Date(), "%Y%m%d"))
}
```


```{r}
config <- yaml::read_yaml("/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/scripts/config_files/config_varying_batch_biovar_dropout_sample.yml")
gene_files <- list.files(config$visualization$preprocessed_dir, full.names = TRUE)
samplesheet <- read.csv(config$samplesheet, sep = "\t", stringsAsFactors = FALSE)
find_matching_files <- function(sample_name, files) {
  matching_files <- grep(sample_name, files, value = TRUE)
  # matching_files <- ifelse(grepl("_merged_seacell", sample_name), matching_files, matching_files[!grepl("_merged_seacell", matching_files)])
  return(matching_files)
}

identify_gene_selections <- function(file_name) {
    if (grepl("_full.tsv", file_name)) {
        return("full")
    } else if (grepl("_hvg.tsv", file_name)) {
        return("hvg")
    } else if (grepl("_rand.tsv", file_name)) {
        return("rand")
    } else if (grepl("_angl.tsv", file_name)) {
        return("angl")
    }
}

expanded <- samplesheet %>%
  rowwise() %>%
  mutate(tsv_files = list(find_matching_files(sample_name, gene_files))) %>%
  tidyr::unnest(tsv_files) %>%
  rowwise() %>%
  mutate(
    gene_selection = identify_gene_selections(tsv_files),
    gene_selection = factor(gene_selection, levels = c("full", "hvg", "rand", "angl"))
  ) %>%
  ungroup()
```

```{r}
outplot <- config$visualization$outplot_dir_compare_gene_selection
if (!dir.exists(outplot)) {
    dir.create(outplot, recursive = TRUE)
}
```


# Plot intersection size of HVGs and anglemania genes
* It looks like the intersection size of angl and hvg decreases with increasing batch effect
* and it increases with increasing biological variability!
* can we show that angl genes are more meaningful?
* plot the intersection size of hvg and angl against the batch effect and biological variability
```{r, loading result}
result <- expanded %>%
  filter(gene_selection %in% c("hvg", "angl")) %>%
  group_by(sample_name) %>%
  mutate(group_size = n()) %>%
         filter(group_size > 1)

result <- result %>%
  pivot_wider(names_from = gene_selection, values_from = tsv_files) %>%
  rowwise() %>%
  dplyr::mutate(angl = list(read.csv(angl, sep = "\t", header = T) %>% pull(hgnc_symbol)),
         hvg = list(read.csv(hvg, sep = "\t", header = T) %>% pull(hgnc_symbol)))
# saveRDS(result, file = "/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/output/compare_gene_selections_result.RDS")
# result <- readRDS("/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/output/compare_gene_selections_result.RDS")
result <- result %>%
  filter(
    de.facLoc != 0.001,
    batch.facLoc != 0.001
  ) %>%
  mutate(
    de.facLoc = case_when(
      de.facLoc == 0 ~ "none",
      de.facLoc == 0.01 ~ "small",
      de.facLoc == 0.05 ~ "mild",
      de.facLoc == 0.1 ~ "moderate",
      de.facLoc == 0.3 ~ "strong",
      de.facLoc == 0.6 ~ "very strong"
    ),
    de.facLoc = factor(de.facLoc, levels = c("none", "small", "mild", "moderate", "strong", "very strong")),
    batch.facLoc = case_when(
      batch.facLoc == 0 ~ "none",
      batch.facLoc == 0.01 ~ "small",
      batch.facLoc == 0.05 ~ "mild",
      batch.facLoc == 0.1 ~ "moderate",
      batch.facLoc == 0.3 ~ "strong",
      batch.facLoc == 0.6 ~ "very strong"
    ),
    batch.facLoc = factor(batch.facLoc, levels = c("none", "small", "mild", "moderate", "strong", "very strong"))
  )
head(result)
result %>%
  ggplot(aes(length(angl)))+
  geom_histogram()

for_plot <- result %>%
  rowwise() %>%
  mutate(intersection = list(intersect(angl, hvg)),
    intersection_size = length(intersection),
    intersection_size_perc = intersection_size / length(angl)*100) %>% 
    select(-angl, -hvg) 
colnames(for_plot)
colnames(result)
cor_plot <- for_plot %>%
  select(
    where(is.numeric),
    de.facLoc, batch.facLoc,
    -groupCells, -group_size
  ) %>%
  mutate(across(everything(), as.numeric))

corrs <- cor(cor_plot$intersection_size_perc, cor_plot%>% select(-intersection_size, -intersection_size_perc))
corrs
library(corrplot)
corrplot(corrs, method = "circle")
```

```{r, plotting intersection size}
p <- for_plot %>%
  ungroup() %>%
  group_by(de.facLoc) %>%
  mutate(median_intersection_size = median(intersection_size_perc)) %>%
  ggplot(aes(as.factor(de.facLoc), intersection_size_perc))+
  geom_boxplot(aes(fill = median_intersection_size)) +
  labs(x = "Biological Variability", y = "Intersection Size (%)") +
  # manually adjust tick labels
  scale_y_continuous(breaks = c(0, 20, 40, 60, 80, 100), limits = c(0,100))+
  scale_fill_continuous(low = "white", high = "blue", limits = c(0, 100)) + 
  theme_minimal() +
  # ggtitle("Percent of anglemania genes that intersect with HVG. Plotted against biological variability.")+
  theme(text = element_text(size = 20, face = "bold"),
        legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
p
ggsave(
  file = file.path(outplot, glue("{today()}_Biological_Variability_vs_intersection_size.pdf")),
  plot = p,
  width = 6,
  height = 5
)

p <- for_plot %>%
  ungroup() %>%
    group_by(batch.facLoc) %>%
    mutate(median_intersection_size = median(intersection_size_perc)) %>%
    ggplot(aes(as.factor(batch.facLoc), intersection_size_perc)) +
      geom_boxplot(aes(fill = median_intersection_size)) +
      labs(x = "Technical Variability", y = "Intersection Size (%)") +
      # manually adjust tick labels
      scale_y_continuous(breaks = c(0, 20, 40, 60, 80, 100), limits = c(0, 100)) +
      scale_fill_continuous(low = "white", high = "blue", limits = c(0, 100)) +
      theme_minimal() +
      # ggtitle("PercenRt of anglemania genes that intersect with HVG. Plotted against biological variability.")+
      theme(
        text = element_text(size = 20, face = "bold"),
        legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1)
  )
p

ggsave(
  file = file.path(outplot, glue("{today()}_Technical_Variability_vs_intersection_size.pdf")),
  plot = p,
  width = 6,
  height = 5
)

p <- for_plot %>%
  ungroup() %>%
  group_by(sparsity) %>%
  mutate(median_intersection_size = median(intersection_size_perc)) %>%
  ggplot(aes(as.factor(sparsity), intersection_size_perc)) +
    geom_boxplot(aes(fill = median_intersection_size)) +
    labs(x = "Sparsity", y = "Intersection Size (%)") +
    # manually adjust tick labels
    scale_y_continuous(breaks = c(0, 20, 40, 60, 80, 100), limits = c(0, 100)) +
    scale_fill_continuous(low = "white", high = "blue", limits = c(0, 100)) +
    theme_minimal() +
    # ggtitle("PercenRt of anglemania genes that intersect with HVG. Plotted against biological variability.")+
    theme(
      text = element_text(size = 20, face = "bold"),
      legend.position = "none",
      axis.text.x = element_text(angle = 20, hjust = 1)
    )
p
ggsave(
  file = file.path(outplot, glue("{today()}_Sparsity_vs_intersection_size.pdf")),
  plot = p,
  width = 6,
  height = 5
)
```


# make score for each gene that describes the bio to batch variability ratio
```{r}
library(scales)
get_rowData <- function(sim_h5ad_path){
  df <- h5read(sim_h5ad_path, "/var") %>% as.data.frame() %>% column_to_rownames("X_index")
  return(df)
}

get_BTVR <- function(rowdata) {
  # Calculate BioVar and TechVar using vectorized operations
  bio_var_cols <- dplyr::select(rowdata, starts_with("DEFacGroup"))
  tech_var_cols <- dplyr::select(rowdata, starts_with("BatchFacBatch"))

  # Use apply function to compute variances row-wise
  raw_BioVar <- apply(bio_var_cols, 1, var)
  BioVar <- raw_BioVar
  BioVar[BioVar > 0] <- raw_BioVar[raw_BioVar > 0] %>%
    log() %>%
    rescale()
  raw_TechVar <- apply(tech_var_cols, 1, var)
  TechVar <- raw_TechVar
  TechVar[TechVar > 0] <- raw_TechVar[raw_TechVar > 0] %>%
    log() %>%
    rescale()

  # TechVar <- raw_TechVar %>%
  #   log() %>%
  #   rescale()

  # Compute BTVR directly
  BTVR <- log((raw_BioVar + 1e-15) / (raw_TechVar + 1e-15)) %>% rescale()
  GeneScore <- BTVR * BioVar # basically "scaled" by BioVar

  # Create result tibble
  summary_df <- tibble(
    gene = rownames(rowdata),
    BTVR = BTVR,
    GeneScore = GeneScore,
    raw_BioVar = raw_BioVar,
    BioVar = BioVar,
    raw_TechVar = raw_TechVar,
    TechVar = TechVar,
    GeneMean = rowdata$GeneMean
  )

  return(summary_df)
}
```


```{r}
btvr <- expanded %>%
  filter(gene_selection %in% c("hvg", "angl")) %>%
  group_by(sample_name) %>%
  mutate(group_size = n()) %>%
         filter(group_size > 1) %>% # make sure to only include the samples that have both the angl and hvg genes already calculated (snakemake pipeline preprocess)
  pivot_wider(names_from = gene_selection, values_from = tsv_files)

# construct nested tibble but don't use tidyverse cause it's freaking slow
rowdatas <- pblapply(btvr$original_h5ad, get_rowData, cl = 20)
h5closeAll()
BTVR <- pblapply(rowdatas, get_BTVR, cl = 20)
angl <- pblapply(btvr$angl, function(x) {
  genes <- read.csv(x, sep = "\t", header = T) %>% pull(hgnc_symbol)
  }, cl = 20)
hvg <- pblapply(btvr$hvg, function(x) {
  genes <- read.csv(x, sep = "\t", header = T) %>% pull(hgnc_symbol)
  }, cl = 20)

btvr$BTVR <- BTVR
btvr$angl <- angl
btvr$hvg <- hvg

btvr_raw <- btvr 
```

```{r}

btvr <- btvr_raw %>% 
  mutate(btvr_dist = pmap(
list(angl, hvg, BTVR),
      ~ { genes <- unique(c(..1, ..2))
        # Create a tibble for genes with their BTVR values
        merged_tbl <- ..3 %>%
          filter(gene %in% genes) %>%
          mutate(
            angl = gene %in% ..1,
            hvg = gene %in% ..2
          )
        # Return the new tibble
        merged_tbl} )) 

btvr <- btvr %>%
  rowwise() %>%
  mutate(btvr_dist = list(btvr_dist %>% pivot_longer(cols = c(angl, hvg), 
    names_to = "gene_selection", 
    values_to = "is_selected") %>% filter(is_selected)))


```


* anglemania genes have a much higher BTVR (bio to technical variability) => this is great!
```{r}

unnested <- btvr %>%
  select(-BTVR, -angl, -hvg) %>%
  unnest(btvr_dist) %>%
  ungroup() %>%
  filter(!is.infinite(BTVR)) # there was one gene (!) that had 0 TechVar so BTVR was infinite, but really only 1 gene lol.
head(unnested)
# unnested %>%
#   ggplot()
unnested %>%
  distinct(ngroup, nbatch)

# Summarize the data and pivot to wide format
summary_df_BTVR <- unnested %>%
  group_by(gene_selection, batch.facLoc, de.facLoc) %>%
  summarise(
    mean = mean(BTVR, na.rm = TRUE),
    median = median(BTVR, na.rm = TRUE),
    sd = sd(BTVR, na.rm = TRUE),
    n = n(),
    .groups = 'drop'  # Avoid unnecessary grouping
  ) 


```

# Plotting
## BTVR
### Mean
```{r}
# Plot heatmap
p <- unnested %>%
  draw_heatmap2(
    df = .,
    score = "BTVR",
    fill = "mean",
    scale_fill_limits = c(0, 1)
  )
p

ggsave(
  file.path(outplot, glue("{today()}_mean_BTVR_heatmap.pdf")),
  p,
  width = 11,
  height = 8
)
```

### Median
```{r}
p <- unnested %>%
  draw_heatmap2(
    df = .,
    score = "BTVR",
    fill = "median",
    scale_fill_limits = c(0, 1)
  )
p
ggsave(
  file.path(outplot, glue("{today()}_median_BTVR_heatmap.pdf")),
  p,
  width = 11,
  height = 8
)

```

## BTVR scaled by BioVar
### Mean
```{r}
# Summarize the data and pivot to wide format
p <- unnested %>%
  draw_heatmap2(
    df = .,
    score = "GeneScore",
    fill = "mean",
    scale_fill_limits = c(0, 1)
  )

p

ggsave(
  file.path(outplot, glue("{today()}_mean_BTVR-Scaled-by-BioVar_heatmap.pdf")),
  p,
  width = 11,
  height = 8
)
```


### Median
```{r}
p <- unnested %>%
  draw_heatmap2(
    df = .,
    score = "GeneScore",
    fill = "median",
    scale_fill_limits = c(0, 1)
  )

p
```

## BioVar
* it looks like the higher BTVR is due to the fact that angelmania captures genes with high biological variability!

### Mean
```{r}
p <- unnested %>%
  draw_heatmap2(
    df = .,
    score = "BioVar",
    fill = "mean"
  )
p

ggsave(
  file.path(outplot, glue("{today()}_mean_BioVar_heatmap.pdf")),
  width = 11,
  height = 8
)
```

### Median
```{r}
p <- unnested %>%
  draw_heatmap2(
    df = .,
    score = "BioVar",
    fill = "median"
  )

p

ggsave(
  file.path(outplot, glue("{today()}_median_BioVar_heatmap.pdf")),
  width = 11,
  height = 8
)
```


## raw_BioVar
* it looks like the higher BTVR is due to the fact that angelmania captures genes with high biological variability!

### Mean
```{r}
p <- unnested %>%
  draw_heatmap2(
    df = .,
    score = "raw_BioVar",
    fill = "mean"
  )

p

ggsave(
  file.path(outplot, glue("{today()}_mean_rawBioVar_heatmap.pdf")),
  width = 11,
  height = 8
)
```

### Median
```{r}
p <- unnested %>%
  draw_heatmap2(
    df = .,
    score = "raw_BioVar",
    fill = "median"
  )

p

ggsave(
  file.path(outplot, glue("{today()}_median_rawBioVar_heatmap.pdf")),
  width = 11,
  height = 8
)
```


## TechVar
* However it still captures genes with high technical variability, just like HVGs, with increasing Technical Variability / Batch effects

### Mean
```{r}
p <- unnested %>%
  draw_heatmap2(
    df = .,
    score = "TechVar",
    fill = "mean"
  )

p

ggsave(
  file.path(outplot, glue("{today()}_mean_TechVar_heatmap.pdf")),
  width = 11,
  height = 8
)

```

### Median
```{r}
p <- unnested %>%
  draw_heatmap2(
    df = .,
    score = "TechVar",
    fill = "median",
    add_geom_text = TRUE
  )

p

p <- unnested %>%
  draw_heatmap2(
    df = .,
    score = "TechVar",
    fill = "median"
  )

p

ggsave(
  file.path(outplot, glue("{today()}_median_TechVar_heatmap.pdf")),
  width = 11,
  height = 8)
```



## rawTechVar
* However it still captures genes with high technical variability, just like HVGs, with increasing Technical Variability / Batch effects

### Mean
```{r}
p <- unnested %>%
  draw_heatmap2(
    df = .,
    score = "raw_TechVar",
    fill = "mean"
  )
p

p <- unnested %>%
  mutate(raw_TechVar = log(raw_TechVar)) %>%
  draw_heatmap2(
    df = .,
    score = "raw_TechVar",
    fill = "mean"
  )
p

ggsave(
  file.path(outplot, glue("{today()}_mean_rawlogTechVar_heatmap.pdf")),
  width = 11,
  height = 8
)

```

### Median
```{r}
p <- unnested %>%
  draw_heatmap2(
    df = .,
    score = "raw_TechVar",
    fill = "median"
  )
p

p <- unnested %>%
  mutate(raw_TechVar = log(raw_TechVar)) %>%
  draw_heatmap2(
    df = .,
    score = "raw_TechVar",
    fill = "median"
  )

p

ggsave(
  file.path(outplot, glue("{today()}_median_rawlogTechVar_heatmap.pdf")),
  width = 11,
  height = 8)
```

```{r}
p <- unnested %>%
  group_by(de.facLoc, sparsity, gene_selection) %>%
  summarise(mean_total = mean(BTVR), .groups = "drop") %>%
  ggplot(aes(x = sparsity, y = mean_total, color = as.factor(de.facLoc))) +
  geom_point(size = 6, alpha = 0.8, aes(shape = gene_selection)) +
  ylim(0, 1) +
  guides(color = guide_legend(ncol = 2)) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    x = "Sparsity [%]",
    y = "BTVR",
    color = "Biological Variability",
    title = glue("Total Score by \nsparsity and de.facLoc")
  )
p
ggsave(
  file.path(outplot, glue("{today()}_scatter_mean_BTVR_by_sparsity.pdf")),
  p,
  width = 11,
  height = 8
)
```

# Statistical tests

Are BTVR, BioVar, and TechVar significantly different between anglemania and HVG selected genes?

**Approach**: per sample, compute median BTVR/BioVar/TechVar for each gene_selection.
Then apply a one-sample Wilcoxon test on the per-sample differences (angl − hvg).
H1 for BTVR/BioVar: angl > hvg. H1 for TechVar: hvg > angl (lower tech noise for angl).

```{r stat_btvr_per_sample}
per_sample <- unnested %>%
  group_by(sample_name, gene_selection, batch.facLoc, de.facLoc, sparsity) %>%
  summarise(
    median_BTVR    = median(BTVR,        na.rm = TRUE),
    median_BioVar  = median(BioVar,      na.rm = TRUE),
    median_TechVar = median(TechVar,     na.rm = TRUE),
    .groups = "drop"
  )

diff_btvr <- per_sample %>%
  pivot_wider(
    names_from  = gene_selection,
    values_from = starts_with("median_")
  ) %>%
  mutate(
    delta_BTVR    = median_BTVR_angl   - median_BTVR_hvg,
    delta_BioVar  = median_BioVar_angl  - median_BioVar_hvg,
    delta_TechVar = median_TechVar_hvg  - median_TechVar_angl  # positive = angl has lower tech var
  )

# Overall test (all conditions pooled)
overall_tests <- tibble(
  metric       = c("BTVR", "BioVar", "TechVar (hvg−angl)"),
  median_delta = c(
    median(diff_btvr$delta_BTVR,    na.rm = TRUE),
    median(diff_btvr$delta_BioVar,  na.rm = TRUE),
    median(diff_btvr$delta_TechVar, na.rm = TRUE)
  ),
  p_value = c(
    wilcox.test(diff_btvr$delta_BTVR,    mu = 0, alternative = "greater")$p.value,
    wilcox.test(diff_btvr$delta_BioVar,  mu = 0, alternative = "greater")$p.value,
    wilcox.test(diff_btvr$delta_TechVar, mu = 0, alternative = "greater")$p.value
  )
) %>%
  mutate(p_fdr = p.adjust(p_value, method = "BH"))

cat("=== Overall test (all conditions pooled) ===\n")
print(overall_tests)
```

```{r stat_btvr_by_sparsity}
# Does the BTVR advantage of anglemania change with sparsity?
stat_by_sparsity <- diff_btvr %>%
  group_by(sparsity) %>%
  summarise(
    n            = n(),
    median_delta = median(delta_BTVR, na.rm = TRUE),
    p_value      = tryCatch(
      wilcox.test(delta_BTVR, mu = 0, alternative = "greater")$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) %>%
  mutate(
    p_fdr = p.adjust(p_value, method = "BH"),
    sig   = case_when(p_fdr < 0.001 ~ "***", p_fdr < 0.01 ~ "**",
                      p_fdr < 0.05  ~ "*",   TRUE ~ "ns")
  )

stat_by_sparsity
```

```{r stat_btvr_per_condition}
# Per-condition Wilcoxon (batch.facLoc × de.facLoc × sparsity)
stat_btvr <- diff_btvr %>%
  group_by(batch.facLoc, de.facLoc) %>%
  summarise(
    n                   = n(),
    median_delta_BTVR   = median(delta_BTVR,  na.rm = TRUE),
    p_BTVR              = tryCatch(wilcox.test(delta_BTVR, mu = 0, alternative = "greater")$p.value, error = function(e) NA_real_),
    .groups = "drop"
  ) %>%
  mutate(
    p_BTVR_fdr = p.adjust(p_BTVR, method = "BH"),
    sig_BTVR   = case_when(p_BTVR_fdr < 0.001 ~ "***", p_BTVR_fdr < 0.01 ~ "**",
                            p_BTVR_fdr < 0.05  ~ "*",   TRUE ~ "ns")
  )

head(stat_btvr)
```

```{r stat_btvr_heatmap, fig.height=6, fig.width=8}
p_sig_btvr <- ggplot(stat_btvr, aes(x = batch.facLoc, y = de.facLoc, fill = median_delta_BTVR)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sig_BTVR), size = 5) +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "#64487A",
    midpoint = 0, limits = c(-0.15, 0.15), oob = scales::squish
  ) +
  coord_equal() +
  labs(
    title    = "Paired BTVR test: anglemania vs HVG (dropout simulations)",
    subtitle = "Fill = median Δ BTVR (angl − hvg); label = BH-corrected significance",
    x = "Batch effect (batch.facLoc)", y = "Biological variability (de.facLoc)",
    fill = "Δ BTVR"
  ) +
  theme_bw() +
  theme(text = element_text(size = 14))

p_sig_btvr
ggsave(
  file.path(outplot, glue("{today()}_stat_BTVR_significance_heatmap.pdf")),
  p_sig_btvr, width = 8, height = 6
)
```

```{r stat_btvr_by_sparsity_plot, fig.height=5}
p_sparsity <- ggplot(diff_btvr, aes(x = factor(sparsity), y = delta_BTVR)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_boxplot(fill = "#64487A", alpha = 0.7, outlier.size = 0.5) +
  geom_text(
    data = stat_by_sparsity,
    aes(x = factor(sparsity), y = max(diff_btvr$delta_BTVR, na.rm = TRUE) * 1.05, label = sig),
    size = 5
  ) +
  labs(
    title    = "BTVR advantage of anglemania by sparsity level",
    subtitle = "Stars = BH-corrected Wilcoxon p-value (angl > hvg)",
    x = "Sparsity [%]", y = "Δ median BTVR (angl − hvg)"
  ) +
  theme_bw() +
  theme(text = element_text(size = 14))

p_sparsity
ggsave(
  file.path(outplot, glue("{today()}_stat_BTVR_by_sparsity.pdf")),
  p_sparsity, width = 8, height = 5
)
```

# sessionInfo
```{r}
sessionInfo()
```