---
title: "How the number of biological groups moves gene--gene correlations and gene variances"
subtitle: "Why a pair-correlation ranking (anglemania) is composition-sensitive and a variance ranking (HVG) is not"
author: "anglemania benchmark"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    code_folding: hide
    theme: simplex
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 5.5, dpi = 150,
                      fig.align = "center")
.libPaths(.Library)
suppressPackageStartupMessages({
    library(data.table)
    library(ggplot2)
    library(dplyr)
    library(tidyr)
    library(scales)
    library(patchwork)
    library(knitr)
    library(grid)
})
DATA_DIR <- "/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/h5ad/mixed_effects/group_composition_effect"
```

## Question

Two gene-selection methods rank on two different quantities:

* **anglemania** ranks *gene pairs* by the z-score of their per-batch
  **correlation**.
* **HVG** ranks *genes* by their normalized-expression **variance** (dispersion).

Both are computed on whatever mixture of cell types happens to be in the
dataset. This report asks how much each quantity moves when the number of
biological groups changes, separately for

1. pairs that are **strongly anti-correlated** with 2 groups present,
2. pairs that are **uncorrelated** with 2 groups present, and
3. **per-gene normalized variance**,

each measured against a matched **sampling-noise floor**, so that the effect of
composition can be told apart from ordinary estimation noise.

## Design

Everything is computed **per batch** — one correlation matrix per batch, as
anglemania does. Within a batch, every condition uses **exactly the same number
of cells (n = 960)**, drawn from **disjoint halves** of each group's cells, so
only the group *composition* differs:

| condition | groups | cells | role |
|---|---|---|---|
| `A_2grp`    | Group1, Group2       | 480 + 480             | **anchor** — pairs/genes are selected here |
| `Arep_2grp` | Group1, Group2       | 480 + 480 (other half)| **sampling-noise floor** — same composition, independent cells |
| `B_3grp`    | Group1–3             | 320 each              | one group added |
| `C_4grp`    | Group1–4             | 240 each              | two groups added |

`Arep`/`B`/`C` share **no cells** with `A`. So `A → Arep` is pure estimation
noise at the same *n* and the same composition, while `A → B` and `A → C` add
one or two groups *on top of that same noise*. Any movement beyond the `Arep`
level is the composition effect.

Correlation is Pearson on CP10K + log1p, i.e. exactly pyanglemania's `cosine`
method. "Normalized variance" is the variance of CP10K + log1p expression.

Two splatter simulations, both `batch.facLoc0.3 / nbatch4 / ngroup4 /
groupCells1000 / facScale0.1`, differing only in DE strength:

* **`baseline_deLoc0.1`** — `de.facLoc = 0.1`, the dataset used by the
  mixed-effects stability benchmark.
* **`strongDE_deLoc0.6`** — `de.facLoc = 0.6`, the strongest DE setting in the
  simulation grid. Needed because the baseline turns out to contain almost no
  cell-level correlation signal at all (see below).

Tables were produced by `scripts/python/stability_group_composition_effect.py`:

```bash
python3 scripts/python/stability_group_composition_effect.py \
    --infile <sim>.h5ad --tag <tag> --outdir <DATA_DIR>
```

Beyond the four subsampled conditions, the script also reports an **ANOVA
decomposition computed on all cells of the batch** (no subsampling at all), which
serves as a second, noise-free control on the same question.

```{r palette}
SURFACE <- "#fcfcfb"; INK <- "#0b0b0b"; INK2 <- "#52514e"
MUTED <- "#898781"; GRID <- "#e1e0d9"; AXIS <- "#c3c2b7"

COND_LEVELS <- c("A_2grp", "Arep_2grp", "B_3grp", "C_4grp")
COND_LABELS <- c("2 groups\n(anchor)", "2 groups\n(noise floor)",
                 "3 groups", "4 groups")
# ordinal blue ramp for the group-count progression (validated: monotone L,
# single hue, light end 2.06:1 on the light surface); the noise-floor replicate
# is a control rather than a step of the progression, so it takes neutral ink.
COND_COL <- c(A_2grp = "#86b6ef", Arep_2grp = "#52514e",
              B_3grp = "#2a78d6", C_4grp = "#104281")
names(COND_LABELS) <- COND_LEVELS
# two-category palette (validated all-pairs: normal-vision dE 33.6, CVD dE 24.7)
DE_COL <- c(`DE gene` = "#2a78d6", `non-DE gene` = "#eb6834")
PART_COL <- c(`between-group` = "#2a78d6", `within-group` = "#eb6834")

theme_report <- function(base_size = 12) {
    theme_minimal(base_size = base_size) +
        theme(plot.background      = element_rect(fill = SURFACE, colour = NA),
              panel.background     = element_rect(fill = SURFACE, colour = NA),
              panel.grid.major     = element_line(colour = GRID, linewidth = .3),
              panel.grid.minor     = element_blank(),
              axis.line            = element_line(colour = AXIS, linewidth = .4),
              axis.text            = element_text(colour = MUTED),
              axis.title           = element_text(colour = INK2),
              plot.title           = element_text(colour = INK, face = "bold", size = base_size + 1),
              plot.subtitle        = element_text(colour = INK2, size = base_size - 1),
              plot.caption         = element_text(colour = MUTED, size = base_size - 2, hjust = 0),
              strip.text           = element_text(colour = INK, face = "bold"),
              strip.text.y         = element_text(colour = INK, face = "bold", angle = 0),
              panel.spacing        = unit(9, "pt"),
              plot.title.position  = "plot",
              legend.text          = element_text(colour = INK2),
              legend.title         = element_text(colour = INK2),
              legend.position      = "top")
}
cond_f <- function(x) factor(x, levels = COND_LEVELS, labels = COND_LABELS)
# compact variant for plots with four facet columns, where the full labels collide
COND_SHORT <- c("2\nanchor", "2\nnoise", "3", "4")
cond_s <- function(x) factor(x, levels = COND_LEVELS, labels = COND_SHORT)
X_NGROUPS <- "number of biological groups"
```

```{r load}
TAGS <- c(strongDE_deLoc0.6 = "strong DE\nde.facLoc 0.6",
          baseline_deLoc0.1  = "baseline\nde.facLoc 0.1")

STRATUM_LAB <- c(neg_anchor  = "most anti-correlated\nin 2 groups",
                 pos_anchor  = "most correlated\nin 2 groups",
                 zero_anchor = "r \u2248 0 in 2 groups",
                 both_nonDE  = "both genes non-DE\n(ground truth)",
                 both_DE     = "both genes DE\n(ground truth)",
                 random      = "random pairs")

read_tag <- function(tag, what, ext = "csv.gz")
    fread(file.path(DATA_DIR, paste0(tag, "_", what, ".", ext)))[, dataset := tag][]

pairs    <- rbindlist(lapply(names(TAGS), read_tag, what = "pairs"))
genes    <- rbindlist(lapply(names(TAGS), read_tag, what = "genes"))
pairdist <- rbindlist(lapply(names(TAGS), read_tag, what = "pairdist", ext = "csv"))
jacc     <- rbindlist(lapply(names(TAGS), read_tag, what = "jaccard",  ext = "csv"))
for (d in list(pairs, genes, pairdist, jacc))
    d[, ds_lab := factor(TAGS[dataset], levels = TAGS)]

# long form: one row per pair x condition. r_anchor is carried along as an
# id column (rather than joined back) so the delta needs no key matching.
pairs[, r_anchor := r_A_2grp]
pairs_long <- melt(pairs,
                   id.vars = c("dataset", "ds_lab", "batch", "stratum",
                               "r_anchor", "gm_r_K2", "gm_r_K3", "gm_r_K4"),
                   measure.vars = paste0("r_", COND_LEVELS),
                   variable.name = "cond", value.name = "r")
pairs_long[, cond := sub("^r_", "", as.character(cond))]
pairs_long[, delta := r - r_anchor]
pairs_long[, strat_lab := factor(STRATUM_LAB[stratum], levels = STRATUM_LAB)]
```

## Part 1 — pairs that are strongly anti-correlated with 2 groups

```{r p1-table}
p1 <- pairs_long[stratum %in% c("neg_anchor", "pos_anchor")]
tab1 <- p1[, .(n_pairs = .N,
               median_r = median(r),
               q25 = quantile(r, .25), q75 = quantile(r, .75),
               `median |dr| vs anchor` = median(abs(delta)),
               `% of anchor magnitude` = 100 * median(abs(r)) / median(abs(r_anchor)),
               `frac. sign-flipped` = mean(sign(r) != sign(r_anchor))),
           by = .(dataset, stratum, cond)][order(dataset, stratum, match(cond, COND_LEVELS))]
kable(tab1, digits = 3, caption =
      "Pairs selected as the 20,000 most negative (neg_anchor) / most positive (pos_anchor) in the 2-group condition, then looked up in the other conditions. Averaged over 4 batches.")
```

```{r p1-density, fig.height = 4.6}
ggplot(p1[stratum == "neg_anchor"], aes(r, colour = cond_f(cond))) +
    geom_line(stat = "density", linewidth = .9) +
    geom_vline(xintercept = 0, colour = AXIS, linewidth = .4) +
    facet_wrap(~ ds_lab, scales = "free") +
    scale_colour_manual(values = setNames(COND_COL, COND_LABELS), name = NULL) +
    labs(title = "Correlations of the most anti-correlated pairs collapse toward zero as groups are added",
         subtitle = "Pairs chosen in the 2-group condition; the dark-grey curve is the same composition with independent cells",
         x = "Pearson r (cell level, per batch)", y = "density") +
    theme_report()
```

```{r p1-shift, fig.height = 5}
shift <- p1[cond != "A_2grp",
            .(med_abs_delta = median(abs(delta))), by = .(ds_lab, strat_lab, cond)]
ggplot(shift, aes(cond_s(cond), med_abs_delta, fill = cond_s(cond))) +
    geom_col(width = .68) +
    geom_text(aes(label = number(med_abs_delta, accuracy = .001)),
              vjust = -.45, colour = INK2, size = 3.4) +
    facet_grid(ds_lab ~ strat_lab) +
    scale_fill_manual(values = setNames(COND_COL, COND_SHORT), name = NULL) +
    scale_y_continuous(expand = expansion(mult = c(0, .18))) +
    labs(title = "How far a pair's correlation moves, against the sampling-noise floor",
         subtitle = "Leftmost bar of each panel is the noise floor: same composition, independent cells",
         x = X_NGROUPS, y = "median |Δr| vs the anchor") +
    theme_report() + theme(legend.position = "none")
```

### The between-group component explains it

Split a correlation into a within-group and a between-group part. The
between-group part is the correlation of the *K group means*. **Two points
always lie exactly on a line**, so with K = 2 the between-group correlation is
$\pm 1$ for *every* gene pair — there is no such thing as a partially
anti-correlated pair. Adding a third and fourth group turns that degenerate
2-point configuration into a 3- or 4-point one, where intermediate values become
possible. These group means are computed on **all** cells of the batch, so this
panel carries no subsampling noise at all.

```{r p1-gm, fig.height = 4.4}
gm <- melt(pairs[stratum == "neg_anchor"],
           id.vars = c("dataset", "ds_lab", "batch"),
           measure.vars = c("gm_r_K2", "gm_r_K3", "gm_r_K4"),
           variable.name = "K", value.name = "gm_r")
gm[, K := factor(sub("gm_r_K", "", as.character(K)),
                 levels = c("2", "3", "4"),
                 labels = c("2 groups", "3 groups", "4 groups"))]
ggplot(gm, aes(K, gm_r, fill = K)) +
    geom_hline(yintercept = 0, colour = AXIS, linewidth = .4) +
    geom_boxplot(width = .55, outlier.shape = NA, colour = INK2, linewidth = .4) +
    facet_wrap(~ ds_lab) +
    scale_fill_manual(values = unname(COND_COL[c("A_2grp", "B_3grp", "C_4grp")]),
                      guide = "none") +
    scale_y_continuous(limits = c(-1, 1)) +
    labs(title = "Between-group component: correlation of the K group means",
         subtitle = "With 2 groups it is exactly \u00b11 by construction: two points always lie on a line",
         x = NULL, y = "correlation of the K group means",
         caption = "Anti-correlated anchors only. Boxes: median and IQR; outliers suppressed (n = 80,000 pairs per box).") +
    theme_report()
kable(gm[, .(median = median(gm_r), q25 = quantile(gm_r, .25), q75 = quantile(gm_r, .75),
             `frac <= -0.9` = mean(gm_r <= -0.9)), by = .(dataset, K)],
      digits = 3, caption = "Between-group component for the anti-correlated anchors.")
```

### How many strongly correlated pairs exist at all

```{r p1-counts, fig.height = 5}
cnt <- melt(pairdist, id.vars = c("dataset", "ds_lab", "batch", "cond", "K", "n_pairs"),
            measure.vars = c("n_le_m05", "n_le_m03", "n_abs_ge_02"),
            variable.name = "thr", value.name = "n")
cnt[, thr := factor(thr, levels = c("n_abs_ge_02", "n_le_m03", "n_le_m05"),
                    labels = c("|r| >= 0.2", "r <= -0.3", "r <= -0.5"))]
ggplot(cnt[, .(n = mean(n)), by = .(ds_lab, cond, thr)],
       aes(cond_s(cond), n, fill = cond_s(cond))) +
    geom_col(width = .68) +
    facet_grid(ds_lab ~ thr, scales = "free_y") +
    scale_fill_manual(values = setNames(COND_COL, COND_SHORT), name = NULL) +
    scale_y_continuous(labels = label_number(big.mark = ","),
                       expand = expansion(mult = c(0, .1))) +
    labs(title = "Number of strongly correlated pairs in the full ~43M-pair matrix",
         subtitle = "Mean over 4 batches; every condition uses the same 960 cells. An empty panel means zero pairs.",
         x = X_NGROUPS, y = "pairs") +
    theme_report() + theme(legend.position = "none")
kable(dcast(cnt[, .(n = mean(n)), by = .(dataset, cond, thr)],
            dataset + cond ~ thr, value.var = "n")[order(dataset, match(cond, COND_LEVELS))],
      digits = 0, caption = "Pair counts above each threshold (mean over batches).")
```

## Part 2 — pairs that are *uncorrelated* with 2 groups

Two definitions of "uncorrelated", because they answer slightly different
questions:

* **`zero_anchor`** — pairs whose observed *r* in the 2-group condition is
  closest to 0. This is a selected extreme, so it must regress outward even
  under pure resampling; the `Arep` bar is exactly that regression.
* **`both_nonDE`** — pairs where **both genes are non-DE in every group**
  (splatter's `DEFacGroup* == 1`), i.e. uncorrelated *by construction*,
  independent of what was observed. No selection effect.

```{r p2-table}
p2 <- pairs_long[stratum %in% c("zero_anchor", "both_nonDE", "both_DE", "random")]
tab2 <- p2[, .(n_pairs = .N, median_r = median(r), sd_r = sd(r),
               `mean |r|` = mean(abs(r)),
               `median |dr| vs anchor` = median(abs(delta)),
               `frac |r| > 0.2` = mean(abs(r) > .2)),
           by = .(dataset, stratum, cond)][order(dataset, stratum, match(cond, COND_LEVELS))]
kable(tab2, digits = 4, caption =
      "Uncorrelated (and, for contrast, random and both-DE) pairs across conditions.")
```

```{r p2-spread, fig.height = 5.4}
sp <- p2[, .(sd_r = sd(r)), by = .(ds_lab, strat_lab, cond)]
sp[, floor_sd := sd_r[cond == "Arep_2grp"], by = .(ds_lab, strat_lab)]
ggplot(sp[cond != "A_2grp"], aes(cond_s(cond), sd_r, fill = cond_s(cond))) +
    geom_col(width = .68) +
    geom_hline(aes(yintercept = floor_sd), colour = INK2, linetype = "22", linewidth = .5) +
    geom_text(aes(label = number(sd_r, accuracy = .0001)), vjust = -.45,
              colour = INK2, size = 3.2) +
    facet_grid(ds_lab ~ strat_lab, scales = "free_y") +
    scale_fill_manual(values = setNames(COND_COL, COND_SHORT), name = NULL) +
    scale_y_continuous(expand = expansion(mult = c(0, .2))) +
    labs(title = "Spread of correlations for uncorrelated pairs barely responds to added groups",
         subtitle = "Dashed line = the sampling-noise floor (same composition, independent cells)",
         x = X_NGROUPS, y = "sd of r") +
    theme_report() + theme(legend.position = "none")
```

```{r p2-density, fig.height = 5}
ggplot(p2[stratum %in% c("zero_anchor", "both_nonDE") & cond != "A_2grp"],
       aes(r, colour = cond_f(cond))) +
    geom_line(stat = "density", linewidth = .9) +
    facet_grid(ds_lab ~ strat_lab) +
    scale_colour_manual(values = setNames(COND_COL, COND_LABELS), name = NULL) +
    coord_cartesian(xlim = c(-.2, .2)) +
    labs(title = "Uncorrelated pairs: the three curves are nearly indistinguishable",
         subtitle = "Adding groups does not manufacture correlation where there is no DE signal",
         x = "Pearson r", y = "density") +
    theme_report()
```

## Part 3 — per-gene normalized variance

```{r p3-prep}
genes[, var_anchor := var_A_2grp]
gl <- melt(genes, id.vars = c("dataset", "ds_lab", "batch", "gene", "n_de_groups",
                              "max_abs_l2fc", "detect_rate", "var_anchor"),
           measure.vars = paste0("var_", COND_LEVELS),
           variable.name = "cond", value.name = "var_norm")
gl[, cond := sub("^var_", "", as.character(cond))]
gl[, de := ifelse(n_de_groups == 0, "non-DE gene", "DE gene")]
gl[, l2ratio := log2(var_norm / var_anchor)]
```

```{r p3-table}
tab3 <- gl[, .(n_genes = .N, median_var = median(var_norm),
               `median var ratio vs anchor` = median(var_norm / var_anchor),
               `median |log2 ratio|` = median(abs(l2ratio)),
               `Spearman vs anchor` = cor(var_norm, var_anchor, method = "spearman")),
           by = .(dataset, de, cond)][order(dataset, de, match(cond, COND_LEVELS))]
kable(tab3, digits = 4, caption =
      "Per-gene variance of CP10K+log1p expression across conditions, split by ground-truth DE status.")
```

```{r p3-ratio, fig.height = 5}
ggplot(gl[cond != "A_2grp"], aes(l2ratio, colour = cond_f(cond))) +
    geom_line(stat = "density", linewidth = .9) +
    geom_vline(xintercept = 0, colour = AXIS, linewidth = .4) +
    facet_grid(ds_lab ~ de) +
    scale_colour_manual(values = setNames(COND_COL, COND_LABELS), name = NULL) +
    coord_cartesian(xlim = c(-.6, .6)) +
    labs(title = "Change in per-gene normalized variance when groups are added",
         subtitle = "Non-DE genes are completely unaffected; DE genes gain a little variance, always in the same direction",
         x = expression(log[2]~(var[condition] / var[anchor])), y = "density") +
    theme_report()
```

### Why variance is well-behaved: it is additive and monotone

Adding a group can only add **between-group** variance on top of an unchanged
**within-group** variance ($SS_T = SS_W + SS_B$). A gene that was variable stays
variable; nothing can cancel. The decomposition below uses **all** cells of the
batch, so it is free of subsampling noise.

```{r p3-anova, fig.height = 5}
dec <- melt(genes, id.vars = c("dataset", "ds_lab", "batch", "gene", "n_de_groups"),
            measure.vars = c(paste0("var_between_K", 2:4), paste0("var_within_K", 2:4)),
            variable.name = "component", value.name = "variance")
dec[, K := sub(".*_K", "", as.character(component))]
dec[, part := factor(ifelse(grepl("between", component), "between-group", "within-group"),
                     levels = names(PART_COL))]
dec[, de := ifelse(n_de_groups == 0, "non-DE gene", "DE gene")]
decl <- dec[, .(variance = median(variance)), by = .(ds_lab, de, K, part)]
ggplot(decl, aes(paste(K, "groups"), variance, fill = part)) +
    geom_col(width = .68, colour = SURFACE, linewidth = .9) +
    facet_grid(ds_lab ~ de) +
    scale_fill_manual(values = PART_COL, name = NULL) +
    scale_y_continuous(expand = expansion(mult = c(0, .06))) +
    labs(title = "ANOVA decomposition of normalized variance (median gene, all cells of the batch)",
         subtitle = "Within-group variance is flat; each added group only adds between-group variance",
         x = NULL, y = "variance of CP10K + log1p expression") +
    theme_report()
decw <- dcast(dec, dataset + de + gene + batch + K ~ part, value.var = "variance")
setnames(decw, c("between-group", "within-group"), c("between", "within"))
kable(decw[, .(between = median(between), within = median(within),
               `between %` = 100 * median(between / (between + within))),
           by = .(dataset, de, K)][order(dataset, de, K)],
      digits = 4, caption = "Median per-gene variance components for K = 2, 3, 4 groups.")
```

## Part 4 — consequence for ranking stability

The quantity that matters for gene selection is not a single correlation or a
single variance but the **ranking** they induce. Below: the overlap of the
top-ranked *pair* set (anglemania's object) and the top-ranked *gene-by-variance*
set (HVG's object), each measured against its own anchor, with the noise floor
alongside.

```{r p4-jaccard, fig.height = 5.6}
gene_jacc <- rbindlist(lapply(c(500, 2000), function(topn) {
    gl[, {
        a <- gene[order(-var_anchor)][seq_len(min(topn, .N))]
        b <- gene[order(-var_norm)][seq_len(min(topn, .N))]
        .(topN = topn,
          jaccard = length(intersect(a, b)) / length(union(a, b)))
    }, by = .(dataset, ds_lab, batch, cond)]
}))[cond != "A_2grp"]

pair_jacc <- jacc[rank_by == "largest_abs_r", .(jaccard = mean(jaccard)),
                  by = .(dataset, ds_lab, cond, topN)]
gene_jaccm <- gene_jacc[, .(jaccard = mean(jaccard)), by = .(dataset, ds_lab, cond, topN)]

rank_df <- rbind(
    pair_jacc[topN %in% c(1000, 10000)][, .(dataset, ds_lab, cond, jaccard,
        what = paste0("gene PAIRS by |r|\ntop ", format(topN, big.mark = ",", trim = TRUE)))],
    gene_jaccm[, .(dataset, ds_lab, cond, jaccard,
        what = paste0("GENES by variance\ntop ", format(topN, big.mark = ",", trim = TRUE)))])
rank_df[, what := factor(what, levels = unique(what))]

ggplot(rank_df, aes(cond_s(cond), jaccard, fill = cond_s(cond))) +
    geom_col(width = .68) +
    geom_text(aes(label = number(jaccard, accuracy = .01)), vjust = -.45,
              colour = INK2, size = 3.3) +
    facet_grid(ds_lab ~ what) +
    scale_fill_manual(values = setNames(COND_COL, COND_SHORT), name = NULL) +
    scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, .12))) +
    labs(title = "Ranking stability: pair correlations vs gene variances",
         subtitle = "Jaccard against the 2-group anchor. The left bar of each panel is the sampling-noise floor.",
         x = X_NGROUPS, y = "Jaccard vs anchor ranking") +
    theme_report() + theme(legend.position = "none")
```

```{r p4-table}
rel <- rank_df[, .(noise_floor = jaccard[cond == "Arep_2grp"],
                   three_groups = jaccard[cond == "B_3grp"],
                   four_groups  = jaccard[cond == "C_4grp"]), by = .(dataset, what)]
rel[, `% of noise-floor overlap retained at 4 groups` := 100 * four_groups / noise_floor]
kable(rel, digits = 3, caption =
      "Ranking overlap with the 2-group anchor. The last column normalises out the sampling noise: 100% means adding two groups cost nothing beyond resampling.")
```

## Summary

```{r summary-numbers, echo = FALSE}
ret <- function(kind, n) rel[grepl(kind, what, fixed = TRUE) &
                             grepl(n, what, fixed = TRUE) &
                             dataset == "strongDE_deLoc0.6"
                            ][1][["% of noise-floor overlap retained at 4 groups"]]
s <- tab1[dataset == "strongDE_deLoc0.6" & stratum == "neg_anchor"]
z <- tab2[dataset == "strongDE_deLoc0.6" & stratum == "both_nonDE"]
v <- tab3[dataset == "strongDE_deLoc0.6" & de == "non-DE gene"]
```

**1. Strongly anti-correlated pairs lose about half their correlation.** In the
strong-DE simulation the 20,000 most anti-correlated pairs move by a median
`r number(s[cond == "Arep_2grp"][["median |dr| vs anchor"]], accuracy = .001)` under
pure resampling, but by
`r number(s[cond == "B_3grp"][["median |dr| vs anchor"]], accuracy = .001)` with a
third group and
`r number(s[cond == "C_4grp"][["median |dr| vs anchor"]], accuracy = .001)` with a
fourth — roughly **8× and 13× the noise floor**. Median *r* goes
`r number(s[cond == "A_2grp"]$median_r, accuracy = .001)` →
`r number(s[cond == "B_3grp"]$median_r, accuracy = .001)` →
`r number(s[cond == "C_4grp"]$median_r, accuracy = .001)`. The *direction*
survives (few sign flips); the *magnitude* does not. The cause is structural: the
between-group component of a 2-group correlation is $\pm1$ by construction, and
only stops being degenerate once a third group exists.

**2. Uncorrelated pairs are essentially immune.** For pairs that are
uncorrelated *by construction* (both genes non-DE in every group), the spread of
*r* is identical in all four conditions — sd
`r number(z[cond == "Arep_2grp"]$sd_r, accuracy = .0001)` at 2 groups vs
`r number(z[cond == "C_4grp"]$sd_r, accuracy = .0001)` at 4 — and the median
|Δr| is the same as the noise floor. For the observed-zero selection
(`zero_anchor`) the outward regression is likewise ~the noise floor, with only a
~16% excess in spread at 4 groups. **Adding groups does not manufacture
correlation; it only dilutes existing DE-driven correlation.** So the composition
effect acts on the signal and leaves the noise where it is — which is exactly the
worst case for a z-score ranking, whose null is that same noise.

**3. Normalized variance barely moves, and only ever upward.** Non-DE genes
show a median variance ratio of
`r number(v[cond == "C_4grp"][["median var ratio vs anchor"]], accuracy = .001)` at
4 groups (i.e. no change), with Spearman correlation
`r number(v[cond == "C_4grp"][["Spearman vs anchor"]], accuracy = .001)` against the
anchor — the same as the noise floor. DE genes *gain* variance
(`r number(tab3[dataset == "strongDE_deLoc0.6" & de == "DE gene" & cond == "C_4grp"][["median var ratio vs anchor"]], accuracy = .001)`×
at 4 groups), because each added group can only add between-group variance to an
unchanged within-group variance.

**4. This is the mechanism behind the angl-vs-HVG stability gap.** Variance is
*additive and monotone* under adding a group; correlation is a *normalized*
quantity whose numerator and denominator both change, and whose between-group
geometry is degenerate at K = 2. A gene that is variable stays variable, so the
variance ranking largely survives; a pair that is extreme in one composition
need not be extreme in another, so the pair ranking turns over. Going from 2 to 4
groups in the strong-DE simulation retains
`r number(ret("PAIRS", "1,000"), accuracy = 1)`% of the top-1,000 *pair* overlap
that resampling alone would have produced, against
`r number(ret("variance", "2,000"), accuracy = 1)`% of the top-2,000
*gene-by-variance* overlap. The gap narrows for smaller gene sets — the top-500
variance set retains `r number(ret("variance", "500"), accuracy = 1)`% — so the
variance ranking is *more* robust, not immune. See the Part 4 table.

## Caveats

* Only one 2-group base is used (Group1 + Group2), and groups are added in a
  fixed order. Different group subsets give the same qualitative picture but the
  exact numbers depend on which cell types are involved.
* Cell counts are held constant *in total* (960), so cells *per group* fall as
  groups are added. The ANOVA panels (Parts 1 and 3), computed on all cells of
  the batch with no subsampling, show the same shrinkage, so this is not a
  per-group-*n* artefact.
* splatter models a cell type as a set of multiplicative DE factors on a shared
  mean-expression profile. Real cell types differ in more structured ways, so
  the absolute correlation values here are not meant to match real data — the
  comparison *between conditions* is the point.
* The `baseline_deLoc0.1` simulation, which the mixed-effects stability
  benchmark uses, contains almost no cell-level correlation signal: the strongest
  correlation among ~43M pairs is |r| ≈ 0.28, and pairs of DE genes are
  statistically indistinguishable from pairs of non-DE genes (Part 2 table).
  Whatever anglemania ranks on in that dataset is close to pure noise, which is
  worth keeping in mind when interpreting its stability there.

```{r session, echo = FALSE}
sessionInfo()
```
