---
title: "GTEs analysis"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/output/")})
output:
    html_document:
      toc: yes
      toc_float:
        collapsed: no
        smooth_scroll: no
      df_print: paged
---

# introduction
Chatty gave a pretty good idea on how to use the GTEs package to show how anglemania
genes are less affected by batch effects than HVGs:
- “HBG removal” ablation (smoking gun)
    - Quantify per‑gene batch‑sensitivity, drop the worst offenders from the HVG set,
    and watch integration improve.
- Design: For each gene, compute a gene‑level batch score like the GTE/HBG-style
score from Zhou et al. (2025). Rank genes by batch‑sensitivity; re‑run integration
after removing the top 1–5% from the HBG list.

- What you’ll see: kBET/LISI improve sharply once a handful of HBGs are excluded; in
several datasets, removing only a few genes makes a disproportionate difference. 

- Figure panel suggestion: line plot: “% HBGs removed” (x) vs “iLISI/kBET” (y) for 
HVG‑based integration; a dotted horizontal line for anglemania’s performance to show convergence.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.height = 10)
knitr::opts_chunk$set(fig.width = 14)
```


# load libraries
```{r, load libraries}
library(tidyverse)
library(pbapply)
library(rhdf5)
library(httpgd)
library(GTEs)
library(glue)
library(anndataR)
library(Seurat)
```

# Part I
- Check number of HBGs in HVG list and anglemania genes
- Sequential removal of more and more HBGs from the HVG list 
- Visualization of how integration scores of HVG integration converge with
anglemania integration when you remove HBGs.

## load data
- GTEs needs the counts matrix and the metadata 
```{r}
batch_facloc <- 0.3
de_facloc <- 0.1
sample_name <- glue("batch.facLoc{batch_facloc}_de.facLoc{de_facloc}_nbatch5_ngroup4_groupCells1000")
sim_h5ad_path <- glue("/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/h5ad/facScale0.1/{sample_name}.h5ad")
se <- anndataR::read_h5ad(sim_h5ad_path, to = "Seurat")
counts <- SeuratObject::LayerData(se, "counts")
meta <- se[[]]
anglemania_genes <- read.delim(glue("/data/akalin/akollot/anglemania_benchmark/simulated/facScale0.1/groupCells1000/preprocessed/{sample_name}_angl.tsv")) |> pull(hgnc_symbol)
hvg <- read.delim(glue("/data/akalin/akollot/anglemania_benchmark/simulated/facScale0.1/groupCells1000/preprocessed/{sample_name}_hvg.tsv")) |> pull(hgnc_symbol)
gene_list <- list(
    anglemania = anglemania_genes,
    hvg = hvg
)
```

```{r}

```