---
title: "investigate_sim2_subbatch_anglemania"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../output/")})
output:
    html_document:
      toc: yes
      toc_float:
        collapsed: no
        smooth_scroll: no
      df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.height = 8)
knitr::opts_chunk$set(fig.width = 12)
```

## Background

anglemania selected only **2 genes** for `sim2_subbatch_raw` (instead of 2000). The
suspected cause is the strict gene intersection step: anglemania requires a gene to be
expressed in ≥1 cell in **every** batch. With 16 sub-batches (Batch1Sub1…Batch4Sub4),
tiny/sparse sub-batches may express very few genes, collapsing the intersection.

**sim2 design (from sim2.R):**

- 4 main batches × 4 sub-batches = 16 sub-batches total
- Relative library sizes: Batch1=1.0, Batch2=0.7, Batch3=0.6, **Batch4=0.3** (very sparse)
- Sub-batch proportions: Batch4 sub-batches at 35%, 35%, **15%, 15%** → ~225 cells in smallest
- Sub-batch noise added per sub-batch via `splatSimulate(lib.loc=8, lib.scale=0.5)`

```{r libraries, message=FALSE, warning=FALSE}
.libPaths(.Library)
library(anndataR)
library(SingleCellExperiment)
library(Matrix)
library(tidyverse)
library(ggplot2)
```

## 1. Load Data

```{r load-data}
h5ad_path <- "/local/Projects/AAkalin_Neuroblastoma/data/Luecken_Theis_NatMethods_2021_benchmarking/datasets/sim2_subbatch_raw.h5ad"
sce <- anndataR::read_h5ad(h5ad_path, to = "SingleCellExperiment")

cat("Dimensions:", nrow(sce), "genes x", ncol(sce), "cells\n")
cat("obs columns:", paste(colnames(colData(sce)), collapse = ", "), "\n")
```

```{r identify-batch-col}
# Identify batch column (pipeline uses batch_key: "batch")
# From sim2.R colData: Batch, SubBatch, Sub, Group, etc.
cd <- as.data.frame(colData(sce))
head(cd)
```

```{r set-batch-cols}
# Determine correct column names based on what's in the data
batch_col <- if ("batch" %in% colnames(cd)) "batch" else if ("SubBatch" %in% colnames(cd)) "SubBatch" else stop("No batch column found")
main_batch_col <- if ("Batch" %in% colnames(cd)) "Batch" else NULL

cat("Using batch column:", batch_col, "\n")
cat("Unique batches:", length(unique(cd[[batch_col]])), "\n")
sort(unique(cd[[batch_col]]))
```

## 2. Cells per Batch

```{r cells-per-batch, fig.cap="Number of cells per sub-batch. Unequal sizes reflect the simulation design."}
cells_per_batch <- cd %>%
    count(.data[[batch_col]], name = "n_cells") %>%
    arrange(.data[[batch_col]])

ggplot(cells_per_batch, aes(x = .data[[batch_col]], y = n_cells)) +
    geom_col(fill = "steelblue") +
    geom_text(aes(label = n_cells), vjust = -0.3, size = 3) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
        title = "Cells per sub-batch (sim2_subbatch_raw)",
        x = "Sub-batch", y = "Number of cells"
    )
```

```{r cells-per-main-batch, eval=!is.null(main_batch_col)}
if (!is.null(main_batch_col)) {
    cells_per_main <- cd %>%
        count(.data[[main_batch_col]], name = "n_cells")

    ggplot(cells_per_main, aes(x = .data[[main_batch_col]], y = n_cells)) +
        geom_col(fill = "darkorange") +
        geom_text(aes(label = n_cells), vjust = -0.3, size = 3) +
        theme_bw() +
        labs(
            title = "Cells per main batch",
            x = "Batch", y = "Number of cells"
        )
}
```

## 3. Genes Expressed per Batch

For each sub-batch, count genes where **sum of counts > 1** (user request) and also
genes with **any expression > 0** (what anglemania uses for prefiltering).

```{r genes-per-batch}
batches <- sort(unique(cd[[batch_col]]))
count_mat <- counts(sce)

gene_stats <- do.call(rbind, lapply(batches, function(b) {
    batch_cells <- rownames(cd)[cd[[batch_col]] == b]
    mat <- count_mat[, batch_cells, drop = FALSE]
    data.frame(
        batch = b,
        n_cells = length(batch_cells),
        n_genes_any_expr = sum(Matrix::rowSums(mat > 0) >= 1),
        n_genes_sum_gt1  = sum(Matrix::rowSums(mat) > 1)
    )
}))

gene_stats
```

```{r plot-genes-any-expr, fig.cap="Number of genes with any expression (≥1 cell with count > 0) per sub-batch. This is what anglemania uses for per-batch gene prefiltering."}
gene_stats_long <- gene_stats %>%
    pivot_longer(
        cols = c(n_genes_any_expr, n_genes_sum_gt1),
        names_to = "metric",
        values_to = "n_genes"
    ) %>%
    mutate(metric = recode(metric,
        n_genes_any_expr = "rowSums > 0 (≥1 cell expressed)",
        n_genes_sum_gt1  = "total sum > 1"
    ))

ggplot(gene_stats_long, aes(x = batch, y = n_genes, fill = metric)) +
    geom_col(position = "dodge") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
        title = "Number of expressed genes per sub-batch",
        x = "Sub-batch", y = "Number of genes",
        fill = "Criterion"
    ) +
    scale_fill_manual(values = c("steelblue", "tomato"))
```

## 4. Gene Intersection Analysis

Simulate what anglemania does: take the strict intersection of expressed gene sets
across all sub-batches. Show how the intersection size shrinks as batches are added.

```{r gene-sets}
gene_sets <- lapply(batches, function(b) {
    batch_cells <- rownames(cd)[cd[[batch_col]] == b]
    mat <- count_mat[, batch_cells, drop = FALSE]
    rownames(mat)[Matrix::rowSums(mat > 0) >= 1]
})
names(gene_sets) <- batches

cat("Gene set sizes per batch:\n")
sizes <- sapply(gene_sets, length)
sort(sizes)
```

```{r cumulative-intersection, fig.cap="How the gene intersection shrinks as sub-batches are added. Batches are ordered from largest to smallest expressed gene set. The final intersection is what anglemania works with."}
# Order batches from most to fewest expressed genes (best case → worst case)
ordered_large_to_small <- names(sort(sizes, decreasing = TRUE))
cum_intersect_large_first <- sapply(seq_along(ordered_large_to_small), function(i) {
    length(Reduce(intersect, gene_sets[ordered_large_to_small[1:i]]))
})

# Also show natural alphabetical order
cum_intersect_alpha <- sapply(seq_along(batches), function(i) {
    length(Reduce(intersect, gene_sets[batches[1:i]]))
})

df_cum <- bind_rows(
    data.frame(
        n_batches = seq_along(ordered_large_to_small),
        intersection_size = cum_intersect_large_first,
        order = "Largest gene set first"
    ),
    data.frame(
        n_batches = seq_along(batches),
        intersection_size = cum_intersect_alpha,
        order = "Alphabetical order"
    )
)

ggplot(df_cum, aes(x = n_batches, y = intersection_size, color = order)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    theme_bw() +
    labs(
        title = "Cumulative gene intersection as sub-batches are added",
        x = "Number of sub-batches included",
        y = "Intersection size (# genes)",
        color = "Batch order"
    ) +
    scale_color_manual(values = c("steelblue", "tomato")) +
    geom_vline(xintercept = length(batches), linetype = "dashed", color = "gray50") +
    annotate("text", x = length(batches) - 0.5,
             y = max(df_cum$intersection_size) * 0.8,
             label = paste0("Final: ", tail(cum_intersect_large_first, 1)),
             hjust = 1, color = "steelblue")
```

```{r show-bottleneck-batches}
# Which batch(es) cause the biggest drop in intersection?
drops <- diff(cum_intersect_large_first)
bottleneck_idx <- which(drops == min(drops))
cat("Biggest intersection drop when adding batch #", bottleneck_idx + 1, ":\n")
cat("  Batch:", ordered_large_to_small[bottleneck_idx + 1], "\n")
cat("  Drop:", drops[bottleneck_idx], "genes\n")
cat("  Gene set size of this batch:", sizes[ordered_large_to_small[bottleneck_idx + 1]], "\n")
```

## 5. Total Count Distribution in Sparse Sub-batches

Look at the distribution of per-gene total counts in the smallest sub-batches
to understand why genes drop out.

```{r sparse-batch-counts, fig.cap="Distribution of per-gene total count sums in the 4 smallest sub-batches. Genes with sum=0 are unexpressed and dropped from the intersection."}
smallest_batches <- names(sort(sizes))[1:4]

count_dist <- do.call(rbind, lapply(smallest_batches, function(b) {
    batch_cells <- rownames(cd)[cd[[batch_col]] == b]
    mat <- count_mat[, batch_cells, drop = FALSE]
    gene_sums <- Matrix::rowSums(mat)
    data.frame(
        batch = b,
        gene_sum = as.numeric(gene_sums)
    )
}))

ggplot(count_dist, aes(x = gene_sum + 1)) +
    geom_histogram(bins = 50, fill = "steelblue", color = "white") +
    facet_wrap(~batch, scales = "free_y") +
    scale_x_log10() +
    theme_bw() +
    labs(
        title = "Per-gene count sum distribution in the 4 smallest sub-batches",
        x = "Total count sum per gene (log10 scale, +1)",
        y = "Number of genes"
    ) +
    geom_vline(xintercept = 1 + 1, linetype = "dashed", color = "tomato") +
    annotate("text", x = 2.5, y = Inf, vjust = 1.5,
             label = "sum > 1 threshold", color = "tomato", size = 3)
```

```{r zero-gene-fraction}
zero_frac <- do.call(rbind, lapply(batches, function(b) {
    batch_cells <- rownames(cd)[cd[[batch_col]] == b]
    mat <- count_mat[, batch_cells, drop = FALSE]
    gene_sums <- Matrix::rowSums(mat)
    data.frame(
        batch = b,
        n_cells = length(batch_cells),
        n_genes_total = nrow(mat),
        n_genes_zero  = sum(gene_sums == 0),
        frac_zero = mean(gene_sums == 0)
    )
})) %>% arrange(desc(frac_zero))

zero_frac
```

```{r plot-zero-frac, fig.cap="Fraction of genes with zero total counts per sub-batch. High zero fraction in small/sparse sub-batches directly limits the intersection."}
ggplot(zero_frac, aes(x = reorder(batch, frac_zero), y = frac_zero)) +
    geom_col(fill = "tomato") +
    geom_text(aes(label = scales::percent(frac_zero, accuracy = 0.1)),
              hjust = -0.1, size = 3) +
    coord_flip() +
    theme_bw() +
    scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.15))) +
    labs(
        title = "Fraction of genes with zero total counts per sub-batch",
        x = "Sub-batch", y = "Fraction of genes with sum = 0"
    )
```

## 6. Summary

```{r summary}
final_intersection_size <- length(Reduce(intersect, gene_sets))
cat("Total genes in dataset:", nrow(sce), "\n")
cat("Number of sub-batches:", length(batches), "\n")
cat("Smallest sub-batch:", names(which.min(sizes)), "with", min(sizes), "expressed genes\n")
cat("Strict intersection across all", length(batches), "sub-batches:", final_intersection_size, "genes\n")
cat("\nConclusion: anglemania's strict gene intersection across", length(batches),
    "sub-batches reduces the candidate gene pool to", final_intersection_size,
    "genes before any angle computation, explaining the 2-gene output.\n")
```
