---
title: "pyanglemania_test_visualization"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../output/")})
output:
    html_document:
      toc: yes
      toc_float:
        collapsed: no
        smooth_scroll: no
      df_print: paged

---

Smoke-test visualization for `config_pyanglemania_test.yml`: a single splatter-simulated
sample (3 batches, 3 groups, 1000 cells/group, de.facScale 0.1). With only one sample,
the per-condition heatmaps and paired Wilcoxon tests used in `bm_simulated_visualization.Rmd`
/ `compare_gene_selections.Rmd` don't apply -- this just sanity-checks that `angl` and `hvg`
produced sensible scores and gene sets for this one run.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.height = 6)
knitr::opts_chunk$set(fig.width = 8)
```

```{r, load libraries}
.libPaths(.Library)
library(tidyverse)
library(yaml)
library(rhdf5)
library(scales)
library(glue)
```

# Load config
```{r}
cfg <- read_yaml("/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/scripts/config_files/config_pyanglemania_test.yml")
r_era_cfg <- read_yaml("/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/scripts/config_files/config_varying_batch_biovar_facScale0.1.yml")
```

# Load metrics
```{r}
metrics_path <- cfg$visualization$metrics_file
samplesheet_path <- cfg$samplesheet
preprocessed_dir <- cfg$visualization$preprocessed_dir
outplot_bm <- cfg$visualization$outplot_dir_bm_visualization
outplot_cgs <- cfg$visualization$outplot_dir_compare_gene_selection
for (d in c(outplot_bm, outplot_cgs)) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE, showWarnings = FALSE)
}

samplesheet <- read.delim(samplesheet_path)
metrics <- read.delim(metrics_path) %>%
  left_join(samplesheet %>% select(-original_h5ad), by = join_by(sample == sample_name)) %>%
  mutate(
    gene_selection = factor(gene_selection, levels = c("hvg", "angl")),
    integration_method = factor(integration_method)
  )
metrics

r_era_sample <- "batch.facLoc0.1_de.facLoc0.1_nbatch4_ngroup5_groupCells1000"
r_era_metrics <- read.delim(r_era_cfg$visualization$metrics_file) %>%
  filter(sample == r_era_sample, gene_selection == "angl") %>%
  mutate(gene_selection = "r_angl")

metrics_extended <- bind_rows(metrics, r_era_metrics) %>%
  mutate(
    gene_selection = factor(gene_selection, levels = c("hvg", "angl", "r_angl")),
    integration_method = factor(integration_method)
  )
metrics_extended
```

# Integration scores: angl vs hvg vs r_angl
```{r}
p <- metrics_extended %>%
  select(gene_selection, integration_method, Total, Bio.conservation, Batch.correction) %>%
  pivot_longer(cols = c(Total, Bio.conservation, Batch.correction), names_to = "score_type", values_to = "score") %>%
  mutate(score_type = factor(score_type, levels = c("Total", "Bio.conservation", "Batch.correction"))) %>%
  ggplot(aes(x = gene_selection, y = score, fill = gene_selection)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = round(score, 2)), vjust = -0.3) +
  facet_grid(score_type ~ integration_method) +
  scale_fill_manual(values = c("angl" = "#64487A", "r_angl" = "#A67DB8", "hvg" = "grey")) +
  ylim(0, 1) +
  theme_bw() +
  labs(title = cfg$name, x = NULL, y = "Score") +
  theme(text = element_text(size = 14), legend.position = "none")
p
ggsave(file.path(outplot_bm, "scores_barplot.pdf"), p, width = 8, height = 8)
```

## Individual metrics
```{r individual_metrics_dotplot, fig.width=10, fig.height=10}
bio_metrics   <- c("Silhouette.label", "Isolated.labels", "bNMI", "cLISI", "ldfDiff")
batch_metrics <- c("PCR.comparison", "iLISI", "cms", "BRAS")

p_dots <- metrics_extended %>%
  select(-any_of(c("Total", "Bio.conservation", "Batch.correction"))) %>%
  pivot_longer(cols = where(is.numeric), names_to = "metric", values_to = "value") %>%
  mutate(metric = factor(metric, levels = c(bio_metrics, batch_metrics))) %>%
  ggplot(aes(x = gene_selection, y = value, color = gene_selection)) +
  geom_point(size = 4) +
  facet_grid(metric ~ integration_method, scales = "free_y") +
  scale_color_manual(values = c("angl" = "#64487A", "r_angl" = "#A67DB8", "hvg" = "grey40")) +
  theme_bw() +
  labs(title = cfg$name, x = NULL, y = "Score") +
  theme(text = element_text(size = 12), legend.position = "none",
        strip.text.y = element_text(angle = 0))
p_dots
ggsave(file.path(outplot_bm, "individual_metrics_dotplot.pdf"), p_dots, width = 10, height = 10)
```

# Compare gene selections: angl vs hvg
```{r}
sample_name <- samplesheet$sample_name[1]
angl_genes <- read.delim(file.path(preprocessed_dir, paste0(sample_name, "_angl.tsv"))) %>% pull(hgnc_symbol)
hvg_genes  <- read.delim(file.path(preprocessed_dir, paste0(sample_name, "_hvg.tsv"))) %>% pull(hgnc_symbol)
intersection <- intersect(angl_genes, hvg_genes)

glue(
  "angl: {length(angl_genes)} genes | hvg: {length(hvg_genes)} genes | ",
  "intersection: {length(intersection)} ({round(100 * length(intersection) / length(angl_genes), 1)}% of angl)"
)
```

## Compare against the original R `anglemania` implementation

`pyanglemania` is a GPU-accelerated Python port of the original R `anglemania` package.
As a migration sanity-check, we compare its gene selection against R-era results from a
previous benchmark run (`facScale0.1 / groupCells1000`). The test sample used above
(nbatch=3, ngroup=3) was not included in that earlier run; the closest available sample
with the same simulation parameters (batch.facLoc=0.1, de.facLoc=0.1) is
`nbatch4_ngroup5_groupCells1000`, which is used here. Because the two datasets are
independent splatter simulations, the overlap quantifies method reproducibility across
similar conditions, not same-dataset agreement.

```{r load_r_era_genes}
r_era_preprocessed_dir <- r_era_cfg$visualization$preprocessed_dir

r_angl_genes <- read.delim(
  file.path(r_era_preprocessed_dir, paste0(r_era_sample, "_angl.tsv"))
) %>% pull(hgnc_symbol)

glue("R-era anglemania ({r_era_sample}): {length(r_angl_genes)} genes")
```

```{r compare_r_vs_py}
r_vs_py <- intersect(r_angl_genes, angl_genes)
glue(
  "R anglemania (nbatch2/ngroup2): {length(r_angl_genes)} genes | ",
  "pyanglemania (nbatch3/ngroup3): {length(angl_genes)} genes | ",
  "intersection: {length(r_vs_py)} ({round(100 * length(r_vs_py) / length(r_angl_genes), 1)}% of R genes also selected by Python)"
)
```

## Quantify overlap between all gene selections

Pairwise and three-way overlap between `hvg`, the Python port (`angl`), and the original R
implementation (`r_angl`).

```{r overlap_table}
gene_sets <- list(hvg = hvg_genes, angl = angl_genes, r_angl = r_angl_genes)

pairwise_overlap <- combn(names(gene_sets), 2, function(pair) {
  a <- gene_sets[[pair[1]]]
  b <- gene_sets[[pair[2]]]
  inter <- intersect(a, b)
  tibble(
    set_a = pair[1], set_b = pair[2],
    n_a = length(a), n_b = length(b),
    n_intersection = length(inter),
    jaccard = length(inter) / length(union(a, b)),
    pct_of_a = 100 * length(inter) / length(a),
    pct_of_b = 100 * length(inter) / length(b)
  )
}, simplify = FALSE) %>% bind_rows()

pairwise_overlap

glue("Genes selected by all three methods: {length(Reduce(intersect, gene_sets))}")
```

```{r upset_plot, fig.width=8, fig.height=5}
all_genes <- unique(unlist(gene_sets))
binary_matrix <- sapply(gene_sets, function(s) all_genes %in% s)
rownames(binary_matrix) <- all_genes

comb_mat <- ComplexHeatmap::make_comb_mat(binary_matrix, mode = "distinct")
ComplexHeatmap::UpSet(
  comb_mat,
  comb_order = order(ComplexHeatmap::comb_size(comb_mat), decreasing = TRUE),
  comb_col = c("red", "blue", "black")[ComplexHeatmap::comb_degree(comb_mat)]
)
```

## BTVR (Between-to-Total Variance Ratio) by gene selection
* anglemania genes (R or Python) are expected to have a higher BTVR (more biological, less
  technical variance) than HVGs, and the two anglemania implementations are expected to look
  similar to each other.
```{r}
rowdata <- h5read(samplesheet$file_path[1], "/var") %>% as.data.frame() %>% column_to_rownames("X_index")
h5closeAll()

bio_var <- apply(dplyr::select(rowdata, starts_with("DEFacGroup")), 1, var)
tech_var <- apply(dplyr::select(rowdata, starts_with("BatchFacBatch")), 1, var)

btvr_base <- tibble(
  gene = rownames(rowdata),
  BTVR = rescale(log((bio_var + 1e-15) / (tech_var + 1e-15)))
)

# A gene can belong to more than one selection method, so each method's membership is
# computed independently (a gene in both gets a row in each group) rather than via a single
# case_when label, which would silently drop it from all but the first matching group.
btvr <- purrr::imap_dfr(gene_sets, function(genes, name) {
  btvr_base %>% filter(gene %in% genes) %>% mutate(gene_selection = name)
}) %>%
  mutate(gene_selection = factor(gene_selection, levels = c("hvg", "angl", "r_angl")))

p <- btvr %>%
  ggplot(aes(x = gene_selection, y = BTVR, fill = gene_selection)) +
  geom_violin() +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  scale_fill_manual(values = c("angl" = "#64487A", "r_angl" = "#A67DB8", "hvg" = "grey")) +
  ylim(0, 1) +
  theme_bw() +
  labs(title = glue("{cfg$name}: BTVR by gene selection"), x = NULL) +
  theme(text = element_text(size = 14), legend.position = "none")
p
ggsave(file.path(outplot_cgs, "BTVR_violin.pdf"), p, width = 7, height = 5)

btvr %>%
  group_by(gene_selection) %>%
  summarise(mean_BTVR = mean(BTVR), median_BTVR = median(BTVR), n = n())
```

## BTVR: full gene distribution with selection-group landmarks
```{r btvr_full_distribution, fig.width=8, fig.height=5}
btvr_summary <- btvr %>%
  group_by(gene_selection) %>%
  summarise(mean_BTVR = mean(BTVR), median_BTVR = median(BTVR))

p_dist <- btvr_base %>%
  ggplot(aes(x = BTVR)) +
  geom_histogram(bins = 60, fill = "lightgrey", color = "white") +
  geom_vline(data = btvr_summary,
             aes(xintercept = median_BTVR, color = gene_selection, linetype = "median"),
             linewidth = 1) +
  geom_vline(data = btvr_summary,
             aes(xintercept = mean_BTVR, color = gene_selection, linetype = "mean"),
             linewidth = 1) +
  scale_color_manual(values = c("angl" = "#64487A", "r_angl" = "#A67DB8", "hvg" = "grey40")) +
  scale_linetype_manual(values = c("median" = "solid", "mean" = "dashed")) +
  theme_bw() +
  labs(
    title = glue("{cfg$name}: BTVR distribution (all genes)"),
    subtitle = "Solid = median, dashed = mean per gene selection",
    x = "BTVR (rescaled log bio/tech variance ratio)", y = "Count",
    color = "Gene selection", linetype = NULL
  ) +
  theme(text = element_text(size = 13))
p_dist
ggsave(file.path(outplot_cgs, "BTVR_full_distribution.pdf"), p_dist, width = 8, height = 5)
```

# Session Info
```{r}
sessionInfo()
```
