---
title: "pyanglemania_test_visualization"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../output/")})
output:
    html_document:
      toc: yes
      toc_float:
        collapsed: no
        smooth_scroll: no
      df_print: paged

---

Smoke-test visualization for `config_pyanglemania_test.yml`: a single splatter-simulated
sample (3 batches, 3 groups, 1000 cells/group, de.facScale 0.1). With only one sample,
the per-condition heatmaps and paired Wilcoxon tests used in `bm_simulated_visualization.Rmd`
/ `compare_gene_selections.Rmd` don't apply -- this just sanity-checks that `angl` and `hvg`
produced sensible scores and gene sets for this one run.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.height = 6)
knitr::opts_chunk$set(fig.width = 8)
```

```{r, load libraries}
.libPaths(.Library)
library(tidyverse)
library(yaml)
library(rhdf5)
library(scales)
library(glue)
```

# Load config
```{r}
cfg <- read_yaml("/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark/scripts/config_files/config_pyanglemania_test.yml")
```

# Load metrics
```{r}
metrics_path <- cfg$visualization$metrics_file
samplesheet_path <- cfg$samplesheet
preprocessed_dir <- cfg$visualization$preprocessed_dir
outplot_bm <- cfg$visualization$outplot_dir_bm_visualization
outplot_cgs <- cfg$visualization$outplot_dir_compare_gene_selection
for (d in c(outplot_bm, outplot_cgs)) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE, showWarnings = FALSE)
}

samplesheet <- read.delim(samplesheet_path)
metrics <- read.delim(metrics_path) %>%
  left_join(samplesheet %>% select(-original_h5ad), by = join_by(sample == sample_name)) %>%
  mutate(
    gene_selection = factor(gene_selection, levels = c("hvg", "angl")),
    integration_method = factor(integration_method)
  )
metrics
```

# Integration scores: angl vs hvg
```{r}
p <- metrics %>%
  select(gene_selection, integration_method, Total, Bio.conservation, Batch.correction) %>%
  pivot_longer(cols = c(Total, Bio.conservation, Batch.correction), names_to = "score_type", values_to = "score") %>%
  mutate(score_type = factor(score_type, levels = c("Total", "Bio.conservation", "Batch.correction"))) %>%
  ggplot(aes(x = gene_selection, y = score, fill = gene_selection)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = round(score, 2)), vjust = -0.3) +
  facet_grid(score_type ~ integration_method) +
  scale_fill_manual(values = c("angl" = "#64487A", "hvg" = "grey")) +
  ylim(0, 1) +
  theme_bw() +
  labs(title = cfg$name, x = NULL, y = "Score") +
  theme(text = element_text(size = 14), legend.position = "none")
p
ggsave(file.path(outplot_bm, "scores_barplot.pdf"), p, width = 8, height = 8)
```

# Compare gene selections: angl vs hvg
```{r}
sample_name <- samplesheet$sample_name[1]
angl_genes <- read.delim(file.path(preprocessed_dir, paste0(sample_name, "_angl.tsv"))) %>% pull(hgnc_symbol)
hvg_genes  <- read.delim(file.path(preprocessed_dir, paste0(sample_name, "_hvg.tsv"))) %>% pull(hgnc_symbol)
intersection <- intersect(angl_genes, hvg_genes)

glue(
  "angl: {length(angl_genes)} genes | hvg: {length(hvg_genes)} genes | ",
  "intersection: {length(intersection)} ({round(100 * length(intersection) / length(angl_genes), 1)}% of angl)"
)
```

## Compare against the original R `anglemania` implementation

`pyanglemania` is a from-scratch GPU-capable Python port of the original R `anglemania`
package. As a migration sanity-check, this runs the legacy R package directly on the exact
same input file and parameters used above, and compares the resulting gene set against the
Python (`angl`) selection above. No GPU is needed for the R implementation, but it is
considerably slower than the Python port (~5 min vs ~1.5 min on this dataset on CPU) -- this
chunk is the slow part of the notebook.

```{r run_r_anglemania}
suppressPackageStartupMessages({
  library(anndataR)
  library(anglemania)
})

ad_r <- anndataR::read_h5ad(samplesheet$file_path[1])
sce <- ad_r$to_SingleCellExperiment()
sce <- anglemania(
  sce,
  batch_key = cfg$batch_key,
  dataset_key = NULL,
  min_cells_per_gene = 1,
  max_n_genes = cfg$n_genes,
  method = cfg$anglemania_mode,
  permutation_function = cfg$permutation_function
)
r_angl_genes <- get_anglemania_genes(sce)
length(r_angl_genes)
```

```{r compare_r_vs_py}
r_vs_py <- intersect(r_angl_genes, angl_genes)
glue(
  "R anglemania: {length(r_angl_genes)} genes | pyanglemania: {length(angl_genes)} genes | ",
  "intersection: {length(r_vs_py)} ({round(100 * length(r_vs_py) / length(r_angl_genes), 1)}% agreement)"
)
```

## Quantify overlap between all gene selections

Pairwise and three-way overlap between `hvg`, the Python port (`angl`), and the original R
implementation (`r_angl`).

```{r overlap_table}
gene_sets <- list(hvg = hvg_genes, angl = angl_genes, r_angl = r_angl_genes)

pairwise_overlap <- combn(names(gene_sets), 2, function(pair) {
  a <- gene_sets[[pair[1]]]
  b <- gene_sets[[pair[2]]]
  inter <- intersect(a, b)
  tibble(
    set_a = pair[1], set_b = pair[2],
    n_a = length(a), n_b = length(b),
    n_intersection = length(inter),
    jaccard = length(inter) / length(union(a, b)),
    pct_of_a = 100 * length(inter) / length(a),
    pct_of_b = 100 * length(inter) / length(b)
  )
}, simplify = FALSE) %>% bind_rows()

pairwise_overlap

glue("Genes selected by all three methods: {length(Reduce(intersect, gene_sets))}")
```

```{r upset_plot, fig.width=8, fig.height=5}
all_genes <- unique(unlist(gene_sets))
binary_matrix <- sapply(gene_sets, function(s) all_genes %in% s)
rownames(binary_matrix) <- all_genes

comb_mat <- ComplexHeatmap::make_comb_mat(binary_matrix, mode = "distinct")
ComplexHeatmap::UpSet(
  comb_mat,
  comb_order = order(ComplexHeatmap::comb_size(comb_mat), decreasing = TRUE),
  comb_col = c("red", "blue", "black")[ComplexHeatmap::comb_degree(comb_mat)]
)
```

## BTVR (Between-to-Total Variance Ratio) by gene selection
* anglemania genes (R or Python) are expected to have a higher BTVR (more biological, less
  technical variance) than HVGs, and the two anglemania implementations are expected to look
  similar to each other.
```{r}
rowdata <- h5read(samplesheet$file_path[1], "/var") %>% as.data.frame() %>% column_to_rownames("X_index")
h5closeAll()

bio_var <- apply(dplyr::select(rowdata, starts_with("DEFacGroup")), 1, var)
tech_var <- apply(dplyr::select(rowdata, starts_with("BatchFacBatch")), 1, var)

btvr_base <- tibble(
  gene = rownames(rowdata),
  BTVR = rescale(log((bio_var + 1e-15) / (tech_var + 1e-15)))
)

# A gene can belong to more than one selection method, so each method's membership is
# computed independently (a gene in both gets a row in each group) rather than via a single
# case_when label, which would silently drop it from all but the first matching group.
btvr <- purrr::imap_dfr(gene_sets, function(genes, name) {
  btvr_base %>% filter(gene %in% genes) %>% mutate(gene_selection = name)
}) %>%
  mutate(gene_selection = factor(gene_selection, levels = c("hvg", "angl", "r_angl")))

p <- btvr %>%
  ggplot(aes(x = gene_selection, y = BTVR, fill = gene_selection)) +
  geom_violin() +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  scale_fill_manual(values = c("angl" = "#64487A", "r_angl" = "#A67DB8", "hvg" = "grey")) +
  ylim(0, 1) +
  theme_bw() +
  labs(title = glue("{cfg$name}: BTVR by gene selection"), x = NULL) +
  theme(text = element_text(size = 14), legend.position = "none")
p
ggsave(file.path(outplot_cgs, "BTVR_violin.pdf"), p, width = 7, height = 5)

btvr %>%
  group_by(gene_selection) %>%
  summarise(mean_BTVR = mean(BTVR), median_BTVR = median(BTVR), n = n())
```

# Session Info
```{r}
sessionInfo()
```
