---
title: "Sanity Check: Top vs Bottom BTVR Genes for Integration"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../output/")})
output:
    html_document:
      toc: yes
      toc_float:
        collapsed: no
        smooth_scroll: no
      df_print: paged
---

**Rationale**: BTVR (Between-to-Total Variance Ratio) captures the ratio of
biological to technical gene variance. Genes with high BTVR are biologically
informative and batch-robust — exactly what good feature selection for
integration should target. This notebook verifies this by running scvi and
scanorama integration on the top 500 vs bottom 500 BTVR genes from a simulated
dataset (where ground-truth variance factors are known), then comparing
integration metrics between the two groups. Good integration metrics with
top-BTVR genes and poor metrics with bottom-BTVR genes confirm the hypothesis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
knitr::opts_chunk$set(fig.height = 7)
knitr::opts_chunk$set(fig.width = 12)
```

# Parameters

```{r params}
.libPaths(.Library)

# ---- paths ----
base_dir   <- "/home/akollot/projects/CRC1588/dataset_integration/anglemania_benchmark"
out_dir    <- file.path(base_dir, "output", "sanity_check_btvr_top_vs_bottom")
scripts_dir <- file.path(base_dir, "scripts")
pipeline_dir <- file.path(scripts_dir, "pipeline_scripts")

# ---- simulation ----
n_genes      <- 5000
n_batches    <- 3
n_groups     <- 3
cells_per_group <- 500   # cells per group per batch
batch_facLoc <- 0.3      # moderate batch effects
de_facLoc    <- 0.3      # moderate biological signal
set_seed     <- 42

# ---- gene selection ----
n_genes_select <- 500    # top/bottom N by BTVR

# ---- integration ----
batch_key  <- "Batch"
label_key  <- "Group"
integration_methods <- c("scvi", "scanorama")
gene_sets   <- c("top_btvr", "bottom_btvr")
```

# Setup

```{r libraries}
suppressPackageStartupMessages({
    library(splatter)
    library(zellkonverter)
    library(SingleCellExperiment)
    library(dplyr)
    library(tidyr)
    library(ggplot2)
    library(scales)
    library(patchwork)
    library(glue)
    library(uwot)
})

source(file.path(scripts_dir, "R", "ggplot_utils.R"))
today <- format(Sys.Date(), "%Y%m%d")
```

```{r create_dirs}
dirs_needed <- c(
    out_dir,
    file.path(out_dir, "preprocessed"),
    file.path(out_dir, "plots")
)
for (d in dirs_needed) {
    if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}
for (method in integration_methods) {
    for (gs in gene_sets) {
        d <- file.path(out_dir, "integration", "sim", method)
        if (!dir.exists(d)) dir.create(d, recursive = TRUE)
        d <- file.path(out_dir, "metrics", "sim", method)
        if (!dir.exists(d)) dir.create(d, recursive = TRUE)
    }
}
```

# Simulate data

```{r simulate, cache=TRUE}
sim_h5ad <- file.path(out_dir, "sim.h5ad")

set.seed(set_seed)
sim <- splatSimulate(
    batchCells  = rep(cells_per_group * n_groups, n_batches),
    group.prob  = rep(1 / n_groups, n_groups),
    batch.facLoc  = batch_facLoc,
    batch.facScale = 0.1,
    method      = "groups",
    verbose     = FALSE,
    out.prob    = 0.001,
    de.prob     = 0.1,
    de.facLoc   = de_facLoc,
    de.facScale = 0.4,
    bcv.common  = 0.1,
    nGenes      = n_genes
)

message("Simulated: ", ncol(sim), " cells × ", nrow(sim), " genes")
message("Batches: ", paste(unique(sim$Batch), collapse=", "))
message("Groups: ",  paste(unique(sim$Group),  collapse=", "))

if (!file.exists(sim_h5ad)) {
    # Write raw counts to X for compatibility with integration.py
    writeH5AD(sim, file = sim_h5ad, X_name = "counts")
    message("Wrote simulated data to: ", sim_h5ad)
} else {
    message("Simulated h5ad already exists: ", sim_h5ad)
}
```

# Compute BTVR from ground truth

BTVR = log( (BioVar + ε) / (TechVar + ε) ), where:
- **BioVar** = variance of differential expression factors across groups (`DEFacGroup*`)
- **TechVar** = variance of batch factors across batches (`BatchFacBatch*`)

Higher BTVR → more biologically variable, less batch-driven → better for integration.

```{r compute_btvr}
compute_btvr <- function(rowdata) {
    bio_cols  <- dplyr::select(as.data.frame(rowdata), starts_with("DEFacGroup"))
    tech_cols <- dplyr::select(as.data.frame(rowdata), starts_with("BatchFacBatch"))

    raw_BioVar  <- apply(bio_cols,  1, var)
    raw_TechVar <- apply(tech_cols, 1, var)

    # log-ratio: high = biologically variable, not technically variable
    BTVR_raw <- log((raw_BioVar + 1e-10) / (raw_TechVar + 1e-10))

    tibble(
        gene       = rownames(rowdata),
        raw_BioVar = raw_BioVar,
        raw_TechVar = raw_TechVar,
        BTVR       = BTVR_raw,
        BTVR_scaled = rescale(BTVR_raw)
    )
}

btvr_df <- compute_btvr(rowData(sim))
btvr_df <- btvr_df[order(btvr_df$BTVR, decreasing = TRUE), ]
head(btvr_df)
```

# Visualise BTVR distribution

```{r plot_btvr_dist, fig.height=5}
p_dist <- ggplot(btvr_df, aes(x = BTVR)) +
    geom_histogram(bins = 80, fill = "#64487A", color = "white", linewidth = 0.2) +
    geom_vline(
        xintercept = btvr_df$BTVR[n_genes_select],
        linetype = "dashed", color = "steelblue", linewidth = 0.8
    ) +
    geom_vline(
        xintercept = btvr_df$BTVR[nrow(btvr_df) - n_genes_select + 1],
        linetype = "dashed", color = "tomato", linewidth = 0.8
    ) +
    annotate("text", x = btvr_df$BTVR[n_genes_select] + 0.05,
             y = Inf, vjust = 1.5, hjust = 0, size = 3.5, color = "steelblue",
             label = glue("top {n_genes_select} cutoff")) +
    annotate("text", x = btvr_df$BTVR[nrow(btvr_df) - n_genes_select + 1] - 0.05,
             y = Inf, vjust = 1.5, hjust = 1, size = 3.5, color = "tomato",
             label = glue("bottom {n_genes_select} cutoff")) +
    labs(
        title = glue("BTVR distribution across {n_genes} simulated genes"),
        x = "log( BioVar / TechVar )",
        y = "Count"
    )
p_dist

ggsave(
    file.path(out_dir, "plots", glue("{today}_btvr_distribution.pdf")),
    p_dist, width = 8, height = 4
)
```

```{r plot_biovar_vs_techvar, fig.height=5}
btvr_df_labeled <- btvr_df |>
    mutate(gene_set = case_when(
        row_number() <= n_genes_select          ~ "top BTVR",
        row_number() > n() - n_genes_select     ~ "bottom BTVR",
        TRUE                                    ~ "other"
    ))

p_scatter <- ggplot(btvr_df_labeled, aes(x = raw_TechVar, y = raw_BioVar, color = gene_set)) +
    geom_point(size = 0.4, alpha = 0.5) +
    scale_color_manual(values = c("top BTVR" = "steelblue", "bottom BTVR" = "tomato", "other" = "grey70")) +
    scale_x_log10() + scale_y_log10() +
    labs(
        title = "BioVar vs TechVar (ground truth)",
        x = "TechVar (BatchFacBatch variance)",
        y = "BioVar (DEFacGroup variance)",
        color = "Gene set"
    )
p_scatter

ggsave(
    file.path(out_dir, "plots", glue("{today}_biovar_vs_techvar.pdf")),
    p_scatter, width = 7, height = 5
)
```

# Select gene sets

```{r select_genes}
top_genes    <- btvr_df$gene[seq_len(n_genes_select)]
bottom_genes <- btvr_df$gene[seq(nrow(btvr_df) - n_genes_select + 1, nrow(btvr_df))]

message("Top BTVR genes: BTVR range [",
        round(min(btvr_df$BTVR[seq_len(n_genes_select)]), 2), ", ",
        round(max(btvr_df$BTVR[seq_len(n_genes_select)]), 2), "]")
message("Bottom BTVR genes: BTVR range [",
        round(min(btvr_df$BTVR[seq(nrow(btvr_df) - n_genes_select + 1, nrow(btvr_df))]), 2), ", ",
        round(max(btvr_df$BTVR[seq(nrow(btvr_df) - n_genes_select + 1, nrow(btvr_df))]), 2), "]")

# Write gene list TSVs (same format as prepare_inputs.R output)
gene_lists <- list(
    top_btvr    = top_genes,
    bottom_btvr = bottom_genes
)

gene_tsv_paths <- list()
for (gs in names(gene_lists)) {
    path <- file.path(out_dir, "preprocessed", glue("sim_{gs}.tsv"))
    write.table(
        data.frame(hgnc_symbol = gene_lists[[gs]]),
        path, sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE
    )
    gene_tsv_paths[[gs]] <- path
    message("Wrote gene list: ", path)
}
```

# Run integrations

Integration is done via `integration.py` (same script as in the pipeline), which
normalises internally and stores the embedding in `obsm["X_emb"]`.

```{r integrate}
integrated_paths <- list()

for (method in integration_methods) {
    for (gs in gene_sets) {
        outfile <- file.path(out_dir, "integration", "sim", method,
                             glue("sim_{gs}.h5ad"))
        integrated_paths[[glue("{method}_{gs}")]] <- outfile

        if (file.exists(outfile)) {
            message("[skip] Integration already done: ", basename(outfile))
            next
        }

        message("\n>>> Integrating: method=", method, "  gene_set=", gs)
        ret <- system2(
            "python3",
            args = c(
                file.path(pipeline_dir, "integration.py"),
                glue("--infile={sim_h5ad}"),
                glue("--feature_subset={gene_tsv_paths[[gs]]}"),
                glue("--outfile={outfile}"),
                glue("--integration_method={method}"),
                glue("--batch_key={batch_key}"),
                glue("--label_key={label_key}")
            ),
            stdout = TRUE, stderr = TRUE
        )
        message(paste(ret, collapse = "\n"))
        if (!file.exists(outfile)) {
            warning("Integration output not found after running: ", outfile)
        }
    }
}
message("All integrations done.")
```

# Run metrics

Metrics are computed for each integration using the pipeline scripts directly.

```{r metrics}
metric_paths <- list()

run_metric <- function(metric_name, cmd, args, outfile) {
    if (file.exists(outfile)) {
        message("[skip] Metric already computed: ", basename(outfile))
        return(invisible(NULL))
    }
    message("\n>>> Computing ", metric_name, ": ", basename(outfile))
    ret <- system2(cmd, args = args, stdout = TRUE, stderr = TRUE)
    message(paste(ret, collapse = "\n"))
    if (!file.exists(outfile)) {
        warning(metric_name, " output not found: ", outfile)
    }
}

for (method in integration_methods) {
    for (gs in gene_sets) {
        integrated_h5ad <- integrated_paths[[glue("{method}_{gs}")]]
        metrics_dir     <- file.path(out_dir, "metrics", "sim", method)
        sample_label    <- "sim"

        if (!file.exists(integrated_h5ad)) {
            message("[skip] No integration output found for ", method, " / ", gs)
            next
        }

        # ---- scib_metrics ----
        scib_out <- file.path(metrics_dir, glue("sim_{gs}_scibmetrics.tsv"))
        metric_paths[[glue("scib_{method}_{gs}")]] <- scib_out
        run_metric("scib_metrics", "python3",
            args = c(
                file.path(pipeline_dir, "metrics-scib_metrics.py"),
                glue("--original_h5ad={sim_h5ad}"),
                glue("--integrated_h5ad={integrated_h5ad}"),
                glue("--outfile={scib_out}"),
                glue("--sample={sample_label}"),
                glue("--gene_selection={gs}"),
                glue("--integration_method={method}"),
                glue("--batch_key={batch_key}"),
                glue("--label_key={label_key}")
            ),
            outfile = scib_out
        )

        # ---- bNMI ----
        bnmi_out <- file.path(metrics_dir, glue("sim_{gs}_bnmi.tsv"))
        metric_paths[[glue("bnmi_{method}_{gs}")]] <- bnmi_out
        run_metric("bNMI", "python3",
            args = c(
                file.path(pipeline_dir, "metrics-bNMI.py"),
                glue("--integrated_h5ad={integrated_h5ad}"),
                glue("--outfile={bnmi_out}"),
                glue("--sample={sample_label}"),
                glue("--gene_selection={gs}"),
                glue("--integration_method={method}"),
                glue("--label_key={label_key}")
            ),
            outfile = bnmi_out
        )

        # ---- CMS ----
        cms_out <- file.path(metrics_dir, glue("sim_{gs}_cms.tsv"))
        metric_paths[[glue("cms_{method}_{gs}")]] <- cms_out
        run_metric("CMS", "Rscript",
            args = c(
                file.path(pipeline_dir, "metrics-cms.R"),
                "--integrated_h5ad", integrated_h5ad,
                "--batch_key",       batch_key,
                "--outfile",         cms_out,
                "--sample",          sample_label,
                "--gene_selection",  gs,
                "--integration_method", method
            ),
            outfile = cms_out
        )

        # ---- ldfDiff ----
        ldf_out <- file.path(metrics_dir, glue("sim_{gs}_ldfDiff.tsv"))
        metric_paths[[glue("ldf_{method}_{gs}")]] <- ldf_out
        run_metric("ldfDiff", "Rscript",
            args = c(
                file.path(pipeline_dir, "metrics-ldfDiff.R"),
                "--original_h5ad",   sim_h5ad,
                "--integrated_h5ad", integrated_h5ad,
                "--batch_key",       batch_key,
                "--label_key",       label_key,
                "--outfile",         ldf_out,
                "--sample",          sample_label,
                "--gene_selection",  gs,
                "--integration_method", method
            ),
            outfile = ldf_out
        )
    }
}
message("All metrics done.")
```

# Combine metrics

```{r combine_metrics}
combined_out <- file.path(out_dir, "metrics", "combined_metrics.tsv")

existing_scib <- Filter(file.exists, unname(metric_paths[grepl("^scib_", names(metric_paths))]))
existing_bnmi <- Filter(file.exists, unname(metric_paths[grepl("^bnmi_", names(metric_paths))]))
existing_cms  <- Filter(file.exists, unname(metric_paths[grepl("^cms_",  names(metric_paths))]))
existing_ldf  <- Filter(file.exists, unname(metric_paths[grepl("^ldf_",  names(metric_paths))]))

if (length(existing_scib) > 0 && length(existing_bnmi) > 0 &&
    length(existing_cms) > 0  && length(existing_ldf) > 0) {

    scib_arg <- paste(existing_scib, collapse = " ")
    bnmi_arg <- paste(existing_bnmi, collapse = " ")
    cms_arg  <- paste(existing_cms,  collapse = " ")
    ldf_arg  <- paste(existing_ldf,  collapse = " ")

    ret <- system2(
        "python3",
        args = c(
            file.path(pipeline_dir, "combine_metrics.py"),
            "--scib_metrics", existing_scib,
            "--bnmi",         existing_bnmi,
            "--cms",          existing_cms,
            "--ldfdiff",      existing_ldf,
            "--outfile",      combined_out
        ),
        stdout = TRUE, stderr = TRUE
    )
    message(paste(ret, collapse = "\n"))
} else {
    message("Not all metric files available yet — skipping combine step.")
    message("Missing: scib=", length(existing_scib), " bnmi=", length(existing_bnmi),
            " cms=", length(existing_cms), " ldf=", length(existing_ldf))
}
```

# Visualise metrics

```{r load_metrics}
if (!file.exists(combined_out)) {
    stop("Combined metrics file not found: ", combined_out,
         "\nMake sure all integrations and metrics ran successfully.")
}

metrics <- read.delim(combined_out) |>
    mutate(
        gene_selection   = factor(gene_selection,   levels = c("top_btvr", "bottom_btvr")),
        integration_method = factor(integration_method, levels = integration_methods),
        condition        = glue("{integration_method} / {gene_selection}")
    )

metrics
```

```{r plot_metrics_bar, fig.height=8}
metric_cols <- c("Silhouette label", "Isolated labels", "bNMI", "cLISI", "ldfDiff",
                 "PCR comparison", "iLISI", "cms", "BRAS",
                 "Bio conservation", "Batch correction", "Total")
metric_cols_present <- intersect(metric_cols, colnames(metrics))

metrics_long <- metrics |>
    select(integration_method, gene_selection, all_of(metric_cols_present)) |>
    pivot_longer(
        cols      = all_of(metric_cols_present),
        names_to  = "metric",
        values_to = "score"
    ) |>
    mutate(
        metric_type = case_when(
            metric %in% c("Silhouette label", "Isolated labels", "bNMI", "cLISI", "ldfDiff") ~ "Bio conservation",
            metric %in% c("PCR comparison", "iLISI", "cms", "BRAS")                           ~ "Batch correction",
            metric %in% c("Bio conservation", "Batch correction", "Total")                    ~ "Summary",
            TRUE ~ "Other"
        ),
        metric = factor(metric, levels = metric_cols_present),
        gene_selection = factor(gene_selection, levels = c("top_btvr", "bottom_btvr"),
                                labels = c("Top BTVR", "Bottom BTVR"))
    )

p_metrics <- ggplot(
        metrics_long |> filter(metric_type != "Summary"),
        aes(x = gene_selection, y = score, fill = gene_selection)
    ) +
    geom_col(position = "dodge", width = 0.7) +
    scale_fill_manual(values = c("Top BTVR" = "steelblue", "Bottom BTVR" = "tomato")) +
    facet_grid(metric_type ~ interaction(metric, integration_method), scales = "free_y") +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
        title = glue("Integration metrics: top {n_genes_select} vs bottom {n_genes_select} BTVR genes"),
        x = NULL, y = "Score (higher = better)", fill = "Gene selection"
    ) +
    theme(
        axis.text.x    = element_blank(),
        axis.ticks.x   = element_blank(),
        strip.text     = element_text(size = 7),
        legend.position = "bottom"
    )
p_metrics

ggsave(
    file.path(out_dir, "plots", glue("{today}_metrics_top_vs_bottom_btvr.pdf")),
    p_metrics, width = 16, height = 7
)
```

```{r plot_summary_metrics, fig.height=5}
p_summary <- ggplot(
        metrics_long |> filter(metric_type == "Summary"),
        aes(x = gene_selection, y = score, fill = gene_selection)
    ) +
    geom_col(width = 0.6) +
    geom_text(aes(label = round(score, 3)), vjust = -0.4, size = 3.5) +
    scale_fill_manual(values = c("Top BTVR" = "steelblue", "Bottom BTVR" = "tomato")) +
    scale_y_continuous(limits = c(0, 1.05)) +
    facet_grid(metric ~ integration_method) +
    labs(
        title = "Summary scores",
        x = NULL, y = "Score", fill = "Gene selection"
    ) +
    theme(legend.position = "none")
p_summary

ggsave(
    file.path(out_dir, "plots", glue("{today}_summary_scores_top_vs_bottom_btvr.pdf")),
    p_summary, width = 9, height = 7
)
```

# UMAPs

Load integrated embeddings, run 2D UMAP, and plot coloured by batch and cell type.

```{r umap_helper}
plot_umap_from_h5ad <- function(h5ad_path, title) {
    if (!file.exists(h5ad_path)) {
        message("File not found, skipping UMAP: ", h5ad_path)
        return(NULL)
    }
    adt <- anndataR::read_h5ad(h5ad_path)
    x_emb <- adt$obsm[["X_emb"]]
    if (is.null(x_emb)) {
        message("No X_emb found in: ", h5ad_path)
        return(NULL)
    }
    set.seed(42)
    umap_coords <- uwot::umap(x_emb, n_components = 2, n_threads = 4)
    df_umap <- data.frame(
        UMAP1 = umap_coords[, 1],
        UMAP2 = umap_coords[, 2],
        Batch = adt$obs[[batch_key]],
        Group = adt$obs[[label_key]]
    )
    p_batch <- ggplot(df_umap, aes(UMAP1, UMAP2, color = Batch)) +
        geom_point(size = 0.3, alpha = 0.4) +
        labs(title = paste0(title, "\n(by batch)")) +
        theme(legend.position = "right", plot.title = element_text(size = 9))

    p_group <- ggplot(df_umap, aes(UMAP1, UMAP2, color = Group)) +
        geom_point(size = 0.3, alpha = 0.4) +
        labs(title = paste0(title, "\n(by cell type)")) +
        theme(legend.position = "right", plot.title = element_text(size = 9))

    p_batch + p_group
}
```

```{r umaps, fig.height=6, results='asis'}
for (method in integration_methods) {
    cat(glue("\n\n## {method}\n\n"))
    for (gs in gene_sets) {
        integrated_h5ad <- integrated_paths[[glue("{method}_{gs}")]]
        gs_label <- if (gs == "top_btvr") glue("Top {n_genes_select} BTVR") else glue("Bottom {n_genes_select} BTVR")
        p <- plot_umap_from_h5ad(integrated_h5ad, glue("{method} — {gs_label}"))
        if (!is.null(p)) {
            cat(glue("\n\n### {gs_label}\n\n"))
            print(p)
            ggsave(
                file.path(out_dir, "plots", glue("{today}_umap_{method}_{gs}.pdf")),
                p, width = 10, height = 4
            )
        }
    }
}
```

# Session info

```{r session_info}
sessionInfo()
```
