---
title: "Simulate mixed-effects scenarios for gene selection stability"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../output/")})
output:
    html_document:
      toc: yes
      toc_float:
        collapsed: no
        smooth_scroll: no
      df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Base scenario: `batch.facLoc0.3_de.facLoc0.1_nbatch4_ngroup4_groupCells1000`
(facScale0.1) -- 4 batches, 4 cell-type groups, moderate batch effect,
moderate DE signal. This notebook derives all perturbed datasets needed for
the anglemania vs HVG gene-selection stability analysis:

1. **Leave-one-group-out (LOGO)** -- drop each of the 4 groups from the base
   simulation, one at a time (4 datasets).
2. **Leave-one-batch-out (LOBO)** -- drop each of the 4 batches from the base
   simulation, one at a time (4 datasets).
3. **Sparsity** -- a fresh splatter simulation with the same
   batch.facLoc/de.facLoc/nbatch/ngroup/groupCells parameters, but with
   `dropout.mid=3.5, dropout.shape=-1` (same dropout settings used for the
   `ngroup=3` reference in `with_dropout/own_sim`, which lands at ~88%
   sparsity there).
4. **Composition skew** -- unlike LOGO, all 4 groups stay present, just
   reweighted by subsampling cells per group per batch: (a) a graded skew
   (Group1/2/3/4 = 10/20/30/40%) and (b) one group made "rare" (Group1 = 1%,
   Groups 2/3/4 split the remaining 99% evenly). Both keep the same
   per-batch cell budget (`groupCells` = 1000) as the baseline, so total
   dataset size is unchanged -- only the group proportions differ.

The unperturbed baseline itself already exists at
`h5ad/facScale0.1/batch.facLoc0.3_de.facLoc0.1_nbatch4_ngroup4_groupCells1000.h5ad`
and is referenced directly (not copied). Only the perturbed datasets are
written to `h5ad/mixed_effects/`. All outputs are collected into a single
samplesheet for the benchmark pipeline.

```{r libraries, message=FALSE, warning=FALSE}
.libPaths(.Library)
suppressPackageStartupMessages({
    library(splatter)
    library(stringr)
    library(dplyr)
    library(tidyverse)
    library(tibble)
    library(anndataR)
    library(SingleCellExperiment)
})
```

```{r paths}
BASELINE_H5AD <- "/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/h5ad/facScale0.1/batch.facLoc0.3_de.facLoc0.1_nbatch4_ngroup4_groupCells1000.h5ad"
OUTDIR <- "/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/h5ad/mixed_effects"
dir.create(OUTDIR, recursive = TRUE, showWarnings = FALSE)

BASE_PARAMS <- list(
    batch.facLoc = 0.3,
    de.facLoc = 0.1,
    nbatch = 4,
    ngroup = 4,
    groupCells = 1000,
    batch.facScale = 0.1,
    de.facScale = 0.2,
    nGenes = 10000
)
```

## Simulation / save helpers

Reused verbatim from `simulate_single_cell_with_dropout.rmd` (dropout
generation) plus a generalization of its `save_sim()` helper that can save
any `SingleCellExperiment` (not just a fresh `splatSimulate()` output) --
needed to write out the LOGO/LOBO cell subsets of the baseline.

```{r helpers}
splatparams_to_list <- function(obj) {
    p <- metadata(obj)$Params
    slots <- slotNames(p)
    l <- setNames(lapply(slots, function(s) slot(p, s)), slots)
    return(l)
}

save_sce_as_h5ad <- function(sce, outfile, uns = NULL) {
    counts_mat <- t(counts(sce)) %>% as.matrix()
    obs_df <- colData(sce) %>% as.data.frame()
    var_df <- rowData(sce) %>% as.data.frame()
    if (is.null(uns)) uns <- S4Vectors::metadata(sce)

    message("Creating AnnData object")
    adata <- anndataR::AnnData(
        X = counts_mat,
        obs = obs_df,
        var = var_df,
        uns = uns
    )
    message(paste0("Saving ", outfile))
    anndataR::write_h5ad(adata, path = outfile, mode = "w", compression = "none")
}

create_and_save_sim <- function(
    sample_name,
    ngroup,
    nbatch,
    groupCells,
    batch.facLoc,
    de.facLoc,
    batch.facScale = 0.1,
    de.facScale = 0.2,
    nGenes = 10000,
    dropout.mid,
    dropout.shape,
    dropout.type = "experiment",
    outfile,
    ...) {
    if (!file.exists(outfile)) {
        print(paste0("Processing ", sample_name, " with params:"))
        print(knitr::kable(cbind(ngroup, nbatch, batch.facLoc, de.facLoc, dropout.mid, dropout.shape)))
        set.seed(42)
        message("Simulating data")

        sim <- splatSimulate(
            batchCells = rep(groupCells * ngroup, nbatch),
            group.prob = rep(1 / ngroup, ngroup),
            batch.facLoc = batch.facLoc,
            batch.facScale = batch.facScale,
            batch.rmEffect = ifelse(batch.facLoc == 0, TRUE, FALSE),
            method = "groups",
            verbose = FALSE,
            out.prob = 0.001,
            de.prob = 0.1,
            de.facLoc = de.facLoc,
            de.facScale = ifelse(de.facLoc == 0, 0, de.facScale),
            bcv.common = 0.1,
            nGenes = nGenes,
            dropout.mid = dropout.mid,
            dropout.shape = dropout.shape,
            dropout.type = dropout.type,
            ...
        )

        save_sce_as_h5ad(sim, outfile, uns = splatparams_to_list(sim))
        message(paste0("Saved ", outfile))
    } else {
        message(paste0(outfile, " already exists. Skipping..."))
    }
}

compute_sparsity <- function(h5ad_path) {
    ad <- anndataR::read_h5ad(h5ad_path)
    x <- ad$X
    100 * Matrix::mean(x == 0)
}
```

## 1. Sparsity scenario: fresh dropout simulation (ngroup=4)

```{r sparsity-sim}
dropout.mid <- 3.5
dropout.shape <- -1

sparsity_sample_name <- with(BASE_PARAMS, paste0(
    "batch.facLoc", batch.facLoc, "_de.facLoc", de.facLoc,
    "_nbatch", nbatch, "_ngroup", ngroup, "_groupCells", groupCells,
    "_dropout.mid", dropout.mid, "_dropout.shape", dropout.shape
))
sparsity_outfile <- file.path(OUTDIR, paste0(sparsity_sample_name, ".h5ad"))

create_and_save_sim(
    sample_name = sparsity_sample_name,
    ngroup = BASE_PARAMS$ngroup,
    nbatch = BASE_PARAMS$nbatch,
    groupCells = BASE_PARAMS$groupCells,
    batch.facLoc = BASE_PARAMS$batch.facLoc,
    de.facLoc = BASE_PARAMS$de.facLoc,
    batch.facScale = BASE_PARAMS$batch.facScale,
    de.facScale = BASE_PARAMS$de.facScale,
    nGenes = BASE_PARAMS$nGenes,
    dropout.mid = dropout.mid,
    dropout.shape = dropout.shape,
    outfile = sparsity_outfile
)
```

## 2. Load baseline and check Batch / Group columns

```{r load-baseline}
baseline_sce <- anndataR::read_h5ad(BASELINE_H5AD, to = "SingleCellExperiment")
baseline_obs <- colData(baseline_sce) %>% as.data.frame()

stopifnot(all(c("Batch", "Group") %in% colnames(baseline_obs)))
cat("Groups:", paste(sort(unique(baseline_obs$Group)), collapse = ", "), "\n")
cat("Batches:", paste(sort(unique(baseline_obs$Batch)), collapse = ", "), "\n")
table(baseline_obs$Batch, baseline_obs$Group)
```

## 3. Leave-one-group-out (LOGO)

```{r logo}
groups <- sort(unique(baseline_obs$Group))

logo_rows <- lapply(groups, function(g) {
    sample_name <- with(BASE_PARAMS, paste0(
        "batch.facLoc", batch.facLoc, "_de.facLoc", de.facLoc,
        "_nbatch", nbatch, "_ngroup", ngroup, "_groupCells", groupCells,
        "_dropGroup", g
    ))
    outfile <- file.path(OUTDIR, paste0(sample_name, ".h5ad"))
    if (!file.exists(outfile)) {
        keep <- baseline_obs$Group != g
        sub_sce <- baseline_sce[, keep]
        save_sce_as_h5ad(sub_sce, outfile)
    } else {
        message(paste0(outfile, " already exists. Skipping..."))
    }
    data.frame(
        sample_name = sample_name,
        file_path = outfile,
        perturbation_type = "logo",
        perturbation_detail = g,
        stringsAsFactors = FALSE
    )
})
logo_rows <- bind_rows(logo_rows)
logo_rows
```

## 4. Leave-one-batch-out (LOBO)

```{r lobo}
batches <- sort(unique(baseline_obs$Batch))

lobo_rows <- lapply(batches, function(b) {
    sample_name <- with(BASE_PARAMS, paste0(
        "batch.facLoc", batch.facLoc, "_de.facLoc", de.facLoc,
        "_nbatch", nbatch, "_ngroup", ngroup, "_groupCells", groupCells,
        "_dropBatch", b
    ))
    outfile <- file.path(OUTDIR, paste0(sample_name, ".h5ad"))
    if (!file.exists(outfile)) {
        keep <- baseline_obs$Batch != b
        sub_sce <- baseline_sce[, keep]
        save_sce_as_h5ad(sub_sce, outfile)
    } else {
        message(paste0(outfile, " already exists. Skipping..."))
    }
    data.frame(
        sample_name = sample_name,
        file_path = outfile,
        perturbation_type = "lobo",
        perturbation_detail = b,
        stringsAsFactors = FALSE
    )
})
lobo_rows <- bind_rows(lobo_rows)
lobo_rows
```

## 5. Composition skew

Unlike LOGO, every group stays present -- cells are subsampled per group
*within each batch* to hit the target proportions, keeping the same
per-batch total cell budget (`groupCells` = 1000) as the baseline so batch
structure and dataset size are otherwise unchanged.

```{r composition-helper}
subset_composition <- function(sce, obs, target_props, budget, seed = 42) {
    set.seed(seed)
    stopifnot(abs(sum(target_props) - 1) < 1e-8)
    batches <- sort(unique(obs$Batch))
    keep_idx <- integer(0)
    for (b in batches) {
        batch_idx <- which(obs$Batch == b)
        batch_group <- obs$Group[batch_idx]
        for (g in names(target_props)) {
            n_target <- round(target_props[[g]] * budget)
            g_idx <- batch_idx[batch_group == g]
            n_draw <- min(n_target, length(g_idx))
            if (n_draw < n_target) {
                warning(sprintf("Batch %s, %s: wanted %d cells, only %d available", b, g, n_target, length(g_idx)))
            }
            keep_idx <- c(keep_idx, sample(g_idx, n_draw))
        }
    }
    sce[, sort(keep_idx)]
}

composition_scenarios <- list(
    skewed10_20_30_40 = c(Group1 = 0.10, Group2 = 0.20, Group3 = 0.30, Group4 = 0.40),
    rareGroup1_1pct    = c(Group1 = 0.01, Group2 = 0.33, Group3 = 0.33, Group4 = 0.33)
)
```

```{r composition}
composition_rows <- lapply(names(composition_scenarios), function(scenario_name) {
    target_props <- composition_scenarios[[scenario_name]]
    sample_name <- with(BASE_PARAMS, paste0(
        "batch.facLoc", batch.facLoc, "_de.facLoc", de.facLoc,
        "_nbatch", nbatch, "_ngroup", ngroup, "_groupCells", groupCells,
        "_composition_", scenario_name
    ))
    outfile <- file.path(OUTDIR, paste0(sample_name, ".h5ad"))
    if (!file.exists(outfile)) {
        sub_sce <- subset_composition(baseline_sce, baseline_obs, target_props, budget = BASE_PARAMS$groupCells)
        save_sce_as_h5ad(sub_sce, outfile)
    } else {
        message(paste0(outfile, " already exists. Skipping..."))
    }
    data.frame(
        sample_name = sample_name,
        file_path = outfile,
        perturbation_type = "composition",
        perturbation_detail = paste(names(target_props), paste0(round(target_props * 100), "%"), sep = "=", collapse = ";"),
        stringsAsFactors = FALSE
    )
})
composition_rows <- bind_rows(composition_rows)
composition_rows
```

## 6. Assemble samplesheet

```{r samplesheet}
baseline_sample_name <- with(BASE_PARAMS, paste0(
    "batch.facLoc", batch.facLoc, "_de.facLoc", de.facLoc,
    "_nbatch", nbatch, "_ngroup", ngroup, "_groupCells", groupCells
))

baseline_row <- data.frame(
    sample_name = baseline_sample_name,
    file_path = BASELINE_H5AD,
    perturbation_type = "baseline",
    perturbation_detail = NA_character_,
    stringsAsFactors = FALSE
)

sparsity_row <- data.frame(
    sample_name = sparsity_sample_name,
    file_path = sparsity_outfile,
    perturbation_type = "sparsity",
    perturbation_detail = paste0("dropout.mid", dropout.mid),
    stringsAsFactors = FALSE
)

samplesheet <- bind_rows(baseline_row, logo_rows, lobo_rows, sparsity_row, composition_rows) %>%
    mutate(
        batch.facLoc = BASE_PARAMS$batch.facLoc,
        de.facLoc = BASE_PARAMS$de.facLoc,
        nbatch = BASE_PARAMS$nbatch,
        ngroup = BASE_PARAMS$ngroup,
        groupCells = BASE_PARAMS$groupCells
    )

samplesheet
```

## 7. Sparsity check across all samples

```{r sparsity-check}
samplesheet$sparsity_pct <- vapply(samplesheet$file_path, compute_sparsity, numeric(1))
samplesheet %>% select(sample_name, perturbation_type, perturbation_detail, sparsity_pct)
```

```{r write-samplesheet}
samplesheet_path <- file.path(OUTDIR, "samplesheet.tsv")
write.table(samplesheet, file = samplesheet_path, sep = "\t", row.names = FALSE, quote = FALSE)
cat("Wrote samplesheet to", samplesheet_path, "\n")
```

## Session Info

```{r session-info}
sessionInfo()
```
