```{r}
.libPaths(.Library)
suppressPackageStartupMessages({
    library(splatter)
    library(stringr)
    library(dplyr)
    library(SPARSim)
    library(tidyverse)
    library(tibble)
    library(jsonlite)
    library(pbapply)
    library(anndataR)
})
```


# Sim 1
## Sim 1 parameters and samplesheet
```{r}
outdir <- "/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/h5ad/with_dropout"

fixed_params_sim1 <- list(
    # Six batches with 10000 cells each. Needs to be more than we want in the
    # final simulation.
    batchCells = c(1000, 1000, 1000, 1000, 1000, 1000) * 10,
    batch.facLoc = c(0.10, 0.15, 0.12, 0.18, 0.15, 0.25),
    batch.facScale = c(0.10, 0.08, 0.12, 0.10, 0.12, 0.18),
    group.prob = rep(1, 7) / 7,
    # Differential expression by group
    de.prob = c(0.10, 0.12, 0.08, 0.20, 0.12, 0.10, 0.16),
    de.facLoc = c(0.10, 0.08, 0.12, 0.18, 0.06, 0.20, 0.14),
    de.facScale = c(0.40, 0.30, 0.45, 0.48, 0.42, 0.38, 0.36)
    # Seed
)
add_col_from_list <- function(tib, l) {
    if (is.null(names(l))) {
        stop("list must be named")
    }
    l <- lapply(l, function(x) list(x))
    for (x in names(l)) {
        tib <- tib %>% add_column(!!x := l[[x]])
    }
    return(tib)
}

samplesheet_sim1 <- tibble(
    seed = 1,
    lib.loc = 12,
    dropout.type = "experiment",
    dropout.shape = -5,
    dropout.mid = seq(0.5, 3.5, length.out = 10)
) 
    
samplesheet_sim1 <- add_col_from_list(samplesheet_sim1, fixed_params_sim1)
samplesheet_sim1$sim <- "sim1"
samplesheet_sim1$sample_name <- paste0(samplesheet_sim1$sim, "_DropoutSimulation_dropout.type", samplesheet_sim1$dropout.type, "_dropout.shape", abs(samplesheet_sim1$dropout.shape), "_dropout.mid", samplesheet_sim1$dropout.mid)
samplesheet_sim1$file_path <- file.path(outdir, paste0(samplesheet_sim1$sample_name, ".h5ad"))

# if dropout.type is "cells", dropout.mid and dropout.shape have 

```

## sample function from scib-reproducibility github
```{r}
#' Sample group proportions
#'
#' Downsample a Splat simulation to get the desired cell proportions
#'
#' @param sim SingleCellExperiment containing a Splat simulation
#' @param batch_ncells Final number of cells in each batch
#' @param batch_props Final cell proportions in each batch
#'
#' @return Downsampled SingleCellExperiment object
sample_group_props <- function(sim, batch_ncells, batch_props) {
    # Get the cell metadata
    col_data <- as.data.frame(colData(sim))

    # Get the current counts of each group in each batch
    sim_batch_group_count <- table(col_data$Batch, col_data$Group)

    # Get the final number of each group in each batch
    batch_group_ncells <- lapply(seq_along(batch_ncells), function(idx) {
        batch_ncells[idx] * batch_props[[idx]]
    })

    # Check that we aren't trying to get more cells than we have
    lapply(seq_along(batch_group_ncells), function(idx) {
        if (any(batch_group_ncells[[idx]] > sim_batch_group_count[idx, ])) {
            stop("Not enough cells for these proportions in batch ", idx)
        }
    })

    # Downsample cells
    selected <- lapply(seq_along(batch_group_ncells), function(batch) {
        group_cells <- batch_group_ncells[[batch]]
        is_batch <- col_data$Batch == paste0("Batch", batch)
        batch_groups <- col_data$Group[is_batch]
        # Downsample batch
        selected_batch <- sapply(seq_along(group_cells), function(group) {
            is_group <- batch_groups == paste0("Group", group)
            sample(col_data$Cell[is_batch][is_group], group_cells[group])
        })
        unlist(selected_batch)
    })
    selected <- as.character(unlist(selected))

    # Subset SingleCellExperiment
    sim[, selected]
}
```



## Sim 1 functions 
```{r}

aslist2 <- function(tibble) {
    l <- tibble %>% as.list()
    l <- lapply(l, function(x) {
        if (is.list(x)) {
            unlist(x)
        } else {
            x
        }
    })
    return(l)
}

perform_qc <- function(sim, batch_col="Batch"){

    if (batch_col %in% colnames(colData(sim))) {
        message("Calculating QC...")
        sim <- scater::addPerCellQC(sim)
        discard <- lapply(unique(colData(sim)[[batch_col]]), function(batch) {
            in_batch <- colData(sim)[[batch_col]] == batch
            scater::quickPerCellQC(colData(sim)[in_batch, ], nmads = 2)$discard
        })
        discard <- unlist(discard)
        colData(sim)$Discard <- discard

        message("Filtering cells...")
        sim <- sim[, !discard]
        message("Filtering genes...")
        sim <- scater::addPerFeatureQC(sim)
        is_exprs <- rowData(sim)$detected >= 0.01
        sim <- sim[is_exprs, ]
    } else {
        stop(paste0(batch_col, " not found in colData(sim)"))
    }

    return(sim)
}


simulate_row_params_sim1 <- function(row_params){
    params <- newSplatParams()
    params <- setParams(params, update = row_params %>% aslist2())
    sim1 <- splatSimulateGroups(params)

    # Number of cells in each batch in the final simulation
    batch_ncells <- c(300, 250, 220, 200, 180, 100) * 10
    # Proportions in each group in each batch that we want. Should sum to 1.
    batch_props <- list(
        c(0.35, 0.22, 0.15, 0.13, 0.10, 0.05, 0.00),
        c(0.25, 0.33, 0.20, 0.10, 0.00, 0.08, 0.04),
        c(0.30, 0.30, 0.00, 0.20, 0.10, 0.10, 0.00),
        c(0.15, 0.20, 0.35, 0.15, 0.15, 0.00, 0.00),
        c(0.40, 0.05, 0.10, 0.10, 0.17, 0.12, 0.06),
        c(0.45, 0.35, 0.15, 0.05, 0.00, 0.00, 0.00)
    )


    # Downsample cells in our simulation
    message("Downsampling cells...")
    sim1 <- sample_group_props(sim1, batch_ncells, batch_props)
    # Set proportion of counts for each batch (relative library size)
    batch_count_props <- c(
        Batch1 = 0.30,
        Batch2 = 0.25,
        Batch3 = 0.32,
        Batch4 = 0.45,
        Batch5 = 0.28,
        Batch6 = 1.00
    )
    cell_count_props <- batch_count_props[colData(sim1)$Batch]
    # Downsample counts
    message("Downsampling counts...")
    counts(sim1) <- DropletUtils::downsampleMatrix(counts(sim1), cell_count_props,
        bycol = TRUE
    )

    return(sim1)
}

write_h5ad_from_sce <- function(sim, outfile) {
    if (!file.exists(outfile)) {
        message("Extracting counts and metadata...")
        counts <- t(counts(sim)) %>% as.matrix() # for AnnData the matrix is transposed: Gene in columns Cells in rows
        metadata <- colData(sim) %>% as.data.frame()
        var <- rowData(sim) %>% as.data.frame()

        message("Creating AnnData object")
        adata <- anndataR::AnnData(
            X = counts,
            obs = metadata,
            var = var
        )
        message(paste0("Saving ", outfile))
        anndataR::write_h5ad(adata, path = outfile, mode = "w", compression = "none")
    } else {
        message(paste0(outfile, " already exists. Skipping..."))
    }
}

```

## run and save sim 1
```{r}

pblapply(seq_len(nrow(samplesheet_sim1)), function(i) {
    set.seed(42)
    dparam <- samplesheet_sim1[i, ] %>% select(
        seed,
        lib.loc,
        dropout.type,
        dropout.shape,
        dropout.mid,
        batchCells,
        batch.facLoc,
        batch.facScale,
        group.prob,
        de.prob,
        de.facLoc,
        de.facScale
    )
    outfile <- samplesheet_sim1[i, ]$file_path
    if (!file.exists(outfile)) {
        message("beep...")
        sim <- simulate_row_params_sim1(dparam)
        sim <- perform_qc(sim)
        write_h5ad_from_sce(sim, outfile = outfile)
    } else {
        message(paste0(outfile, " already exists. Skipping..."))
    }
}, cl = 10)
```


# Sim 2
## Sim 2 parameters and samplesheet
```{r}
outdir <- "/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/h5ad/with_dropout"

fixed_params_sim2 <- list(
    # Six batches with 10000 cells each. Needs to be more than we want in the
    # final simulation.    
    batchCells     = c(1000, 1000, 1000, 1000) * 10,
    batch.facLoc   = c(0.20, 0.25, 0.22, 0.28),
    batch.facScale = c(0.10, 0.08, 0.12, 0.10),
    # Groups with equal probabilities
    group.prob     = rep(1, 4) / 4,
    # Differential expression by group
    de.prob        = c(0.10, 0.12, 0.08, 0.20),
    de.facLoc      = c(0.10, 0.08, 0.12, 0.18),
    de.facScale    = c(0.40, 0.30, 0.45, 0.48)
    # Seed
)

samplesheet_sim2 <- tibble(
    seed = 1,
    lib.loc = 12,
    dropout.type = "experiment",
    dropout.shape = -5,
    dropout.mid = seq(0.5, 3.5, length.out = 10)
) 
    
samplesheet_sim2 <- add_col_from_list(samplesheet_sim2, fixed_params_sim2)
samplesheet_sim2$sim <- "sim2"
samplesheet_sim2$sample_name <- paste0(samplesheet_sim2$sim, "_DropoutSimulation_dropout.type", samplesheet_sim2$dropout.type, "_dropout.shape", abs(samplesheet_sim2$dropout.shape), "_dropout.mid", samplesheet_sim2$dropout.mid)
samplesheet_sim2$file_path <- file.path(outdir, paste0(samplesheet_sim2$sample_name, ".h5ad"))

# if dropout.type is "cells", dropout.mid and dropout.shape have 

```

## Functions Sim 2
```{r}
# CHUNK FUNCTION FROM BBMISC. COMMENT: check license
#' Chunk elements of vectors into blocks of nearly equal size.
#'
#' In case of shuffling and vectors that cannot be chunked evenly,
#' it is chosen randomly which levels / chunks will receive 1 element less.
#' If you do not shuffle, always the last chunks will receive 1 element less.
#'
#' @param x [ANY]\cr
#'   Vector, list or other type supported by \code{\link[base]{split}}.
#' @param chunk.size [\code{integer(1)}]\cr
#'   Requested number of elements in each chunk.
#'   Cannot be used in combination with \code{n.chunks} or \code{props}.
#'   If \code{x} cannot be evenly chunked, some chunks will have less elements.
#' @param n.chunks [\code{integer(1)}]\cr
#'   Requested number of chunks.
#'   If more chunks than elements in \code{x} are requested, empty chunks are
#'   dropped.
#'   Can not be used in combination with \code{chunks.size} or \code{props}.
#' @param props [\code{numeric}]\cr
#'   Vector of proportions for chunk sizes.
#'   Empty chunks may occur, depending on the length of \code{x} and the given
#'   proportions.
#'   Cannot be used in combination with \code{chunks.size} or \code{n.chunks}.
#' @param shuffle [\code{logical(1)}]\cr
#'   Shuffle \code{x}?
#'   Default is \code{FALSE}.
#' @return [unnamed \code{list}] of chunks.
#' @export
#' @examples
#' xs <- 1:10
#' chunk(xs, chunk.size = 3)
#' chunk(xs, n.chunks = 2)
#' chunk(xs, n.chunks = 2, shuffle = TRUE)
#' chunk(xs, props = c(7, 3))
chunk <- function(x, props, shuffle = FALSE) {
    library(checkmate)
    assertFlag(shuffle)
    nx <- length(x)

    ch <-  {
            assertNumeric(props, min.len = 1L, any.missing = FALSE, lower = 0)
            props <- props / sum(props)
            ch <- factor(rep.int(seq_along(props), round(props * nx, digits = 0L)),
                levels = seq_along(props)
            )
            if (shuffle) sample(ch) else ch
        }

    unname(split(x, ch))
}



add_subbatches <- function(sim, subbatch_props) {
    message("Adding subbatches...")

    params <- metadata(sim)$Params
    n_batches <- getParam(params, "nBatches")

    if (!all(sapply(subbatch_props, sum) == 1)) {
        stop("Not all subbatch props sum to 1")
    }

    colData(sim)$SubBatch <- NA
    # subbatch_list <- list()
    for (batch_idx in seq_len(n_batches)){
        message("Processing Batch ", batch_idx, "...")

        batch <- paste0("Batch", batch_idx)
        is_batch <- colData(sim)$Batch == batch
        batch_size <- sum(is_batch)

        subbatches <- chunk(colnames(sim)[is_batch],
            props = subbatch_props[[batch_idx]],
            shuffle = FALSE
        )

        for (subbatch_idx in seq_along(subbatches)) {
            message("Adding SubBatch ", subbatch_idx, "...")

            subbatch_cells <- subbatches[[subbatch_idx]]

            noise_sim <- splatSimulate(
                batchCells = length(subbatch_cells),
                lib.loc = 8.00,
                lib.scale = 0.50,
                verbose = FALSE,
                seed = batch_idx * subbatch_idx,
                dropout.mid = getParam(params, "dropout.mid"),
                dropout.shape = getParam(params, "dropout.shape")
            )

            message("Adding SubBatch ", subbatch_idx, " counts to simulation...")
            tmp_counts <- counts(sim)[, subbatch_cells, drop = FALSE] + counts(noise_sim)
            tmp_counts <- tmp_counts %>% as.matrix()
            full_counts <- counts(sim) %>% as.matrix()
            full_counts[, subbatch_cells] <- full_counts[, subbatch_cells] + tmp_counts
            full_counts <- as(full_counts, "dgCMatrix")
            counts(sim) <- full_counts
            message("Done adding counts")
            colData(sim)[subbatch_cells, "Sub"] <- paste0("Sub", subbatch_idx)
            colData(sim)[subbatch_cells, "SubBatch"] <- paste0(
                batch,
                "Sub", subbatch_idx
            )

            # subbatch_list[[paste0(batch, "Sub", subbatch_idx)]]$sim <- noise_sim
            # subbatch_list[[paste0(batch, "Sub", subbatch_idx)]]$cells <- subbatch_cells
        }
    }

    return(sim)
}

simulate_row_params_sim2 <- function(row_params) {
    params <- newSplatParams()
    params <- setParams(params, update = row_params %>% aslist2())
    sim2 <- splatSimulateGroups(params)

    # Number of cells in each batch in the final simulation
    batch_ncells <- c(500, 500, 500, 500) * 10

    # Proportions in each group in each batch that we want. Should sum to 1.
    batch_props <- list(
        c(0.35, 0.22, 0.25, 0.18),
        c(0.25, 0.40, 0.27, 0.10),
        c(0.30, 0.15, 0.35, 0.20),
        c(0.10, 0.30, 0.20, 0.40)
    )

    # Downsample cells in our simulation
    message("Downsampling cells...")
    sim2 <- sample_group_props(sim2, batch_ncells, batch_props)

    # Set proportion of counts for each batch (relative library size)
    batch_count_props <- c(
        Batch1 = 1.00,
        Batch2 = 0.70,
        Batch3 = 0.60,
        Batch4 = 0.30
    )

    cell_count_props <- batch_count_props[colData(sim2)$Batch]

    # Add subbatches
    subbatch_props <- list(
        c(0.25, 0.25, 0.25, 0.25),
        c(0.30, 0.30, 0.20, 0.20),
        c(0.40, 0.25, 0.25, 0.10),
        c(0.35, 0.35, 0.15, 0.15)
    )

    sim2 <- add_subbatches(sim2, subbatch_props)
    colData(sim2)["OldBatch"] <- colData(sim2)["Batch"]
    colData(sim2)["Batch"] <- colData(sim2)["SubBatch"]
    # Downsample counts
    message("Downsampling counts...")
    counts(sim2) <- DropletUtils::downsampleMatrix(counts(sim2), cell_count_props,
        bycol = TRUE
    )


    return(sim2)
}



```

## Run and save all sim 2
```{r}

pblapply(seq_len(nrow(samplesheet_sim2)), function(i) {
    # set.seed(42) # seed is already set in splatter simulation parmam
    dparam <- samplesheet_sim2[i, ] %>% select(
        seed,
        lib.loc,
        dropout.type,
        dropout.shape,
        dropout.mid,
        batchCells,
        batch.facLoc,
        batch.facScale,
        group.prob,
        de.prob,
        de.facLoc,
        de.facScale
    )
    outfile <- samplesheet_sim2[i, ]$file_path
    print(paste0("Processing ", outfile))
    if (!file.exists(outfile)) {
        message("beep...")
        sim <- simulate_row_params_sim2(dparam)
        sim <- perform_qc(sim, batch_col = "Batch")
        write_h5ad_from_sce(sim, outfile = outfile)
    } else {
        message(paste0(outfile, " already exists. Skipping..."))
    }
    return(NULL)
}, cl = 10)


```

# add measure of sparsitiy to the samplesheet metadata too
* reload the h5ad files and calculate sparsity of each dataset
```{r}
load_and_calc_sparsity <- function(file_path) {
    adata <- read_h5ad(file_path)
    counts <- adata$X
    total_elements <- nrow(counts) * ncol(counts)
    sparsity <- sum(counts == 0) / total_elements
    return(sparsity)
}
```



# Write samplesheet
```{r}
samplesheet_sim1$sparsity <- pblapply(samplesheet_sim1$file_path, load_and_calc_sparsity, cl = 10)
samplesheet_sim2$sparsity <- pblapply(samplesheet_sim2$file_path, load_and_calc_sparsity, cl = 10)

samplesheet <- rbind(samplesheet_sim1, samplesheet_sim2)

View(head(samplesheet))
samplesheet <- samplesheet %>% mutate(sparsity = round(as.numeric(sparsity), 4) * 100)
write_json(samplesheet, path = "/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/h5ad/with_dropout/samplesheet.json", pretty = TRUE, auto_unbox = TRUE)
```

## write samplesheet for SEACells pipeline
```{r}
colnames(samplesheet)
samplesheet %>%
    select(sample_name, file_path, dropout.mid, sparsity, sim) %>%
    mutate_if(is.list, as.numeric) %>%
    write.table(., file = "/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/h5ad/with_dropout/samplesheet.tsv", sep = "\t")
samplesheet_sim1 %>%
    select(sample_name, file_path, dropout.mid, sparsity, sim) %>%
    mutate_if(is.list, as.numeric) %>%
    write.table(., file = "/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/h5ad/with_dropout/samplesheet_sim1.tsv", sep = "\t")
samplesheet_sim2 %>%
    select(sample_name, file_path, dropout.mid, sparsity, sim) %>%
    mutate_if(is.list, as.numeric) %>%
    write.table(., file = "/local/Projects/AAkalin_Neuroblastoma/Results/simulate_single_cell/h5ad/with_dropout/samplesheet_sim2.tsv", sep = "\t")
```

```{r}
View(head(samplesheet_sim2))
dparam <- samplesheet_sim2 %>%
    select(
        seed,
        lib.loc,
        dropout.type,
        dropout.shape,
        dropout.mid,
        batchCells,
        batch.facLoc,
        batch.facScale,
        group.prob,
        de.prob,
        de.facLoc,
        de.facScale
    ) %>%
    slice(1)
```



